---
title: "Supplementary Material"
execute:
  eval: !expr knitr::is_html_output()
---

::: {.content-visible when-format="pdf"}
Full LLM reasoning traces, assessment summaries, related work, and extended analyses are available in the online supplement at <https://llm-uj-research-eval.netlify.app/appendix_llm_traces.html>.
:::

::: {.content-visible when-format="html"}

## Extended Rating Analysis

The main text reports six core exhibits: cost overview, human-vs-LLM scatter plot, agreement metrics, human baseline (Krippendorff's alpha), criteria-level heatmap, and gap heatmap. The following extended analyses are preserved in the project's git history and available in earlier rendered versions:

- **Krippendorff's alpha by criterion**: Inter-rater agreement treating human mean and each LLM as separate raters
- **CI coverage analysis**: Share of human mean ratings falling within LLM 90% credible intervals
- **Bootstrap 95% CIs**: Nonparametric confidence intervals for Pearson correlations
- **Mutual information**: Information-theoretic measure of human-LLM rating dependence
- **Per-criterion correlations**: Pearson and Spearman by evaluation criterion and model
- **Tier correlation dumbbell plot**: How criteria predict journal tier judgments for humans vs. LLMs
- **Top disagreement cases**: Papers with largest human-LLM rating differences
- **Prompt comparison**: Legacy vs. updated GPT-5 Pro prompt effect on ratings

## Extended Critique Analysis

The main text reports coverage-precision scatter and summary statistics. The following extended analyses are available in the [full online version](https://llm-uj-research-eval.netlify.app/results_critiques.html):

- **Paper-by-paper issue comparisons**: Interactive collapsible panels showing matched issue pairs, unmatched human/LLM issues, and detailed discussion for each paper
- **Aggregate statistics**: Total issue counts by severity, human vs. LLM issue distributions
- **Observable structural differences**: Format, structure, attribution, and length comparisons
- **Manual annotation tool**: Browser-based UI for systematic concordance labeling

---

::: {.content-visible when-format="pdf"}
*The per-paper LLM evaluation traces below are only available in the [online HTML version](https://llm-uj-research-eval.netlify.app/appendix_llm_traces.html).*
:::

::: {.content-visible when-format="html"}
This appendix presents detailed LLM evaluations for each paper, including:

- **Numerical ratings** from all six models and human evaluators (where available)
- **Assessment summaries** (collapsible) — the model's written evaluation; available for all models including Claude Opus 4.6
- **Key issues** (GPT-5.2 Pro only) — numbered list of methodological and evidential concerns
- **Reasoning traces** (collapsible, GPT-5 Pro only) — internal reasoning process; not available for Claude Opus 4.6 (extended thinking was not enabled for this run)

The appendix is organized into two sections:

1. **Focal Papers** — 14 papers evaluated with GPT-5.2 Pro using an extended prompt that produces structured key issues
2. **Other Papers** — remaining papers with standard evaluations

Click on the callout headers to expand/collapse each section.

```{r}
#| label: appendix-setup
#| code-fold: true
#| code-summary: "Setup and helper functions"

library(jsonlite)
library(purrr)
library(dplyr)
library(tidyr)
library(stringr)
library(here)
library(knitr)
library(kableExtra)

`%||%` <- function(x, y) if (!is.null(x)) x else y

# Model directories to scan
MODEL_DIRS <- list(
  "gpt5_pro_updated_jan2026" = "GPT-5 Pro",
  "gpt52_pro_focal_jan2026" = "GPT-5.2 Pro",
  "gpt_4o_mini_2024_07_18" = "GPT-4o-mini",
  "claude_sonnet_4_20250514" = "Claude Sonnet 4",
  "claude_opus_4_6" = "Claude Opus 4.6",
  "gemini_2.0_flash" = "Gemini 2.0 Flash"
)

# Focal papers (evaluated with GPT-5.2 Pro with key_issues)
FOCAL_PAPERS <- c(
  "Acemoglu et al. 2024",
  "Adena and Hager 2024",
  "Benabou et al. 2023",
  "Bilal and Kaenzig 2024",
  "Blimpo and Castaneda-Dower 2025",
  "Bruers 2021",
  "Clancy 2024",
  "Dullaghan and Zhang 2022",
  "Frech et al. 2023",
  "Green et al. 2025",
  "McGuire et al. 2024",
  "Peterman et al. 2025",
  "Weaver et al. 2025",
  "Williams et al. 2024"
)

# Canonical metric names
canon_metric <- function(x) dplyr::recode(

  x,
  "advancing_knowledge" = "Adv. Knowledge",
  "adv_knowledge" = "Adv. Knowledge",
  "open_science" = "Open Science",
  "open_sci" = "Open Science",
  "logic_communication" = "Logic & Comms",
  "logic_comms" = "Logic & Comms",
  "global_relevance" = "Global Relevance",
  "gp_relevance" = "Global Relevance",
  "claims_evidence" = "Claims",
  "claims" = "Claims",
  "overall" = "Overall",
  "methods" = "Methods",
  .default = x
)

# Parse a single JSON response file
parse_paper_json <- function(path, model_name) {
  tryCatch({
    r <- jsonlite::fromJSON(path, simplifyVector = FALSE)

    # Extract parsed metrics (check r$parsed first, then output_text)
    parsed <- NULL
    if (!is.null(r$parsed) && length(r$parsed) > 0) {
      parsed <- r$parsed
    } else if (!is.null(r$output_text) && nchar(r$output_text) > 0) {
      txt <- r$output_text
      txt <- sub("^\\s*```[a-z]*\\s*\n?", "", txt)
      txt <- sub("\\s*```\\s*$", "", txt)
      parsed <- jsonlite::fromJSON(txt, simplifyVector = TRUE)
    } else if (!is.null(r$output)) {
      msg <- purrr::detect(r$output, ~ .x$type == "message", .default = NULL)
      if (!is.null(msg) && length(msg$content) > 0) {
        parsed <- jsonlite::fromJSON(msg$content[[1]]$text, simplifyVector = TRUE)
      }
    }

    # Extract metrics
    metrics_df <- NULL
    if (!is.null(parsed$metrics)) {
      rows <- list()
      for (nm in names(parsed$metrics)) {
        m <- parsed$metrics[[nm]]
        if (nm %in% c("tier_should", "tier_will", "journal_should", "journal_will")) {
          rows[[length(rows) + 1]] <- tibble(
            metric = nm,
            midpoint = m$score,
            lower = m$ci_lower,
            upper = m$ci_upper
          )
        } else {
          rows[[length(rows) + 1]] <- tibble(
            metric = nm,
            midpoint = m$midpoint,
            lower = m$lower_bound,
            upper = m$upper_bound
          )
        }
      }
      metrics_df <- bind_rows(rows) |> mutate(model = model_name)
    }

    # Extract assessment summary
    assessment <- parsed$assessment_summary

    # Extract key_issues (GPT-5.2 Pro focal run only)
    key_issues <- NULL
    if (!is.null(parsed$key_issues) && length(parsed$key_issues) > 0) {
      key_issues <- parsed$key_issues
    }

    # Extract reasoning trace (GPT-5 Pro only)
    reasoning <- NA_character_
    if (!is.null(r$output)) {
      reasoning_block <- purrr::detect(r$output, ~ .x$type == "reasoning", .default = NULL)
      if (!is.null(reasoning_block) && !is.null(reasoning_block$summary)) {
        texts <- purrr::map_chr(reasoning_block$summary, ~ .x$text %||% "")
        reasoning <- paste(texts, collapse = "\n\n")
      }
    }

    list(
      metrics = metrics_df,
      assessment = assessment,
      key_issues = key_issues,
      reasoning = reasoning
    )
  }, error = function(e) {
    NULL
  })
}

# Load all LLM data for a specific paper
load_paper_llm_data <- function(paper_name) {
  all_metrics <- list()
  all_assessments <- list()
  all_key_issues <- list()
  reasoning <- NA_character_

  for (dir_name in names(MODEL_DIRS)) {
    model_name <- MODEL_DIRS[[dir_name]]

    # Try different filename formats
    json_dir <- here("results", dir_name, "json")
    if (!dir.exists(json_dir)) next

    # Look for matching file
    paper_clean <- str_replace_all(paper_name, " ", "_")
    possible_files <- c(
      file.path(json_dir, paste0(paper_name, ".response.json")),
      file.path(json_dir, paste0(paper_clean, ".response.json"))
    )

    path <- NULL
    for (pf in possible_files) {
      if (file.exists(pf)) {
        path <- pf
        break
      }
    }

    if (is.null(path)) next

    result <- parse_paper_json(path, model_name)
    if (!is.null(result)) {
      if (!is.null(result$metrics)) {
        all_metrics[[length(all_metrics) + 1]] <- result$metrics
      }
      if (!is.null(result$assessment) && !is.na(result$assessment)) {
        all_assessments[[model_name]] <- result$assessment
      }
      if (!is.null(result$key_issues) && length(result$key_issues) > 0) {
        all_key_issues[[model_name]] <- result$key_issues
      }
      if (!is.na(result$reasoning) && nchar(result$reasoning) > 0) {
        reasoning <- result$reasoning
      }
    }
  }

  list(
    metrics = if (length(all_metrics) > 0) bind_rows(all_metrics) else NULL,
    assessments = all_assessments,
    key_issues = all_key_issues,
    reasoning = reasoning
  )
}

# Load human data
load_human_ratings <- function() {
  rsx_path <- here("data", "rsx_evalr_rating.csv")
  if (!file.exists(rsx_path)) return(NULL)

  UJmap <- tryCatch(
    read.csv(here("data", "UJ_map.csv"), sep = ";", stringsAsFactors = FALSE) |>
      as_tibble() |>
      transmute(label_paper_title = research, label_paper = paper),
    error = function(e) NULL
  )

  rsx <- read.csv(rsx_path, stringsAsFactors = FALSE) |>
    as_tibble() |>
    janitor::clean_names() |>
    mutate(label_paper_title = research) |>
    select(-research)

  if (!is.null(UJmap)) {
    rsx <- rsx |>
      mutate(label_paper_title = str_trim(label_paper_title)) |>
      left_join(
        UJmap |> distinct(label_paper_title, label_paper),
        by = "label_paper_title",
        relationship = "many-to-one"
      )
  }

  rsx |>
    filter(!is.na(label_paper)) |>
    group_by(paper = label_paper, criteria) |>
    summarise(
      midpoint = mean(as.numeric(middle_rating), na.rm = TRUE),
      lower = mean(as.numeric(lower_ci), na.rm = TRUE),
      upper = mean(as.numeric(upper_ci), na.rm = TRUE),
      n_evals = n(),
      .groups = "drop"
    ) |>
    mutate(
      model = "Human",
      metric = criteria
    ) |>
    select(paper, model, metric, midpoint, lower, upper, n_evals)
}

human_ratings <- load_human_ratings()

# Get all unique papers across all model directories
get_all_papers <- function() {
  papers <- c()

  for (dir_name in names(MODEL_DIRS)) {
    json_dir <- here("results", dir_name, "json")
    if (!dir.exists(json_dir)) next

    files <- list.files(json_dir, pattern = "\\.response\\.json$", full.names = FALSE)
    paper_names <- files |>
      str_replace("\\.response\\.json$", "") |>
      str_replace_all("_", " ")

    papers <- c(papers, paper_names)
  }

  sort(unique(papers))
}

all_papers <- get_all_papers()
cat(sprintf("Found %d papers with LLM evaluations\n", length(all_papers)))
```

```{r}
#| label: appendix-render
#| results: 'asis'
#| echo: false

# Metric display order
metric_order <- c("Overall", "Claims", "Methods", "Adv. Knowledge",
                  "Logic & Comms", "Open Science", "Global Relevance",
                  "tier_should", "tier_will")

# Separate focal papers from others
focal_in_corpus <- all_papers[all_papers %in% FOCAL_PAPERS]
other_papers <- all_papers[!all_papers %in% FOCAL_PAPERS]

# --- Focal Papers Section ---
if (length(focal_in_corpus) > 0) {
  cat("\n\n# Focal Papers (GPT-5.2 Pro with Key Issues)\n\n")
  cat("These 14 papers were evaluated with GPT-5.2 Pro using an extended prompt that identifies key methodological and evidential issues.\n\n")
}

for (paper in c(focal_in_corpus, other_papers)) {

  # Add section header when transitioning to other papers

  if (paper == other_papers[1] && length(other_papers) > 0) {
    cat("\n\n# Other Papers\n\n")
  }

  is_focal <- paper %in% FOCAL_PAPERS
  focal_badge <- if (is_focal) " {.focal}" else ""

  cat("\n\n## ", paper, "\n\n", sep = "")

  # Load data for this paper
  llm_data <- load_paper_llm_data(paper)

  # --- Numerical Ratings Table (percentile metrics only) ---

  cat("### Ratings Comparison\n\n")

  if (!is.null(llm_data$metrics) && nrow(llm_data$metrics) > 0) {

    # Add human ratings for this paper if available
    combined_metrics <- llm_data$metrics |>
      mutate(
        metric_display = canon_metric(metric),
        n_evals = NA_integer_
      )

    if (!is.null(human_ratings)) {
      human_paper <- human_ratings |>
        filter(paper == !!paper) |>
        mutate(metric_display = canon_metric(metric)) |>
        filter(!is.na(metric_display), is.finite(midpoint), is.finite(lower), is.finite(upper))

      if (nrow(human_paper) > 0) {
        combined_metrics <- bind_rows(
          combined_metrics |> select(model, metric_display, midpoint, lower, upper, n_evals),
          human_paper    |> select(model, metric_display, midpoint, lower, upper, n_evals)
        )
      }
    }

    # Drop tier metrics for this appendix table to avoid cross-model inconsistencies
    combined_metrics <- combined_metrics |>
      filter(!metric_display %in% c("tier_should", "tier_will"))

    # Remove rows with missing metric labels or non-finite values (e.g., spurious rows)
    combined_metrics <- combined_metrics |>
      filter(!is.na(metric_display), is.finite(midpoint), is.finite(lower), is.finite(upper))

    # Collapse duplicates (e.g., multiple human evaluators) before pivoting
    combined_metrics <- combined_metrics |>
      group_by(model, metric_display) |>
      summarise(
        midpoint = mean(midpoint, na.rm = TRUE),
        lower    = mean(lower, na.rm = TRUE),
        upper    = mean(upper, na.rm = TRUE),
        n_evals  = suppressWarnings(max(n_evals, na.rm = TRUE)),
        .groups  = "drop"
      ) |>
      mutate(n_evals = ifelse(is.finite(n_evals), n_evals, NA_integer_))

    # Pivot to wide format for table
    ratings_wide <- combined_metrics |>
      mutate(
        rating_str = dplyr::case_when(
          model == "Human" & is.finite(n_evals) ~ sprintf("%.0f [%.0f–%.0f] (n=%d)", midpoint, lower, upper, n_evals),
          TRUE ~ sprintf("%.0f [%.0f–%.0f]", midpoint, lower, upper)
        )
      ) |>
      select(model, metric_display, rating_str) |>
      pivot_wider(names_from = model, values_from = rating_str, values_fill = "—")

    # Order metrics
    ratings_wide <- ratings_wide |>
      mutate(metric_display = factor(metric_display, levels = metric_order)) |>
      arrange(metric_display) |>
      rename(Metric = metric_display)

    # Print table
    print(knitr::kable(ratings_wide, align = c("l", rep("r", ncol(ratings_wide) - 1))) |>
      kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                                 full_width = FALSE, font_size = 12))
    cat("\n\n")

  } else {
    cat("*No numerical ratings available.*\n\n")
  }

  # --- Assessment Summaries (Collapsible) ---

  if (length(llm_data$assessments) > 0) {
    for (model_name in names(llm_data$assessments)) {
      assessment <- llm_data$assessments[[model_name]]
      if (!is.null(assessment) && !is.na(assessment) && nchar(assessment) > 0) {

        cat(sprintf("::: {.callout-note collapse=\"true\"}\n"))
        cat(sprintf("## %s Assessment\n\n", model_name))
        cat(assessment, "\n")
        cat(":::\n\n")
      }
    }
  } else {
    cat("::: {.callout-note collapse=\"true\"}\n")
    cat("## Assessment Summary\n\n")
    cat("*No assessment summary available.*\n")
    cat(":::\n\n")
  }

  # --- Key Issues (Collapsible, GPT-5.2 Pro focal run only) ---

  if (length(llm_data$key_issues) > 0) {
    for (model_name in names(llm_data$key_issues)) {
      issues <- llm_data$key_issues[[model_name]]
      if (length(issues) > 0) {
        cat(sprintf("::: {.callout-warning collapse=\"false\"}\n"))
        cat(sprintf("## %s: Key Issues Identified\n\n", model_name))
        for (i in seq_along(issues)) {
          cat(sprintf("%d. %s\n", i, issues[i]))
        }
        cat("\n:::\n\n")
      }
    }
  }

  # --- Reasoning Trace (Collapsible, GPT-5 Pro only) ---
  # Reasoning traces contain raw LLM "thinking" text with special characters
  # (e.g., literal \n references) that break LaTeX. Show in HTML only.

  if (!is.na(llm_data$reasoning) && nchar(llm_data$reasoning) > 50) {
    if (knitr::is_html_output()) {
      cat("::: {.callout-tip collapse=\"true\"}\n")
      cat("## GPT-5 Pro Reasoning Trace\n\n")
      cat("::: {.small}\n")
      cat(llm_data$reasoning, "\n")
      cat(":::\n")
      cat(":::\n\n")
    } else {
      cat("*GPT-5 Pro reasoning trace available in the [online HTML version](https://llm-uj-research-eval.netlify.app/appendix_llm_traces.html).*\n\n")
    }
  }

  cat("\n---\n\n")
}
```

:::

:::
