---
title: "Did Authors Respond to Unjournal Evaluations?"
author: "The Unjournal"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

::: {.callout-note}
Manual coding of 57 tracked evaluations, last updated April 2026. Two data sources: (a) manual assessment of author responses and paper revisions by Unjournal staff; (b) Claude Opus 4.6 change-attribution analysis comparing pre-evaluation PDFs against current NBER/arxiv versions, assessing whether paper updates reflect evaluator suggestions (see [Combined Evidence](#combined-evidence) below).
:::

## Overview

The Unjournal commissions open, structured evaluations of working papers in economics and social science. A natural question is whether this process actually changes research: do authors engage with the feedback and revise their work?

This page tabulates evidence from approximately 57 tracked evaluations based on manual assessment. For each paper where sufficient time has passed, Unjournal staff assessed whether the authors (a) formally or informally responded to the evaluation, and (b) produced an updated version of the paper that reflects the feedback.

For a subset of papers, automated PDF change detection (line-level diffing of before/after versions) supplements the manual coding. The broader background and framing is discussed in [The Unjournal's knowledge base](https://globalimpact.gitbook.io/the-unjournal-project-and-communication-space/benefits-and-features/more-reliable-and-useful-evaluation/author-engagement-evidence).

## Summary

```{r}
#| label: setup
#| code-summary: "Load data and define helpers"

library(dplyr)
library(ggplot2)
library(readr)
library(DT)
library(tidyr)
library(stringr)
library(glue)
library(jsonlite)

df <- read_csv("data/author_adjustment_manual.csv", show_col_types = FALSE)

# Rename columns for easier use
df <- df |>
  rename(
    paper_title     = label_paper_title,
    pubpub_link     = dup_pubpub_final_links,
    adj_status      = `Adjusted_paper?`,
    updated_manual  = `Updated since UJ report -- manual confirmation`,
    deposit_after   = `deposit date > unjournal pub date`,
    research_area   = main_cause_cat_abbrev,
    nb_link         = notebookLM_link,
    pub_status      = publication_status,
    wp_date         = working_paper_release_date,
    uj_pub_date     = publication_date_unjournal
  )

# Define ordered factor for adjustment status
adj_levels <- c(
  "evidence of updating",
  "Stated intention to update",
  "mixed evidence/minor updating",
  "Unlikely to update",
  ""
)

adj_labels <- c(
  "Evidence of updating",
  "Stated intention to update",
  "Mixed / minor updating",
  "Unlikely to update",
  "Not yet assessed"
)

df <- df |>
  mutate(
    adj_status_clean = case_when(
      adj_status == "evidence of updating"        ~ "Evidence of updating",
      adj_status == "Stated intention to update"  ~ "Stated intention to update",
      adj_status == "mixed evidence/minor updating" ~ "Mixed / minor updating",
      adj_status == "Unlikely to update"          ~ "Unlikely to update",
      TRUE                                        ~ "Not yet assessed"
    ),
    adj_status_clean = factor(adj_status_clean, levels = adj_labels),
    author_response_clean = case_when(
      author_response == "Formal response"        ~ "Formal response",
      author_response == "Informal"               ~ "Informal response",
      author_response == "None?"                  ~ "No response",
      TRUE                                        ~ "Not yet coded"
    )
  )

# Color palette aligned with project theme
adj_colors <- c(
  "Evidence of updating"        = "#2D9D5E",
  "Stated intention to update"  = "#5AE08A",
  "Mixed / minor updating"      = "#E8722A",
  "Unlikely to update"          = "#C45A1E",
  "Not yet assessed"            = "#CBD5E1"
)

theme_uj <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      axis.text = element_text(size = base_size * 0.85),
      plot.caption = element_text(size = base_size * 0.75, color = "grey50"),
      legend.position = "right"
    )
}
```

```{r}
#| label: summary-numbers
#| code-summary: "Compute headline statistics"

n_total <- nrow(df)
n_assessed <- df |> filter(adj_status_clean != "Not yet assessed") |> nrow()
n_evidence  <- df |> filter(adj_status_clean == "Evidence of updating") |> nrow()
n_intention <- df |> filter(adj_status_clean == "Stated intention to update") |> nrow()
n_any_positive <- df |>
  filter(adj_status_clean %in% c("Evidence of updating", "Stated intention to update", "Mixed / minor updating")) |>
  nrow()
n_formal <- df |> filter(author_response_clean == "Formal response") |> nrow()
n_any_response <- df |> filter(author_response_clean %in% c("Formal response", "Informal response")) |> nrow()
```

Of **`r n_total` tracked evaluations**:

- **`r n_any_response`** papers received a formal or informal author response
- **`r n_formal`** papers received a **formal written response** to the evaluation
- **`r n_evidence`** papers show **clear evidence of substantive updating** in response to the evaluation
- **`r n_intention`** authors stated intention to update their paper
- **`r n_any_positive`** papers show at least some positive signal (updating, intention, or minor revisions)
- **`r n_assessed`** of `r n_total` have been manually assessed; the remainder are too recent or pending follow-up

```{r}
#| label: fig-adj-bar
#| fig-cap: "Distribution of adjustment outcomes across tracked Unjournal evaluations. 'Not yet assessed' includes papers that are too recent or where follow-up is pending."
#| fig-height: 4

df |>
  count(adj_status_clean) |>
  ggplot(aes(x = reorder(adj_status_clean, -n), y = n, fill = adj_status_clean)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = adj_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Number of papers",
       title = "Author adjustment outcomes following Unjournal evaluations") +
  theme_uj() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

```{r}
#| label: fig-response-bar
#| fig-cap: "Type of author response to Unjournal evaluation."
#| fig-height: 3.5

resp_colors <- c(
  "Formal response"    = "#2B7CE9",
  "Informal response"  = "#5DA3F5",
  "No response"        = "#CBD5E1",
  "Not yet coded"      = "#E2E8F0"
)

df |>
  count(author_response_clean) |>
  ggplot(aes(x = reorder(author_response_clean, -n), y = n, fill = author_response_clean)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = resp_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Number of papers",
       title = "Type of author response to evaluation") +
  theme_uj() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

## Full Tabulation

The table below covers all 57 tracked evaluations. Click column headers to sort; use the search box to filter. Adjustment status is color-coded: green = positive signal, orange = minor/uncertain, grey = unlikely or not yet assessed.

```{r}
#| label: tbl-full
#| code-summary: "Build interactive DT table"

# Build display table
tbl <- df |>
  mutate(
    # Linked paper title
    title_linked = if_else(
      !is.na(pubpub_link) & pubpub_link != "",
      paste0('<a href="', pubpub_link, '" target="_blank">', str_trunc(paper_title, 70), '</a>'),
      str_trunc(paper_title, 70)
    ),
    # Shorten authors
    authors_short = str_trunc(authors, 45),
    # NotebookLM icon link
    nb_icon = if_else(
      !is.na(nb_link) & nb_link != "",
      paste0('<a href="', nb_link, '" target="_blank" title="NotebookLM analysis">&#128214;</a>'),
      ""
    ),
    # DOI link
    doi_linked = if_else(
      !is.na(doi) & doi != "" & !str_detect(doi, "^doi\\.org"),
      paste0('<a href="https://doi.org/', str_remove(doi, "^https://doi\\.org/"), '" target="_blank">DOI</a>'),
      if_else(
        !is.na(doi) & doi != "",
        paste0('<a href="', doi, '" target="_blank">DOI</a>'),
        ""
      )
    )
  ) |>
  select(
    Paper = title_linked,
    Authors = authors_short,
    Area = research_area,
    Response = author_response_clean,
    Adjustment = adj_status_clean,
    `Pub. status` = pub_status,
    NB = nb_icon
  )

# Color rows by adjustment status
adj_bg <- c(
  "Evidence of updating"        = "#d4edda",
  "Stated intention to update"  = "#e8f5e9",
  "Mixed / minor updating"      = "#fff3e0",
  "Unlikely to update"          = "#fce4d4",
  "Not yet assessed"            = "#f8fafc"
)

datatable(
  tbl,
  escape = FALSE,
  rownames = FALSE,
  filter = "top",
  options = list(
    pageLength = 15,
    autoWidth = FALSE,
    columnDefs = list(
      list(width = "35%", targets = 0),
      list(width = "20%", targets = 1),
      list(width = "10%", targets = 2),
      list(width = "10%", targets = 3),
      list(width = "12%", targets = 4),
      list(width = "10%", targets = 5),
      list(width = "3%",  targets = 6)
    ),
    dom = "lfrtip"
  ),
  caption = "All tracked Unjournal evaluations with author response and adjustment data. NB = link to NotebookLM analysis."
) |>
  formatStyle(
    "Adjustment",
    backgroundColor = styleEqual(
      names(adj_bg),
      unname(adj_bg)
    )
  )
```

## Combined Evidence of Paper Updating {#combined-evidence}

Two evidence streams are combined below:

- **Manual classification** (all 57 papers): Unjournal staff assessed whether authors engaged with feedback and revised their work.
- **LLM change-attribution** (8 papers): For papers with fetchable NBER/arxiv DOIs and meaningful text changes, Claude Opus 4.6 compared pre-evaluation PDFs against current versions in a 3-step pipeline — (a) identify major changes, (b) extract evaluator suggestions, (c) assess whether changes plausibly reflect suggestions.

::: {.callout-note collapse="true"}
## Pipeline and caveats

**Scripts** (run from project root): `scripts/fetch_latest_papers.py` → `scripts/run_paper_change_llm.py`

- **Before version**: PDF evaluated by Unjournal evaluators (`papers/`); **After version**: latest from NBER/arxiv, fetched April 2026
- **Model**: Claude Opus 4.6, temperature 0.2; instructed to be conservative — only "direct" or "indirect" attribution when there is clear conceptual alignment with a specific suggestion
- **Coverage**: 30 papers had fetchable DOIs; 8 showed meaningful text changes (≥15 line edits or ≥0.5% size change) and matched evaluation files; 3 papers lacked pre-evaluation PDFs; the 3 manually-confirmed updates were also LLM-analyzed using `--paper` to override the skip filter
- **Temporal caveat**: newer version ≠ evaluation-driven change; some papers were already in revision. The `deposit date > UJ pub date` flag helps but is not definitive
- **LLM limitations**: text truncated at 60,000 chars per version; may miss changes in mathematical/figure-dense sections
:::

```{r}
#| label: llm-attribution-load
#| code-summary: "Load and merge LLM + manual data"

nn <- function(x, default = NA) if (!is.null(x)) x else default

llm_results_path <- "data/paper_change_llm_results.json"

if (file.exists(llm_results_path)) {
  llm_raw <- fromJSON(llm_results_path, simplifyVector = FALSE)

  llm_df <- lapply(llm_raw, function(r) {
    oa <- r$overall_assessment
    data.frame(
      paper_title      = nn(r$paper_title, ""),
      adj_status_csv   = nn(r$adj_status, ""),
      deposit_after_uj = isTRUE(r$deposit_after_uj),
      text_chg_pct     = nn(r$text_length_change_pct, NA_real_),
      additions        = nn(r$additions_count, NA_integer_),
      deletions        = nn(r$deletions_count, NA_integer_),
      n_major_changes  = length(nn(r$major_changes, list())),
      n_suggestions    = length(nn(r$evaluator_suggestions, list())),
      pct_influenced   = if (!is.null(oa)) nn(oa$pct_likely_influenced, NA_real_) else NA_real_,
      attr_confidence  = if (!is.null(oa)) nn(oa$confidence, NA_real_) else NA_real_,
      narrative        = if (!is.null(oa)) nn(oa$narrative, "") else "",
      eval_found       = !is.null(r$eval_file) && nchar(nn(r$eval_file, "")) > 0,
      skipped          = !is.null(r$skipped_reason),
      stringsAsFactors = FALSE
    )
  }) |> bind_rows()

  llm_analyzed <- llm_df |>
    filter(!skipped & eval_found & n_major_changes > 0 & !is.na(pct_influenced))

  # Enrich with area and PubPub links from main df
  llm_enriched <- llm_analyzed |>
    left_join(
      df |> select(paper_title, research_area, pubpub_link),
      by = "paper_title"
    ) |>
    mutate(
      manual_tier = case_when(
        adj_status_csv == "evidence of updating"          ~ "Confirmed update",
        adj_status_csv == "Stated intention to update"    ~ "Stated intention",
        adj_status_csv == "mixed evidence/minor updating" ~ "Mixed / minor",
        adj_status_csv == "Unlikely to update"            ~ "Unlikely",
        TRUE                                              ~ "Unclassified"
      ),
      combined_tier = case_when(
        manual_tier == "Confirmed update" & pct_influenced >= 40 ~ "Confirmed + strong LLM",
        manual_tier == "Confirmed update"                        ~ "Confirmed + weaker LLM",
        pct_influenced >= 30                                     ~ "LLM: likely influenced",
        pct_influenced >= 10                                     ~ "LLM: possibly influenced",
        TRUE                                                     ~ "LLM: largely independent"
      ),
      combined_tier = factor(combined_tier, levels = c(
        "Confirmed + strong LLM", "Confirmed + weaker LLM",
        "LLM: likely influenced", "LLM: possibly influenced",
        "LLM: largely independent"
      ))
    )

  n_analyzed      <- nrow(llm_enriched)
  n_total_fetched <- nrow(llm_df)
  med_pct         <- median(llm_enriched$pct_influenced, na.rm = TRUE)
  n_high          <- sum(llm_enriched$pct_influenced >= 30, na.rm = TRUE)
} else {
  message("LLM results not found — run scripts/run_paper_change_llm.py first")
  llm_df <- llm_analyzed <- llm_enriched <- data.frame()
  n_analyzed <- n_total_fetched <- n_high <- 0
  med_pct <- NA_real_
}
```

```{r}
#| label: llm-attribution-summary
#| code-summary: "Summary callout"

if (nrow(llm_enriched) > 0) {
  n_confirmed   <- sum(llm_enriched$manual_tier == "Confirmed update")
  n_llm_high    <- sum(llm_enriched$manual_tier == "Unclassified" &
                         llm_enriched$pct_influenced >= 30)
  glue::glue(
    "Of **{n_total_fetched}** papers with fetchable DOIs, **{n_analyzed}** had ",
    "meaningful text changes and a matched evaluation. ",
    "**{n_confirmed}** are manually confirmed updates; of the remaining ",
    "{n_analyzed - n_confirmed}, the LLM finds **{n_llm_high}** with ≥30% of changes ",
    "likely driven by evaluator feedback (median across all: **{round(med_pct)}%**)."
  ) |> cat()
}
```

```{r}
#| label: fig-combined-dot
#| fig-cap: "LLM attribution score for the 8 papers with pre/post versions. x-axis: estimated % of major changes attributed to evaluator feedback (Claude Opus 4.6, conservative). Point size scales with % text change. Color distinguishes papers manually confirmed as updates (green) from unclassified papers (blue). Dashed line at 30%."
#| fig-height: 4.2

if (nrow(llm_enriched) > 0) {
  dot_colors <- c(
    "Confirmed update" = "#2D9D5E",
    "Unclassified"     = "#2B7CE9"
  )

  llm_enriched |>
    mutate(paper_short = str_trunc(paper_title, 48)) |>
    ggplot(aes(
      x     = pct_influenced,
      y     = reorder(paper_short, pct_influenced),
      color = manual_tier,
      size  = text_chg_pct
    )) +
    geom_vline(xintercept = 30, linetype = "dashed",
               color = "grey55", linewidth = 0.5) +
    geom_point(alpha = 0.85) +
    scale_color_manual(values = dot_colors, name = "Manual classification") +
    scale_size_continuous(range = c(3, 9), name = "Text change %") +
    scale_x_continuous(limits = c(0, 72),
                       labels = \(x) paste0(x, "%"),
                       breaks = c(0, 15, 30, 50, 70)) +
    labs(x = "% of major changes attributed to evaluator feedback",
         y = NULL) +
    theme_uj() +
    theme(legend.position = "bottom", legend.box = "vertical",
          legend.margin = margin(0, 0, 0, 0))
}
```

```{r}
#| label: fig-combined-overview
#| fig-cap: "Combined evidence distribution across all 57 tracked papers. Papers with LLM attribution results are subdivided by combined tier; the remainder use manual classification only. 'Not fetched' = no NBER/arxiv DOI available."
#| fig-height: 4

if (nrow(llm_enriched) > 0) {
  # LLM-analyzed papers: use combined tier
  llm_summary <- llm_enriched |>
    count(combined_tier) |>
    rename(group = combined_tier) |>
    mutate(group = as.character(group), source = "LLM-analyzed")

  # All other papers: manual classification, excluding those in LLM set
  llm_titles <- llm_enriched$paper_title
  other_summary <- df |>
    filter(!paper_title %in% llm_titles) |>
    mutate(group = case_when(
      adj_status_clean == "Evidence of updating"       ~ "Confirmed update\n(manual only)",
      adj_status_clean == "Stated intention to update" ~ "Stated intention",
      adj_status_clean == "Mixed / minor updating"     ~ "Mixed / minor",
      adj_status_clean == "Unlikely to update"         ~ "Unlikely to update",
      TRUE                                             ~ "Not yet assessed"
    )) |>
    count(group) |>
    mutate(source = "Manual only")

  # Separate bucket for not-fetched (skipped, no DOI)
  n_not_fetched <- llm_df |> filter(skipped) |> nrow()
  if (n_not_fetched > 0) {
    other_summary <- bind_rows(
      other_summary,
      data.frame(group = "No version fetched", n = n_not_fetched, source = "Manual only")
    )
  }

  combined_summary <- bind_rows(llm_summary, other_summary) |>
  mutate(
    group = factor(group, levels = c(
      "Confirmed + strong LLM", "Confirmed + weaker LLM",
      "LLM: likely influenced", "LLM: possibly influenced",
      "LLM: largely independent",
      "Confirmed update\n(manual only)", "Stated intention",
      "Mixed / minor", "Unlikely to update",
      "No version fetched", "Not yet assessed"
    ))
  )

  overview_colors <- c(
    "Confirmed + strong LLM"        = "#1a7a47",
    "Confirmed + weaker LLM"        = "#2D9D5E",
    "LLM: likely influenced"        = "#5AE08A",
    "LLM: possibly influenced"      = "#a8d5b5",
    "LLM: largely independent"      = "#d0e8d8",
    "Confirmed update\n(manual only)" = "#2B7CE9",
    "Stated intention"              = "#7EB8F7",
    "Mixed / minor"                 = "#E8722A",
    "Unlikely to update"            = "#C45A1E",
    "No version fetched"            = "#E2E8F0",
    "Not yet assessed"              = "#CBD5E1"
  )

  ggplot(combined_summary, aes(
    x = n, y = reorder(group, n), fill = group
  )) +
    geom_col(width = 0.65, show.legend = FALSE) +
    geom_text(aes(label = n), hjust = -0.3, size = 3.2) +
    scale_fill_manual(values = overview_colors) +
    scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
    labs(x = "Number of papers", y = NULL,
         title = "Combined evidence: all 57 tracked evaluations") +
    theme_uj() +
    theme(axis.text.y = element_text(size = 9))
}
```

```{r}
#| label: tbl-llm-attribution
#| tbl-cap: "Papers with LLM attribution results, sorted by estimated % of evaluation-driven changes. 'Confirmed update' papers (Bilal, Kremer, Liang) were manually verified and also LLM-analyzed. Signal tier combines both evidence streams."
#| code-summary: "Combined evidence table"

if (nrow(llm_enriched) > 0) {
  tier_bg <- c(
    "Confirmed + strong LLM"   = "#c8e6c9",
    "Confirmed + weaker LLM"   = "#d4edda",
    "LLM: likely influenced"   = "#e8f5e9",
    "LLM: possibly influenced" = "#fff3e0",
    "LLM: largely independent" = "#f8fafc"
  )

  tbl_llm <- llm_enriched |>
    arrange(desc(pct_influenced)) |>
    mutate(
      Title = ifelse(
        !is.na(pubpub_link) & nchar(pubpub_link) > 0,
        paste0('<a href="', pubpub_link, '" target="_blank">',
               str_trunc(paper_title, 58), '</a>'),
        str_trunc(paper_title, 58)
      ),
      `Manual`          = manual_tier,
      `Dep. after`      = ifelse(deposit_after_uj, "yes", "no"),
      `Text Δ%`         = round(text_chg_pct, 1),
      `LLM %`           = paste0(round(pct_influenced), "%"),
      `Conf.`           = paste0(attr_confidence, "/5"),
      `Signal tier`     = combined_tier,
      `LLM narrative`   = str_trunc(narrative, 200)
    ) |>
    select(Title, Manual, `Dep. after`, `Text Δ%`, `LLM %`,
           `Conf.`, `Signal tier`, `LLM narrative`)

  datatable(
    tbl_llm,
    escape   = FALSE,
    rownames = FALSE,
    options  = list(
      pageLength = 15,
      dom        = "lrtip",
      columnDefs = list(
        list(width = "24%", targets = 0),
        list(width = "12%", targets = 1),
        list(width = "6%",  targets = 2),
        list(width = "6%",  targets = 3),
        list(width = "6%",  targets = 4),
        list(width = "5%",  targets = 5),
        list(width = "15%", targets = 6),
        list(width = "26%", targets = 7)
      )
    )
  ) |>
  formatStyle(
    "Signal tier",
    backgroundColor = styleEqual(names(tier_bg), unname(tier_bg))
  )
} else {
  cat("*No LLM attribution results available yet — run scripts/run_paper_change_llm.py first.*")
}
```
