---
title: "Did Authors Respond to Unjournal Evaluations?"
author: "The Unjournal"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

::: {.callout-note}
Manual coding of 57 tracked evaluations, last updated April 2026. Methodology: two data sources combined — (a) manual assessment of author responses and paper revisions by Unjournal staff, and (b) automated PDF-comparison via the [`paper_change_analysis` tool](https://daaronr.github.io/unjournal_tools_interfaces/paper_change_analysis/) which detects line-level changes between pre- and post-evaluation PDF versions.
:::

## Overview

The Unjournal commissions open, structured evaluations of working papers in economics and social science. A natural question is whether this process actually changes research: do authors engage with the feedback and revise their work?

This page tabulates evidence from approximately 57 tracked evaluations based on manual assessment. For each paper where sufficient time has passed, Unjournal staff assessed whether the authors (a) formally or informally responded to the evaluation, and (b) produced an updated version of the paper that reflects the feedback.

For a subset of papers, automated PDF change detection (line-level diffing of before/after versions) supplements the manual coding. The broader background and framing is discussed in [The Unjournal's knowledge base](https://globalimpact.gitbook.io/the-unjournal-project-and-communication-space/benefits-and-features/more-reliable-and-useful-evaluation/author-engagement-evidence).

## Summary

```{r}
#| label: setup
#| code-summary: "Load data and define helpers"

library(dplyr)
library(ggplot2)
library(readr)
library(DT)
library(tidyr)
library(stringr)

df <- read_csv("data/author_adjustment_manual.csv", show_col_types = FALSE)

# Rename columns for easier use
df <- df |>
  rename(
    paper_title     = label_paper_title,
    pubpub_link     = dup_pubpub_final_links,
    adj_status      = `Adjusted_paper?`,
    updated_manual  = `Updated since UJ report -- manual confirmation`,
    deposit_after   = `deposit date > unjournal pub date`,
    research_area   = main_cause_cat_abbrev,
    nb_link         = notebookLM_link,
    pub_status      = publication_status,
    wp_date         = working_paper_release_date,
    uj_pub_date     = publication_date_unjournal
  )

# Define ordered factor for adjustment status
adj_levels <- c(
  "evidence of updating",
  "Stated intention to update",
  "mixed evidence/minor updating",
  "Unlikely to update",
  ""
)

adj_labels <- c(
  "Evidence of updating",
  "Stated intention to update",
  "Mixed / minor updating",
  "Unlikely to update",
  "Not yet assessed"
)

df <- df |>
  mutate(
    adj_status_clean = case_when(
      adj_status == "evidence of updating"        ~ "Evidence of updating",
      adj_status == "Stated intention to update"  ~ "Stated intention to update",
      adj_status == "mixed evidence/minor updating" ~ "Mixed / minor updating",
      adj_status == "Unlikely to update"          ~ "Unlikely to update",
      TRUE                                        ~ "Not yet assessed"
    ),
    adj_status_clean = factor(adj_status_clean, levels = adj_labels),
    author_response_clean = case_when(
      author_response == "Formal response"        ~ "Formal response",
      author_response == "Informal"               ~ "Informal response",
      author_response == "None?"                  ~ "No response",
      TRUE                                        ~ "Not yet coded"
    )
  )

# Color palette aligned with project theme
adj_colors <- c(
  "Evidence of updating"        = "#2D9D5E",
  "Stated intention to update"  = "#5AE08A",
  "Mixed / minor updating"      = "#E8722A",
  "Unlikely to update"          = "#C45A1E",
  "Not yet assessed"            = "#CBD5E1"
)

theme_uj <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      axis.text = element_text(size = base_size * 0.85),
      plot.caption = element_text(size = base_size * 0.75, color = "grey50"),
      legend.position = "right"
    )
}
```

```{r}
#| label: summary-numbers
#| code-summary: "Compute headline statistics"

n_total <- nrow(df)
n_assessed <- df |> filter(adj_status_clean != "Not yet assessed") |> nrow()
n_evidence  <- df |> filter(adj_status_clean == "Evidence of updating") |> nrow()
n_intention <- df |> filter(adj_status_clean == "Stated intention to update") |> nrow()
n_any_positive <- df |>
  filter(adj_status_clean %in% c("Evidence of updating", "Stated intention to update", "Mixed / minor updating")) |>
  nrow()
n_formal <- df |> filter(author_response_clean == "Formal response") |> nrow()
n_any_response <- df |> filter(author_response_clean %in% c("Formal response", "Informal response")) |> nrow()
```

Of **`r n_total` tracked evaluations**:

- **`r n_any_response`** papers received a formal or informal author response
- **`r n_formal`** papers received a **formal written response** to the evaluation
- **`r n_evidence`** papers show **clear evidence of substantive updating** in response to the evaluation
- **`r n_intention`** authors stated intention to update their paper
- **`r n_any_positive`** papers show at least some positive signal (updating, intention, or minor revisions)
- **`r n_assessed`** of `r n_total` have been manually assessed; the remainder are too recent or pending follow-up

```{r}
#| label: fig-adj-bar
#| fig-cap: "Distribution of adjustment outcomes across tracked Unjournal evaluations. 'Not yet assessed' includes papers that are too recent or where follow-up is pending."
#| fig-height: 4

df |>
  count(adj_status_clean) |>
  ggplot(aes(x = reorder(adj_status_clean, -n), y = n, fill = adj_status_clean)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = adj_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Number of papers",
       title = "Author adjustment outcomes following Unjournal evaluations") +
  theme_uj() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

```{r}
#| label: fig-response-bar
#| fig-cap: "Type of author response to Unjournal evaluation."
#| fig-height: 3.5

resp_colors <- c(
  "Formal response"    = "#2B7CE9",
  "Informal response"  = "#5DA3F5",
  "No response"        = "#CBD5E1",
  "Not yet coded"      = "#E2E8F0"
)

df |>
  count(author_response_clean) |>
  ggplot(aes(x = reorder(author_response_clean, -n), y = n, fill = author_response_clean)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = resp_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Number of papers",
       title = "Type of author response to evaluation") +
  theme_uj() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

## Full Tabulation

The table below covers all 57 tracked evaluations. Click column headers to sort; use the search box to filter. Adjustment status is color-coded: green = positive signal, orange = minor/uncertain, grey = unlikely or not yet assessed.

```{r}
#| label: tbl-full
#| code-summary: "Build interactive DT table"

# Build display table
tbl <- df |>
  mutate(
    # Linked paper title
    title_linked = if_else(
      !is.na(pubpub_link) & pubpub_link != "",
      paste0('<a href="', pubpub_link, '" target="_blank">', str_trunc(paper_title, 70), '</a>'),
      str_trunc(paper_title, 70)
    ),
    # Shorten authors
    authors_short = str_trunc(authors, 45),
    # NotebookLM icon link
    nb_icon = if_else(
      !is.na(nb_link) & nb_link != "",
      paste0('<a href="', nb_link, '" target="_blank" title="NotebookLM analysis">&#128214;</a>'),
      ""
    ),
    # DOI link
    doi_linked = if_else(
      !is.na(doi) & doi != "" & !str_detect(doi, "^doi\\.org"),
      paste0('<a href="https://doi.org/', str_remove(doi, "^https://doi\\.org/"), '" target="_blank">DOI</a>'),
      if_else(
        !is.na(doi) & doi != "",
        paste0('<a href="', doi, '" target="_blank">DOI</a>'),
        ""
      )
    )
  ) |>
  select(
    Paper = title_linked,
    Authors = authors_short,
    Area = research_area,
    Response = author_response_clean,
    Adjustment = adj_status_clean,
    `Pub. status` = pub_status,
    NB = nb_icon
  )

# Color rows by adjustment status
adj_bg <- c(
  "Evidence of updating"        = "#d4edda",
  "Stated intention to update"  = "#e8f5e9",
  "Mixed / minor updating"      = "#fff3e0",
  "Unlikely to update"          = "#fce4d4",
  "Not yet assessed"            = "#f8fafc"
)

datatable(
  tbl,
  escape = FALSE,
  rownames = FALSE,
  filter = "top",
  options = list(
    pageLength = 15,
    autoWidth = FALSE,
    columnDefs = list(
      list(width = "35%", targets = 0),
      list(width = "20%", targets = 1),
      list(width = "10%", targets = 2),
      list(width = "10%", targets = 3),
      list(width = "12%", targets = 4),
      list(width = "10%", targets = 5),
      list(width = "3%",  targets = 6)
    ),
    dom = "lfrtip"
  ),
  caption = "All tracked Unjournal evaluations with author response and adjustment data. NB = link to NotebookLM analysis."
) |>
  formatStyle(
    "Adjustment",
    backgroundColor = styleEqual(
      names(adj_bg),
      unname(adj_bg)
    )
  )
```

## Automated Change Detection (Supplement)

The [paper change analysis tool](https://daaronr.github.io/unjournal_tools_interfaces/paper_change_analysis/) provides automated, reproducible detection of line-level changes between pre- and post-evaluation PDF versions. It:

1. Extracts text from "before" (evaluated) and "after" (latest available) PDF versions
2. Computes line-level additions, deletions, and net text change
3. Optionally uses an LLM to attribute specific changes to evaluator suggestions

This supplements the manual coding above by providing objective, quantitative change metrics — particularly useful for papers where subjective judgment of "updating" is difficult.

::: {.callout-note collapse="true"}
## Running the automated pipeline

Results from the automated pipeline will appear here once re-run on current paper pairs. To reproduce:

```bash
# Install dependencies
pip install pandas pdfplumber rapidfuzz

# Run from the paper_change_analysis directory of unjournal_tools_interfaces
python scripts/analyze_paper_changes.py

# Optional: LLM attribution (requires OpenAI key)
python scripts/llm_change_attribution.py
```

See full documentation at [https://daaronr.github.io/unjournal_tools_interfaces/paper_change_analysis/](https://daaronr.github.io/unjournal_tools_interfaces/paper_change_analysis/).
:::

From an earlier run on 34 paper pairs (a subset of the above), the automated analysis found:

- 20 papers had matched Unjournal evaluations (title-based matching from evaluation markdown files)
- 6 papers showed both a matched evaluation and substantial line-level revisions (>50 line changes)
- 1 paper had over 3,000 net line changes post-evaluation

These automated results are consistent with the manual coding above. The discrepancy in totals (34 vs 57) reflects that the manual tracking has since been extended to more papers.
