---
title: "Did Authors Respond to Unjournal Evaluations?"
author: "The Unjournal"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

::: {.callout-note}
57 Unjournal evaluations tracked as of April 2026, drawing on two evidence streams: (a) manual staff assessment of author responses and paper revisions; (b) Claude Opus 4.6 change-attribution for 8 papers where pre/post PDFs were available. See [Combined Evidence](#combined-evidence) below.
:::

## Overview

The Unjournal commissions open, structured evaluations of working papers in economics and social science. A natural question is whether this process changes research: do authors engage with feedback and revise their work?

This page draws on manual classification of 57 tracked evaluations and, for a subset, automated PDF diffing and LLM attribution. Background: [The Unjournal's knowledge base](https://globalimpact.gitbook.io/the-unjournal-project-and-communication-space/benefits-and-features/more-reliable-and-useful-evaluation/author-engagement-evidence).

```{r}
#| label: setup
#| code-summary: "Load data and define helpers"

library(dplyr)
library(ggplot2)
library(readr)
library(DT)
library(tidyr)
library(stringr)
library(glue)
library(jsonlite)

df <- read_csv("data/author_adjustment_manual.csv", show_col_types = FALSE)

df <- df |>
  rename(
    paper_title    = label_paper_title,
    pubpub_link    = dup_pubpub_final_links,
    adj_status     = `Adjusted_paper?`,
    updated_manual = `Updated since UJ report -- manual confirmation`,
    deposit_after  = `deposit date > unjournal pub date`,
    research_area  = main_cause_cat_abbrev,
    nb_link        = notebookLM_link,
    pub_status     = publication_status,
    wp_date        = working_paper_release_date,
    uj_pub_date    = publication_date_unjournal
  )

adj_labels <- c(
  "Evidence of updating", "Stated intention to update",
  "Mixed / minor updating", "Unlikely to update", "Not yet assessed"
)

df <- df |>
  mutate(
    adj_status_clean = case_when(
      adj_status == "evidence of updating"          ~ "Evidence of updating",
      adj_status == "Stated intention to update"    ~ "Stated intention to update",
      adj_status == "mixed evidence/minor updating" ~ "Mixed / minor updating",
      adj_status == "Unlikely to update"            ~ "Unlikely to update",
      TRUE                                          ~ "Not yet assessed"
    ),
    adj_status_clean = factor(adj_status_clean, levels = adj_labels),
    author_response_clean = case_when(
      author_response == "Formal response" ~ "Formal response",
      author_response == "Informal"        ~ "Informal response",
      author_response == "None?"           ~ "No response",
      TRUE                                 ~ "Not yet coded"
    )
  )

theme_uj <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(panel.grid.minor = element_blank(),
          axis.text = element_text(size = base_size * 0.85),
          plot.caption = element_text(size = base_size * 0.75, color = "grey50"),
          legend.position = "right")
}
```

## Combined Evidence of Paper Updating {#combined-evidence}

Two evidence streams are combined: manual classification for all 57 papers, and Claude Opus 4.6 LLM attribution for 8 papers where before/after PDFs were available and meaningful text changes were detected.

::: {.callout-note collapse="true"}
## Pipeline and caveats

**Scripts**: `scripts/fetch_latest_papers.py` → `scripts/run_paper_change_llm.py`

- **Before version**: PDF sent to Unjournal evaluators (`papers/`); **After version**: latest NBER/arxiv version, fetched April 2026
- **Model**: Claude Opus 4.6, temperature 0.2; conservative instructions — only flags "direct" or "indirect" attribution with clear conceptual alignment to a specific evaluator suggestion
- **Coverage**: 30 papers had fetchable DOIs; 8 had meaningful changes (≥15 line edits or ≥0.5% size change) and a matched evaluation file; the 3 manually-confirmed updates were also LLM-analyzed
- **Temporal caveat**: a newer version does not imply changes were evaluation-driven; some papers were already in revision
- **LLM limitations**: text truncated at 60,000 chars/version; figures and tables not visible to the model
:::

```{r}
#| label: llm-attribution-load
#| code-summary: "Load and merge LLM + manual data"

nn <- function(x, default = NA) if (!is.null(x)) x else default

llm_results_path <- "data/paper_change_llm_results.json"

# Signal tier labels used throughout — define once
tier_levels <- c(
  "Confirmed, LLM \u226540%",
  "Confirmed, LLM <40%",
  "LLM \u226530% (not confirmed)",
  "LLM 10\u201329%",
  "LLM <10%"
)

tier_colors_llm <- c(
  "Confirmed, LLM \u226540%"      = "#1a7a47",
  "Confirmed, LLM <40%"           = "#2D9D5E",
  "LLM \u226530% (not confirmed)" = "#5AE08A",
  "LLM 10\u201329%"               = "#a8d5b5",
  "LLM <10%"                      = "#d0e8d8"
)

if (file.exists(llm_results_path)) {
  llm_raw <- fromJSON(llm_results_path, simplifyVector = FALSE)

  llm_df <- lapply(llm_raw, function(r) {
    oa <- r$overall_assessment
    data.frame(
      paper_title      = nn(r$paper_title, ""),
      adj_status_csv   = nn(r$adj_status, ""),
      deposit_after_uj = isTRUE(r$deposit_after_uj),
      text_chg_pct     = nn(r$text_length_change_pct, NA_real_),
      additions        = nn(r$additions_count, NA_integer_),
      deletions        = nn(r$deletions_count, NA_integer_),
      n_major_changes  = length(nn(r$major_changes, list())),
      n_suggestions    = length(nn(r$evaluator_suggestions, list())),
      pct_influenced   = if (!is.null(oa)) nn(oa$pct_likely_influenced, NA_real_) else NA_real_,
      attr_confidence  = if (!is.null(oa)) nn(oa$confidence, NA_real_) else NA_real_,
      narrative        = if (!is.null(oa)) nn(oa$narrative, "") else "",
      eval_found       = !is.null(r$eval_file) && nchar(nn(r$eval_file, "")) > 0,
      skipped          = !is.null(r$skipped_reason),
      stringsAsFactors = FALSE
    )
  }) |> bind_rows()

  llm_analyzed <- llm_df |>
    filter(!skipped & eval_found & n_major_changes > 0 & !is.na(pct_influenced))

  llm_enriched <- llm_analyzed |>
    left_join(df |> select(paper_title, research_area, pubpub_link), by = "paper_title") |>
    mutate(
      manual_tier = case_when(
        adj_status_csv == "evidence of updating"          ~ "Confirmed update",
        adj_status_csv == "Stated intention to update"    ~ "Stated intention",
        adj_status_csv == "mixed evidence/minor updating" ~ "Mixed / minor",
        adj_status_csv == "Unlikely to update"            ~ "Unlikely",
        TRUE                                              ~ "Unclassified"
      ),
      combined_tier = case_when(
        manual_tier == "Confirmed update" & pct_influenced >= 40 ~ tier_levels[1],
        manual_tier == "Confirmed update"                        ~ tier_levels[2],
        pct_influenced >= 30                                     ~ tier_levels[3],
        pct_influenced >= 10                                     ~ tier_levels[4],
        TRUE                                                     ~ tier_levels[5]
      ),
      combined_tier = factor(combined_tier, levels = tier_levels)
    )

  n_analyzed      <- nrow(llm_enriched)
  n_total_fetched <- nrow(llm_df)
  med_pct         <- median(llm_enriched$pct_influenced, na.rm = TRUE)
  n_high          <- sum(llm_enriched$pct_influenced >= 30, na.rm = TRUE)
} else {
  llm_df <- llm_analyzed <- llm_enriched <- data.frame()
  n_analyzed <- n_total_fetched <- n_high <- 0
  med_pct <- NA_real_
}
```

```{r}
#| label: llm-attribution-summary
#| results: asis
#| code-summary: "Summary"

if (nrow(llm_enriched) > 0) {
  n_confirmed <- sum(llm_enriched$manual_tier == "Confirmed update")
  n_llm_high  <- sum(llm_enriched$manual_tier == "Unclassified" &
                       llm_enriched$pct_influenced >= 30)
  cat(glue::glue(
    "Of {n_total_fetched} papers with fetchable DOIs, {n_analyzed} had meaningful text ",
    "changes and a matched evaluation. {n_confirmed} are manually confirmed updates; ",
    "of the remaining {n_analyzed - n_confirmed}, the LLM finds {n_llm_high} with \u226530% ",
    "of changes likely driven by evaluator feedback (median across all: {round(med_pct)}%)."
  ))
}
```

```{r}
#| label: fig-combined-waffle
#| fig-cap: "All 57 tracked papers by combined evidence tier (each square = 1 paper). Green tones: papers where LLM analysis was run, shaded by attribution score and manual confirmation status. Blue: manually confirmed updates without LLM analysis. Orange/grey: stated intention, weak signal, or not yet assessed."
#| fig-height: 4.5

if (nrow(llm_enriched) > 0) {
  # Build one row per paper with tier assignment
  waffle_llm <- llm_enriched |>
    select(paper_title, combined_tier) |>
    mutate(group = as.character(combined_tier))

  waffle_manual <- df |>
    filter(!paper_title %in% llm_enriched$paper_title) |>
    mutate(group = case_when(
      adj_status_clean == "Evidence of updating"       ~ "Confirmed (manual only)",
      adj_status_clean == "Stated intention to update" ~ "Stated intention",
      adj_status_clean == "Mixed / minor updating"     ~ "Mixed / minor",
      adj_status_clean == "Unlikely to update"         ~ "Unlikely",
      TRUE                                             ~ "Not yet assessed"
    )) |>
    select(paper_title, group)

  waffle_all_groups <- c(tier_levels,
    "Confirmed (manual only)", "Stated intention",
    "Mixed / minor", "Unlikely", "Not yet assessed")

  waffle_colors <- c(
    tier_colors_llm,
    "Confirmed (manual only)" = "#2B7CE9",
    "Stated intention"        = "#7EB8F7",
    "Mixed / minor"           = "#E8722A",
    "Unlikely"                = "#C45A1E",
    "Not yet assessed"        = "#CBD5E1"
  )

  waffle_data <- bind_rows(waffle_llm, waffle_manual) |>
    mutate(group = factor(group, levels = waffle_all_groups)) |>
    arrange(group) |>
    mutate(rank = row_number(),
           x = (rank - 1) %% 9,
           y = -((rank - 1) %/% 9))

  ggplot(waffle_data, aes(x = x, y = y, fill = group)) +
    geom_tile(color = "white", linewidth = 1.5) +
    coord_equal() +
    scale_fill_manual(values = waffle_colors, name = NULL,
                      drop = FALSE) +
    theme_void() +
    theme(legend.position = "right",
          legend.text = element_text(size = 9),
          plot.title = element_text(size = 11, face = "plain",
                                    margin = margin(b = 6))) +
    labs(title = "All 57 tracked papers — each square = 1 paper")
}
```

```{r}
#| label: fig-combined-dot
#| fig-cap: "LLM attribution for the 8 papers with pre/post versions compared. x-axis: estimated % of major changes attributed to evaluator feedback (Claude Opus 4.6, conservative). Point size scales with text change magnitude. Green = manually confirmed update; blue = not yet manually classified. Dashed line at 30%."
#| fig-height: 3.8

if (nrow(llm_enriched) > 0) {
  dot_colors <- c("Confirmed update" = "#2D9D5E", "Unclassified" = "#2B7CE9")

  llm_enriched |>
    mutate(paper_short = str_trunc(paper_title, 46)) |>
    ggplot(aes(x = pct_influenced, y = reorder(paper_short, pct_influenced),
               color = manual_tier, size = text_chg_pct)) +
    geom_vline(xintercept = 30, linetype = "dashed", color = "grey55", linewidth = 0.5) +
    geom_point(alpha = 0.85) +
    scale_color_manual(values = dot_colors, name = "Manual classification") +
    scale_size_continuous(range = c(3, 9), name = "Text change %") +
    scale_x_continuous(limits = c(0, 72), labels = \(x) paste0(x, "%"),
                       breaks = c(0, 15, 30, 50, 70)) +
    labs(x = "% of major changes attributed to evaluator feedback", y = NULL) +
    theme_uj() +
    theme(legend.position = "bottom", legend.box = "vertical",
          legend.margin = margin(0, 0, 0, 0))
}
```

```{r}
#| label: tbl-llm-attribution
#| tbl-cap: "Papers with LLM attribution results, sorted by % of changes attributed to evaluator feedback. 'Confirmed' = manually verified by Unjournal staff. Signal tier: 'Confirmed, LLM ≥40%' = manually confirmed AND LLM finds strong corroboration; 'Confirmed, LLM <40%' = confirmed but weaker LLM signal; 'LLM ≥30% (not confirmed)' = unconfirmed but LLM finds likely influence."
#| code-summary: "Combined evidence table"

if (nrow(llm_enriched) > 0) {
  tier_bg <- c(
    "Confirmed, LLM \u226540%"      = "#c8e6c9",
    "Confirmed, LLM <40%"           = "#d4edda",
    "LLM \u226530% (not confirmed)" = "#e8f5e9",
    "LLM 10\u201329%"               = "#fff3e0",
    "LLM <10%"                      = "#f8fafc"
  )

  tbl_llm <- llm_enriched |>
    arrange(desc(pct_influenced)) |>
    mutate(
      Title = ifelse(
        !is.na(pubpub_link) & nchar(pubpub_link) > 0,
        paste0('<a href="', pubpub_link, '" target="_blank">',
               str_trunc(paper_title, 60), '</a>'),
        str_trunc(paper_title, 60)
      ),
      `Manual`      = manual_tier,
      `LLM %`       = paste0(round(pct_influenced), "%"),
      `Conf.`       = paste0(attr_confidence, "/5"),
      `Signal tier` = combined_tier,
      `LLM narrative` = str_trunc(narrative, 350)
    ) |>
    select(Title, Manual, `LLM %`, Conf., `Signal tier`, `LLM narrative`)

  datatable(
    tbl_llm,
    escape   = FALSE,
    rownames = FALSE,
    options  = list(
      pageLength = 5,
      dom        = "tip",
      scrollX    = TRUE,
      columnDefs = list(
        list(width = "18%", targets = 0),
        list(width = "11%", targets = 1),
        list(width = "6%",  targets = 2),
        list(width = "5%",  targets = 3),
        list(width = "16%", targets = 4),
        list(width = "44%", targets = 5)
      )
    )
  ) |>
  formatStyle("Signal tier",
    backgroundColor = styleEqual(names(tier_bg), unname(tier_bg)))
} else {
  cat("*No LLM attribution results available yet.*")
}
```

## Summary Statistics

```{r}
#| label: summary-numbers
#| code-summary: "Compute headline statistics"

n_total      <- nrow(df)
n_assessed   <- df |> filter(adj_status_clean != "Not yet assessed") |> nrow()
n_evidence   <- df |> filter(adj_status_clean == "Evidence of updating") |> nrow()
n_intention  <- df |> filter(adj_status_clean == "Stated intention to update") |> nrow()
n_any_positive <- df |>
  filter(adj_status_clean %in% c("Evidence of updating",
                                  "Stated intention to update",
                                  "Mixed / minor updating")) |> nrow()
n_formal  <- df |> filter(author_response_clean == "Formal response") |> nrow()
n_any_resp <- df |>
  filter(author_response_clean %in% c("Formal response", "Informal response")) |> nrow()
```

Of `r n_total` tracked evaluations, `r n_assessed` have been manually assessed. Among these: `r n_evidence` show clear evidence of substantive updating, `r n_intention` authors stated intention to update, and `r n_any_positive` show at least some positive signal. On engagement: `r n_any_resp` papers received a formal or informal author response, of which `r n_formal` were formal written responses.

```{r}
#| label: fig-response-bar
#| fig-cap: "Type of author response to Unjournal evaluation. 'Not yet coded' indicates the response status has not been assessed for those papers (37 blanks in data); this is distinct from confirmed non-response (only 1 paper is explicitly coded as no response)."
#| fig-height: 3.2

resp_colors <- c(
  "Formal response"   = "#2B7CE9",
  "Informal response" = "#5DA3F5",
  "No response"       = "#CBD5E1",
  "Not yet coded"     = "#E2E8F0"
)

df |>
  count(author_response_clean) |>
  ggplot(aes(x = reorder(author_response_clean, -n), y = n,
             fill = author_response_clean)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = resp_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Number of papers",
       title = "Type of author response to evaluation") +
  theme_uj() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

## Full Tabulation

::: {.callout-note collapse="true"}
## All 57 tracked evaluations

Click column headers to sort; use the search box to filter. Adjustment status is color-coded: green = positive signal, orange = minor/uncertain, grey = unlikely or not yet assessed.

```{r}
#| label: tbl-full
#| code-summary: "Build interactive DT table"

tbl <- df |>
  mutate(
    title_linked = if_else(
      !is.na(pubpub_link) & pubpub_link != "",
      paste0('<a href="', pubpub_link, '" target="_blank">',
             str_trunc(paper_title, 72), '</a>'),
      str_trunc(paper_title, 72)
    ),
    authors_short = str_trunc(authors, 40),
    nb_icon = if_else(
      !is.na(nb_link) & nb_link != "",
      paste0('<a href="', nb_link, '" target="_blank" title="NotebookLM">&#128214;</a>'),
      ""
    )
  ) |>
  select(Paper = title_linked, Authors = authors_short,
         Area = research_area, Response = author_response_clean,
         Adjustment = adj_status_clean, `Pub.` = pub_status, NB = nb_icon)

adj_bg <- c(
  "Evidence of updating"        = "#d4edda",
  "Stated intention to update"  = "#e8f5e9",
  "Mixed / minor updating"      = "#fff3e0",
  "Unlikely to update"          = "#fce4d4",
  "Not yet assessed"            = "#f8fafc"
)

datatable(tbl, escape = FALSE, rownames = FALSE, filter = "top",
  options = list(
    pageLength = 8, autoWidth = FALSE, scrollX = TRUE, dom = "lfrtip",
    columnDefs = list(
      list(width = "36%", targets = 0),
      list(width = "18%", targets = 1),
      list(width = "8%",  targets = 2),
      list(width = "11%", targets = 3),
      list(width = "16%", targets = 4),
      list(width = "8%",  targets = 5),
      list(width = "3%",  targets = 6)
    )
  ),
  caption = "All tracked Unjournal evaluations. NB = NotebookLM analysis link."
) |>
  formatStyle("Adjustment",
    backgroundColor = styleEqual(names(adj_bg), unname(adj_bg)))
```

:::
