---
title: "Did Authors Respond to Unjournal Evaluations?"
author: "The Unjournal"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

::: {.callout-note}
Manual coding of 57 tracked evaluations, last updated April 2026. Two data sources: (a) manual assessment of author responses and paper revisions by Unjournal staff; (b) GPT-5 Pro change-attribution analysis comparing pre-evaluation PDFs against current NBER/arxiv versions, assessing whether paper updates reflect evaluator suggestions (see [LLM Change Attribution Analysis](#llm-change-attribution-analysis) below).
:::

## Overview

The Unjournal commissions open, structured evaluations of working papers in economics and social science. A natural question is whether this process actually changes research: do authors engage with the feedback and revise their work?

This page tabulates evidence from approximately 57 tracked evaluations based on manual assessment. For each paper where sufficient time has passed, Unjournal staff assessed whether the authors (a) formally or informally responded to the evaluation, and (b) produced an updated version of the paper that reflects the feedback.

For a subset of papers, automated PDF change detection (line-level diffing of before/after versions) supplements the manual coding. The broader background and framing is discussed in [The Unjournal's knowledge base](https://globalimpact.gitbook.io/the-unjournal-project-and-communication-space/benefits-and-features/more-reliable-and-useful-evaluation/author-engagement-evidence).

## Summary

```{r}
#| label: setup
#| code-summary: "Load data and define helpers"

library(dplyr)
library(ggplot2)
library(readr)
library(DT)
library(tidyr)
library(stringr)
library(glue)
library(jsonlite)

df <- read_csv("data/author_adjustment_manual.csv", show_col_types = FALSE)

# Rename columns for easier use
df <- df |>
  rename(
    paper_title     = label_paper_title,
    pubpub_link     = dup_pubpub_final_links,
    adj_status      = `Adjusted_paper?`,
    updated_manual  = `Updated since UJ report -- manual confirmation`,
    deposit_after   = `deposit date > unjournal pub date`,
    research_area   = main_cause_cat_abbrev,
    nb_link         = notebookLM_link,
    pub_status      = publication_status,
    wp_date         = working_paper_release_date,
    uj_pub_date     = publication_date_unjournal
  )

# Define ordered factor for adjustment status
adj_levels <- c(
  "evidence of updating",
  "Stated intention to update",
  "mixed evidence/minor updating",
  "Unlikely to update",
  ""
)

adj_labels <- c(
  "Evidence of updating",
  "Stated intention to update",
  "Mixed / minor updating",
  "Unlikely to update",
  "Not yet assessed"
)

df <- df |>
  mutate(
    adj_status_clean = case_when(
      adj_status == "evidence of updating"        ~ "Evidence of updating",
      adj_status == "Stated intention to update"  ~ "Stated intention to update",
      adj_status == "mixed evidence/minor updating" ~ "Mixed / minor updating",
      adj_status == "Unlikely to update"          ~ "Unlikely to update",
      TRUE                                        ~ "Not yet assessed"
    ),
    adj_status_clean = factor(adj_status_clean, levels = adj_labels),
    author_response_clean = case_when(
      author_response == "Formal response"        ~ "Formal response",
      author_response == "Informal"               ~ "Informal response",
      author_response == "None?"                  ~ "No response",
      TRUE                                        ~ "Not yet coded"
    )
  )

# Color palette aligned with project theme
adj_colors <- c(
  "Evidence of updating"        = "#2D9D5E",
  "Stated intention to update"  = "#5AE08A",
  "Mixed / minor updating"      = "#E8722A",
  "Unlikely to update"          = "#C45A1E",
  "Not yet assessed"            = "#CBD5E1"
)

theme_uj <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      axis.text = element_text(size = base_size * 0.85),
      plot.caption = element_text(size = base_size * 0.75, color = "grey50"),
      legend.position = "right"
    )
}
```

```{r}
#| label: summary-numbers
#| code-summary: "Compute headline statistics"

n_total <- nrow(df)
n_assessed <- df |> filter(adj_status_clean != "Not yet assessed") |> nrow()
n_evidence  <- df |> filter(adj_status_clean == "Evidence of updating") |> nrow()
n_intention <- df |> filter(adj_status_clean == "Stated intention to update") |> nrow()
n_any_positive <- df |>
  filter(adj_status_clean %in% c("Evidence of updating", "Stated intention to update", "Mixed / minor updating")) |>
  nrow()
n_formal <- df |> filter(author_response_clean == "Formal response") |> nrow()
n_any_response <- df |> filter(author_response_clean %in% c("Formal response", "Informal response")) |> nrow()
```

Of **`r n_total` tracked evaluations**:

- **`r n_any_response`** papers received a formal or informal author response
- **`r n_formal`** papers received a **formal written response** to the evaluation
- **`r n_evidence`** papers show **clear evidence of substantive updating** in response to the evaluation
- **`r n_intention`** authors stated intention to update their paper
- **`r n_any_positive`** papers show at least some positive signal (updating, intention, or minor revisions)
- **`r n_assessed`** of `r n_total` have been manually assessed; the remainder are too recent or pending follow-up

```{r}
#| label: fig-adj-bar
#| fig-cap: "Distribution of adjustment outcomes across tracked Unjournal evaluations. 'Not yet assessed' includes papers that are too recent or where follow-up is pending."
#| fig-height: 4

df |>
  count(adj_status_clean) |>
  ggplot(aes(x = reorder(adj_status_clean, -n), y = n, fill = adj_status_clean)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = adj_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Number of papers",
       title = "Author adjustment outcomes following Unjournal evaluations") +
  theme_uj() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

```{r}
#| label: fig-response-bar
#| fig-cap: "Type of author response to Unjournal evaluation."
#| fig-height: 3.5

resp_colors <- c(
  "Formal response"    = "#2B7CE9",
  "Informal response"  = "#5DA3F5",
  "No response"        = "#CBD5E1",
  "Not yet coded"      = "#E2E8F0"
)

df |>
  count(author_response_clean) |>
  ggplot(aes(x = reorder(author_response_clean, -n), y = n, fill = author_response_clean)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = resp_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Number of papers",
       title = "Type of author response to evaluation") +
  theme_uj() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

## Full Tabulation

The table below covers all 57 tracked evaluations. Click column headers to sort; use the search box to filter. Adjustment status is color-coded: green = positive signal, orange = minor/uncertain, grey = unlikely or not yet assessed.

```{r}
#| label: tbl-full
#| code-summary: "Build interactive DT table"

# Build display table
tbl <- df |>
  mutate(
    # Linked paper title
    title_linked = if_else(
      !is.na(pubpub_link) & pubpub_link != "",
      paste0('<a href="', pubpub_link, '" target="_blank">', str_trunc(paper_title, 70), '</a>'),
      str_trunc(paper_title, 70)
    ),
    # Shorten authors
    authors_short = str_trunc(authors, 45),
    # NotebookLM icon link
    nb_icon = if_else(
      !is.na(nb_link) & nb_link != "",
      paste0('<a href="', nb_link, '" target="_blank" title="NotebookLM analysis">&#128214;</a>'),
      ""
    ),
    # DOI link
    doi_linked = if_else(
      !is.na(doi) & doi != "" & !str_detect(doi, "^doi\\.org"),
      paste0('<a href="https://doi.org/', str_remove(doi, "^https://doi\\.org/"), '" target="_blank">DOI</a>'),
      if_else(
        !is.na(doi) & doi != "",
        paste0('<a href="', doi, '" target="_blank">DOI</a>'),
        ""
      )
    )
  ) |>
  select(
    Paper = title_linked,
    Authors = authors_short,
    Area = research_area,
    Response = author_response_clean,
    Adjustment = adj_status_clean,
    `Pub. status` = pub_status,
    NB = nb_icon
  )

# Color rows by adjustment status
adj_bg <- c(
  "Evidence of updating"        = "#d4edda",
  "Stated intention to update"  = "#e8f5e9",
  "Mixed / minor updating"      = "#fff3e0",
  "Unlikely to update"          = "#fce4d4",
  "Not yet assessed"            = "#f8fafc"
)

datatable(
  tbl,
  escape = FALSE,
  rownames = FALSE,
  filter = "top",
  options = list(
    pageLength = 15,
    autoWidth = FALSE,
    columnDefs = list(
      list(width = "35%", targets = 0),
      list(width = "20%", targets = 1),
      list(width = "10%", targets = 2),
      list(width = "10%", targets = 3),
      list(width = "12%", targets = 4),
      list(width = "10%", targets = 5),
      list(width = "3%",  targets = 6)
    ),
    dom = "lfrtip"
  ),
  caption = "All tracked Unjournal evaluations with author response and adjustment data. NB = link to NotebookLM analysis."
) |>
  formatStyle(
    "Adjustment",
    backgroundColor = styleEqual(
      names(adj_bg),
      unname(adj_bg)
    )
  )
```

## LLM Change Attribution Analysis

For papers where a newer version is publicly available (via NBER or arxiv), we compare the pre-evaluation version (the PDF sent to evaluators) against the current version, then use GPT-5 Pro to assess whether the changes plausibly reflect the evaluators' suggestions.

**Pipeline** (scripts in `scripts/`, run from project root):
1. `fetch_latest_papers.py` — downloads latest PDFs from NBER/arxiv for all papers with qualifying DOIs
2. `run_paper_change_llm.py` — extracts text, computes diffs, runs a 3-step LLM analysis per paper: (a) identify major changes, (b) extract evaluator suggestions from the evaluation markdown, (c) assess whether each change aligns with a suggestion

::: {.callout-note collapse="true"}
## Methodology and caveats

- **Before version**: the PDF evaluated by Unjournal evaluators (stored in `papers/`)
- **After version**: latest version fetched from NBER or arxiv in April 2026
- **Attribution conservatism**: GPT-5 Pro is instructed to mark changes as "direct" or "indirect" only when there is clear conceptual alignment with a specific evaluator suggestion; coincidental improvements are labelled "possible" or "unrelated"
- **Temporal caveat**: a newer PDF does not guarantee the change happened *in response* to the evaluation; some papers were already in revision. The `deposit date > UJ pub date` flag in the manual coding helps but is not definitive
- **Coverage**: 30 papers had fetchable DOIs (NBER or arxiv); of these, 11 showed meaningful text changes (≥15 line edits or ≥0.5% size change); 3 papers with no matching pre-evaluation PDF were excluded
- **LLM limitations**: the model sees truncated text (60,000 chars per version) and may miss changes in heavily mathematical or figure-dense sections
:::

```{r}
#| label: llm-attribution-load
#| code-summary: "Load LLM attribution results"

nn <- function(x, default = NA) if (!is.null(x)) x else default

llm_results_path <- "data/paper_change_llm_results.json"

if (file.exists(llm_results_path)) {
  llm_raw <- fromJSON(llm_results_path, simplifyVector = FALSE)

  llm_df <- lapply(llm_raw, function(r) {
    oa <- r$overall_assessment
    data.frame(
      paper_title      = nn(r$paper_title, ""),
      adj_status       = nn(r$adj_status, ""),
      deposit_after_uj = isTRUE(r$deposit_after_uj),
      text_chg_pct     = nn(r$text_length_change_pct, NA_real_),
      additions        = nn(r$additions_count, NA_integer_),
      deletions        = nn(r$deletions_count, NA_integer_),
      n_major_changes  = length(nn(r$major_changes, list())),
      n_suggestions    = length(nn(r$evaluator_suggestions, list())),
      pct_influenced   = if (!is.null(oa)) nn(oa$pct_likely_influenced, NA_real_) else NA_real_,
      attr_confidence  = if (!is.null(oa)) nn(oa$confidence, NA_real_) else NA_real_,
      narrative        = if (!is.null(oa)) nn(oa$narrative, "") else "",
      eval_found       = !is.null(r$eval_file) && nchar(nn(r$eval_file, "")) > 0,
      skipped          = !is.null(r$skipped_reason),
      stringsAsFactors = FALSE
    )
  }) |> bind_rows()

  llm_analyzed    <- llm_df |> filter(!skipped & eval_found & n_major_changes > 0)
  n_analyzed      <- nrow(llm_analyzed)
  n_total_fetched <- nrow(llm_df)
  med_pct         <- median(llm_analyzed$pct_influenced, na.rm = TRUE)
  n_high          <- sum(!is.na(llm_analyzed$pct_influenced) &
                           llm_analyzed$pct_influenced >= 30, na.rm = TRUE)
} else {
  message("LLM results not found — run scripts/run_paper_change_llm.py first")
  llm_df <- llm_analyzed <- data.frame()
  n_analyzed <- n_total_fetched <- n_high <- 0
  med_pct <- NA_real_
}
```

```{r}
#| label: llm-attribution-summary
#| code-summary: "Summary statistics"

if (nrow(llm_analyzed) > 0) {
  cat(glue::glue(
    "Of **{n_total_fetched}** papers with fetchable DOIs (NBER/arxiv), ",
    "**{n_analyzed}** showed meaningful text changes and had a matched evaluation. ",
    "For these, GPT-5 Pro estimated a median of **{round(med_pct)}%** of major changes ",
    "as likely influenced by evaluator suggestions (confidence ≥ 1/5). ",
    "**{n_high}** papers had ≥30% of changes attributed to the evaluation."
  ))
}
```

```{r}
#| label: tbl-llm-attribution
#| code-summary: "Build attribution table"

if (nrow(llm_analyzed) > 0) {
  tbl_llm <- llm_analyzed |>
    mutate(
      Title       = str_trunc(paper_title, 65),
      `Adj. status` = adj_status,
      `Deposit after UJ` = ifelse(deposit_after_uj, "yes", "no"),
      `Text Δ%`   = round(text_chg_pct, 1),
      `Adds/Dels` = paste0(additions, "/", deletions),
      `Major changes` = n_major_changes,
      `Eval. suggestions` = n_suggestions,
      `% attr. to eval` = ifelse(!is.na(pct_influenced), paste0(round(pct_influenced), "%"), "—"),
      `Confidence (1–5)` = ifelse(!is.na(attr_confidence), attr_confidence, "—"),
      `LLM narrative` = str_trunc(narrative, 180)
    ) |>
    select(Title, `Adj. status`, `Deposit after UJ`, `Text Δ%`, `Adds/Dels`,
           `Major changes`, `Eval. suggestions`, `% attr. to eval`,
           `Confidence (1–5)`, `LLM narrative`)

  # Color coding by % attributed
  pct_vals <- llm_analyzed$pct_influenced
  pct_bg <- case_when(
    is.na(pct_vals)      ~ "#f8fafc",
    pct_vals >= 50       ~ "#d4edda",
    pct_vals >= 30       ~ "#fff3e0",
    TRUE                 ~ "#f8fafc"
  )

  datatable(
    tbl_llm,
    escape    = FALSE,
    rownames  = FALSE,
    options   = list(
      pageLength = 15,
      dom        = "lrtip",
      columnDefs = list(
        list(width = "28%", targets = 0),
        list(width = "12%", targets = 1),
        list(width = "8%",  targets = 2),
        list(width = "6%",  targets = 3),
        list(width = "8%",  targets = 4),
        list(width = "6%",  targets = 5),
        list(width = "8%",  targets = 6),
        list(width = "7%",  targets = 7),
        list(width = "7%",  targets = 8),
        list(width = "10%", targets = 9)
      )
    ),
    caption = "GPT-5 Pro attribution analysis: do paper changes reflect evaluator suggestions? Papers without measurable text changes are excluded."
  ) |>
    formatStyle(
      "% attr. to eval",
      backgroundColor = styleEqual(
        paste0(round(pct_vals[!is.na(pct_vals)]), "%"),
        pct_bg[!is.na(pct_vals)]
      )
    )
} else {
  cat("*No LLM attribution results available yet.*")
}
```
