---
title: "Did Authors Respond to Unjournal Evaluations?"
author: "The Unjournal"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

::: {.callout-note}
57 Unjournal evaluations tracked as of April 2026, drawing on two evidence streams: (a) manual staff assessment of author responses and paper revisions; (b) Claude Opus 4.6 change-attribution analysis for 8 papers where pre/post PDFs were available, assessing whether updates reflect evaluator suggestions. See [Combined Evidence](#combined-evidence) below.
:::

## Overview

The Unjournal commissions open, structured evaluations of working papers in economics and social science. A natural question is whether this process changes research: do authors engage with feedback and revise their work?

This page tabulates evidence from 57 tracked evaluations. For each paper where sufficient time has passed, Unjournal staff assessed whether authors (a) formally or informally responded to the evaluation, and (b) produced an updated version reflecting the feedback. For a subset, automated PDF change detection and LLM attribution supplements the manual coding. Background and framing: [The Unjournal's knowledge base](https://globalimpact.gitbook.io/the-unjournal-project-and-communication-space/benefits-and-features/more-reliable-and-useful-evaluation/author-engagement-evidence).

```{r}
#| label: setup
#| code-summary: "Load data and define helpers"

library(dplyr)
library(ggplot2)
library(readr)
library(DT)
library(tidyr)
library(stringr)
library(glue)
library(jsonlite)

df <- read_csv("data/author_adjustment_manual.csv", show_col_types = FALSE)

df <- df |>
  rename(
    paper_title     = label_paper_title,
    pubpub_link     = dup_pubpub_final_links,
    adj_status      = `Adjusted_paper?`,
    updated_manual  = `Updated since UJ report -- manual confirmation`,
    deposit_after   = `deposit date > unjournal pub date`,
    research_area   = main_cause_cat_abbrev,
    nb_link         = notebookLM_link,
    pub_status      = publication_status,
    wp_date         = working_paper_release_date,
    uj_pub_date     = publication_date_unjournal
  )

adj_labels <- c(
  "Evidence of updating",
  "Stated intention to update",
  "Mixed / minor updating",
  "Unlikely to update",
  "Not yet assessed"
)

df <- df |>
  mutate(
    adj_status_clean = case_when(
      adj_status == "evidence of updating"          ~ "Evidence of updating",
      adj_status == "Stated intention to update"    ~ "Stated intention to update",
      adj_status == "mixed evidence/minor updating" ~ "Mixed / minor updating",
      adj_status == "Unlikely to update"            ~ "Unlikely to update",
      TRUE                                          ~ "Not yet assessed"
    ),
    adj_status_clean = factor(adj_status_clean, levels = adj_labels),
    author_response_clean = case_when(
      author_response == "Formal response" ~ "Formal response",
      author_response == "Informal"        ~ "Informal response",
      author_response == "None?"           ~ "No response",
      TRUE                                 ~ "Not yet coded"
    )
  )

adj_colors <- c(
  "Evidence of updating"        = "#2D9D5E",
  "Stated intention to update"  = "#5AE08A",
  "Mixed / minor updating"      = "#E8722A",
  "Unlikely to update"          = "#C45A1E",
  "Not yet assessed"            = "#CBD5E1"
)

theme_uj <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      axis.text = element_text(size = base_size * 0.85),
      plot.caption = element_text(size = base_size * 0.75, color = "grey50"),
      legend.position = "right"
    )
}
```

## Combined Evidence of Paper Updating {#combined-evidence}

Two evidence streams are combined: manual classification for all 57 papers, and Claude Opus 4.6 LLM attribution for the 8 papers where before/after PDFs were available and meaningful text changes were detected.

::: {.callout-note collapse="true"}
## Pipeline and caveats

**Scripts** (run from project root): `scripts/fetch_latest_papers.py` → `scripts/run_paper_change_llm.py`

- **Before version**: PDF evaluated by Unjournal evaluators (`papers/`); **After version**: latest from NBER/arxiv, fetched April 2026
- **Model**: Claude Opus 4.6, temperature 0.2; instructed to be conservative — only "direct" or "indirect" attribution when there is clear conceptual alignment with a specific evaluator suggestion
- **Coverage**: 30 papers had fetchable DOIs; 8 showed meaningful text changes (≥15 line edits or ≥0.5% size change) and matched evaluation files; the 3 manually-confirmed updates were also LLM-analyzed using `--paper`
- **Temporal caveat**: a newer PDF does not mean changes happened in response to the evaluation; some papers were already in revision. The `deposit date > UJ pub date` flag helps but is not definitive
- **LLM limitations**: text truncated at 60,000 chars per version; may miss changes in mathematical or figure-dense sections
:::

```{r}
#| label: llm-attribution-load
#| code-summary: "Load and merge LLM + manual data"

nn <- function(x, default = NA) if (!is.null(x)) x else default

llm_results_path <- "data/paper_change_llm_results.json"

if (file.exists(llm_results_path)) {
  llm_raw <- fromJSON(llm_results_path, simplifyVector = FALSE)

  llm_df <- lapply(llm_raw, function(r) {
    oa <- r$overall_assessment
    data.frame(
      paper_title      = nn(r$paper_title, ""),
      adj_status_csv   = nn(r$adj_status, ""),
      deposit_after_uj = isTRUE(r$deposit_after_uj),
      text_chg_pct     = nn(r$text_length_change_pct, NA_real_),
      additions        = nn(r$additions_count, NA_integer_),
      deletions        = nn(r$deletions_count, NA_integer_),
      n_major_changes  = length(nn(r$major_changes, list())),
      n_suggestions    = length(nn(r$evaluator_suggestions, list())),
      pct_influenced   = if (!is.null(oa)) nn(oa$pct_likely_influenced, NA_real_) else NA_real_,
      attr_confidence  = if (!is.null(oa)) nn(oa$confidence, NA_real_) else NA_real_,
      narrative        = if (!is.null(oa)) nn(oa$narrative, "") else "",
      eval_found       = !is.null(r$eval_file) && nchar(nn(r$eval_file, "")) > 0,
      skipped          = !is.null(r$skipped_reason),
      stringsAsFactors = FALSE
    )
  }) |> bind_rows()

  llm_analyzed <- llm_df |>
    filter(!skipped & eval_found & n_major_changes > 0 & !is.na(pct_influenced))

  llm_enriched <- llm_analyzed |>
    left_join(
      df |> select(paper_title, research_area, pubpub_link),
      by = "paper_title"
    ) |>
    mutate(
      manual_tier = case_when(
        adj_status_csv == "evidence of updating"          ~ "Confirmed update",
        adj_status_csv == "Stated intention to update"    ~ "Stated intention",
        adj_status_csv == "mixed evidence/minor updating" ~ "Mixed / minor",
        adj_status_csv == "Unlikely to update"            ~ "Unlikely",
        TRUE                                              ~ "Unclassified"
      ),
      combined_tier = case_when(
        manual_tier == "Confirmed update" & pct_influenced >= 40 ~ "Confirmed + strong LLM",
        manual_tier == "Confirmed update"                        ~ "Confirmed + weaker LLM",
        pct_influenced >= 30                                     ~ "LLM: likely influenced",
        pct_influenced >= 10                                     ~ "LLM: possibly influenced",
        TRUE                                                     ~ "LLM: largely independent"
      ),
      combined_tier = factor(combined_tier, levels = c(
        "Confirmed + strong LLM", "Confirmed + weaker LLM",
        "LLM: likely influenced", "LLM: possibly influenced",
        "LLM: largely independent"
      ))
    )

  n_analyzed      <- nrow(llm_enriched)
  n_total_fetched <- nrow(llm_df)
  med_pct         <- median(llm_enriched$pct_influenced, na.rm = TRUE)
  n_high          <- sum(llm_enriched$pct_influenced >= 30, na.rm = TRUE)
} else {
  message("LLM results not found — run scripts/run_paper_change_llm.py first")
  llm_df <- llm_analyzed <- llm_enriched <- data.frame()
  n_analyzed <- n_total_fetched <- n_high <- 0
  med_pct <- NA_real_
}
```

```{r}
#| label: llm-attribution-summary
#| code-summary: "Summary text"
#| results: asis

if (nrow(llm_enriched) > 0) {
  n_confirmed <- sum(llm_enriched$manual_tier == "Confirmed update")
  n_llm_high  <- sum(llm_enriched$manual_tier == "Unclassified" &
                       llm_enriched$pct_influenced >= 30)
  cat(glue::glue(
    "Of {n_total_fetched} papers with fetchable DOIs, {n_analyzed} had meaningful text ",
    "changes and a matched evaluation. {n_confirmed} are manually confirmed updates; ",
    "of the remaining {n_analyzed - n_confirmed}, the LLM finds {n_llm_high} with ≥30% ",
    "of changes likely driven by evaluator feedback (median across all: {round(med_pct)}%)."
  ))
}
```

```{r}
#| label: tbl-llm-attribution
#| tbl-cap: "Papers with LLM attribution results, sorted by estimated % of evaluation-driven changes. 'Confirmed update' papers (Bilal, Kremer, Liang) were manually verified and also LLM-analyzed. Signal tier combines both evidence streams."
#| code-summary: "Combined evidence table"

if (nrow(llm_enriched) > 0) {
  tier_bg <- c(
    "Confirmed + strong LLM"   = "#c8e6c9",
    "Confirmed + weaker LLM"   = "#d4edda",
    "LLM: likely influenced"   = "#e8f5e9",
    "LLM: possibly influenced" = "#fff3e0",
    "LLM: largely independent" = "#f8fafc"
  )

  tbl_llm <- llm_enriched |>
    arrange(desc(pct_influenced)) |>
    mutate(
      Title = ifelse(
        !is.na(pubpub_link) & nchar(pubpub_link) > 0,
        paste0('<a href="', pubpub_link, '" target="_blank">',
               str_trunc(paper_title, 60), '</a>'),
        str_trunc(paper_title, 60)
      ),
      `Manual`        = manual_tier,
      `Dep.`          = ifelse(deposit_after_uj, "yes", "no"),
      `Text Δ%`       = round(text_chg_pct, 1),
      `LLM %`         = paste0(round(pct_influenced), "%"),
      `Conf.`         = paste0(attr_confidence, "/5"),
      `Signal tier`   = combined_tier,
      `LLM narrative` = str_trunc(narrative, 280)
    ) |>
    select(Title, Manual, Dep., `Text Δ%`, `LLM %`, Conf., `Signal tier`, `LLM narrative`)

  datatable(
    tbl_llm,
    escape   = FALSE,
    rownames = FALSE,
    options  = list(
      pageLength = 10,
      dom        = "tip",
      scrollX    = TRUE,
      columnDefs = list(
        list(width = "18%", targets = 0),
        list(width = "10%", targets = 1),
        list(width = "4%",  targets = 2),
        list(width = "5%",  targets = 3),
        list(width = "5%",  targets = 4),
        list(width = "4%",  targets = 5),
        list(width = "13%", targets = 6),
        list(width = "41%", targets = 7)
      )
    )
  ) |>
  formatStyle(
    "Signal tier",
    backgroundColor = styleEqual(names(tier_bg), unname(tier_bg))
  )
} else {
  cat("*No LLM attribution results available yet — run scripts/run_paper_change_llm.py first.*")
}
```

```{r}
#| label: fig-combined-dot
#| fig-cap: "LLM attribution for 8 papers with pre/post versions. x-axis: % of major changes attributed to evaluator feedback (Claude Opus 4.6, conservative). Point size scales with % text change. Green = manually confirmed update; blue = not yet manually classified. Dashed line at 30%."
#| fig-height: 4.0

if (nrow(llm_enriched) > 0) {
  dot_colors <- c("Confirmed update" = "#2D9D5E", "Unclassified" = "#2B7CE9")

  llm_enriched |>
    mutate(paper_short = str_trunc(paper_title, 46)) |>
    ggplot(aes(
      x = pct_influenced, y = reorder(paper_short, pct_influenced),
      color = manual_tier, size = text_chg_pct
    )) +
    geom_vline(xintercept = 30, linetype = "dashed", color = "grey55", linewidth = 0.5) +
    geom_point(alpha = 0.85) +
    scale_color_manual(values = dot_colors, name = "Manual classification") +
    scale_size_continuous(range = c(3, 9), name = "Text change %") +
    scale_x_continuous(limits = c(0, 72), labels = \(x) paste0(x, "%"),
                       breaks = c(0, 15, 30, 50, 70)) +
    labs(x = "% of major changes attributed to evaluator feedback", y = NULL) +
    theme_uj() +
    theme(legend.position = "bottom", legend.box = "vertical",
          legend.margin = margin(0, 0, 0, 0))
}
```

```{r}
#| label: fig-combined-overview
#| fig-cap: "All 57 tracked papers by combined evidence tier. LLM-analyzed papers (top 5 tiers, greens) use the combined LLM + manual signal; all others use manual classification only."
#| fig-height: 4.2

if (nrow(llm_enriched) > 0) {
  llm_summary <- llm_enriched |>
    count(combined_tier) |>
    rename(group = combined_tier) |>
    mutate(group = as.character(group), source = "LLM-analyzed")

  llm_titles <- llm_enriched$paper_title
  other_summary <- df |>
    filter(!paper_title %in% llm_titles) |>
    mutate(group = case_when(
      adj_status_clean == "Evidence of updating"       ~ "Confirmed (manual only)",
      adj_status_clean == "Stated intention to update" ~ "Stated intention",
      adj_status_clean == "Mixed / minor updating"     ~ "Mixed / minor",
      adj_status_clean == "Unlikely to update"         ~ "Unlikely to update",
      TRUE                                             ~ "Not yet assessed"
    )) |>
    count(group) |>
    mutate(source = "Manual only")

  n_not_fetched <- llm_df |> filter(skipped) |> nrow()
  if (n_not_fetched > 0) {
    other_summary <- bind_rows(
      other_summary,
      data.frame(group = "No version fetched", n = n_not_fetched, source = "Manual only")
    )
  }

  combined_summary <- bind_rows(llm_summary, other_summary) |>
    mutate(group = factor(group, levels = c(
      "Confirmed + strong LLM", "Confirmed + weaker LLM",
      "LLM: likely influenced", "LLM: possibly influenced",
      "LLM: largely independent",
      "Confirmed (manual only)", "Stated intention",
      "Mixed / minor", "Unlikely to update",
      "No version fetched", "Not yet assessed"
    )))

  overview_colors <- c(
    "Confirmed + strong LLM"  = "#1a7a47",
    "Confirmed + weaker LLM"  = "#2D9D5E",
    "LLM: likely influenced"  = "#5AE08A",
    "LLM: possibly influenced"= "#a8d5b5",
    "LLM: largely independent"= "#d0e8d8",
    "Confirmed (manual only)" = "#2B7CE9",
    "Stated intention"        = "#7EB8F7",
    "Mixed / minor"           = "#E8722A",
    "Unlikely to update"      = "#C45A1E",
    "No version fetched"      = "#E2E8F0",
    "Not yet assessed"        = "#CBD5E1"
  )

  ggplot(combined_summary, aes(x = n, y = reorder(group, n), fill = group)) +
    geom_col(width = 0.65, show.legend = FALSE) +
    geom_text(aes(label = n), hjust = -0.3, size = 3.2) +
    scale_fill_manual(values = overview_colors) +
    scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
    labs(x = "Number of papers", y = NULL,
         title = "Combined evidence: all 57 tracked evaluations") +
    theme_uj() +
    theme(axis.text.y = element_text(size = 9))
}
```

## Summary Statistics

```{r}
#| label: summary-numbers
#| code-summary: "Compute headline statistics"

n_total      <- nrow(df)
n_assessed   <- df |> filter(adj_status_clean != "Not yet assessed") |> nrow()
n_evidence   <- df |> filter(adj_status_clean == "Evidence of updating") |> nrow()
n_intention  <- df |> filter(adj_status_clean == "Stated intention to update") |> nrow()
n_any_positive <- df |>
  filter(adj_status_clean %in% c("Evidence of updating",
                                  "Stated intention to update",
                                  "Mixed / minor updating")) |> nrow()
n_formal     <- df |> filter(author_response_clean == "Formal response") |> nrow()
n_any_resp   <- df |> filter(author_response_clean %in%
                               c("Formal response", "Informal response")) |> nrow()
```

Of `r n_total` tracked evaluations, `r n_assessed` have been manually assessed (the remainder are too recent or pending follow-up). Among assessed papers: `r n_evidence` show clear evidence of substantive updating, `r n_intention` authors stated intention to update, and `r n_any_positive` show at least some positive signal. On author engagement: `r n_any_resp` papers received a formal or informal response, of which `r n_formal` were formal written responses.

```{r}
#| label: fig-adj-bar
#| fig-cap: "Distribution of adjustment outcomes across tracked Unjournal evaluations. 'Not yet assessed' includes papers that are too recent or where follow-up is pending."
#| fig-height: 3.8

df |>
  count(adj_status_clean) |>
  ggplot(aes(x = reorder(adj_status_clean, -n), y = n, fill = adj_status_clean)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = adj_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Number of papers",
       title = "Author adjustment outcomes following Unjournal evaluations") +
  theme_uj() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

```{r}
#| label: fig-response-bar
#| fig-cap: "Type of author response to Unjournal evaluation. 'Not yet coded' = response status not yet assessed for these papers (blank in data); distinct from confirmed non-response."
#| fig-height: 3.5

resp_colors <- c(
  "Formal response"  = "#2B7CE9",
  "Informal response"= "#5DA3F5",
  "No response"      = "#CBD5E1",
  "Not yet coded"    = "#E2E8F0"
)

df |>
  count(author_response_clean) |>
  ggplot(aes(x = reorder(author_response_clean, -n), y = n, fill = author_response_clean)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = resp_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Number of papers",
       title = "Type of author response to evaluation") +
  theme_uj() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

## Full Tabulation {.callout collapse="true"}

::: {.callout-note collapse="true"}
## All 57 tracked evaluations

Click column headers to sort; use the search box to filter. Adjustment status is color-coded: green = positive signal, orange = minor/uncertain, grey = unlikely or not yet assessed.

```{r}
#| label: tbl-full
#| code-summary: "Build interactive DT table"

tbl <- df |>
  mutate(
    title_linked = if_else(
      !is.na(pubpub_link) & pubpub_link != "",
      paste0('<a href="', pubpub_link, '" target="_blank">', str_trunc(paper_title, 72), '</a>'),
      str_trunc(paper_title, 72)
    ),
    authors_short = str_trunc(authors, 40),
    nb_icon = if_else(
      !is.na(nb_link) & nb_link != "",
      paste0('<a href="', nb_link, '" target="_blank" title="NotebookLM">&#128214;</a>'),
      ""
    )
  ) |>
  select(
    Paper      = title_linked,
    Authors    = authors_short,
    Area       = research_area,
    Response   = author_response_clean,
    Adjustment = adj_status_clean,
    `Pub.`     = pub_status,
    NB         = nb_icon
  )

adj_bg <- c(
  "Evidence of updating"        = "#d4edda",
  "Stated intention to update"  = "#e8f5e9",
  "Mixed / minor updating"      = "#fff3e0",
  "Unlikely to update"          = "#fce4d4",
  "Not yet assessed"            = "#f8fafc"
)

datatable(
  tbl,
  escape   = FALSE,
  rownames = FALSE,
  filter   = "top",
  options  = list(
    pageLength = 8,
    autoWidth  = FALSE,
    scrollX    = TRUE,
    dom        = "lfrtip",
    columnDefs = list(
      list(width = "36%", targets = 0),
      list(width = "18%", targets = 1),
      list(width = "8%",  targets = 2),
      list(width = "11%", targets = 3),
      list(width = "16%", targets = 4),
      list(width = "8%",  targets = 5),
      list(width = "3%",  targets = 6)
    )
  ),
  caption = "All tracked Unjournal evaluations. NB = NotebookLM analysis link."
) |>
  formatStyle("Adjustment",
    backgroundColor = styleEqual(names(adj_bg), unname(adj_bg)))
```

:::
