---
title: "Did Authors Respond to Unjournal Evaluations?"
author: "The Unjournal"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

::: {.content-visible when-format="pdf"}
The interactive tables in this appendix are available in the online version at <https://valentinklotzbuecher.github.io/llm-uj-research-eval/paper_response_analysis.html>.
:::

::: {.callout-note}
This page reports an April 2026 snapshot of 57 tracked Unjournal evaluations: manual staff assessment for the full set, plus an exploratory LLM comparison of eight paper pairs. We are now rebuilding the comparison workflow around preserved source files, version checks, and human review. See [Combined Evidence](#combined-evidence) below.
:::

## Overview

The Unjournal commissions open, structured evaluations of working papers in economics and social science. A natural question is whether this process changes research: do authors engage with feedback and revise their work?

This page draws on manual classification of 57 tracked evaluations and, for a subset, automated PDF comparison and exploratory LLM screening. Background: [The Unjournal's knowledge base](https://globalimpact.gitbook.io/the-unjournal-project-and-communication-space/benefits-and-features/more-reliable-and-useful-evaluation/author-engagement-evidence).

```{r}
#| label: setup
#| code-summary: "Load data and define helpers"

library(dplyr)
library(ggplot2)
library(readr)
library(DT)
library(tidyr)
library(stringr)
library(glue)
library(jsonlite)

df <- read_csv("data/author_adjustment_manual.csv", show_col_types = FALSE)

df <- df |>
  rename(
    paper_title    = label_paper_title,
    pubpub_link    = dup_pubpub_final_links,
    adj_status     = `Adjusted_paper?`,
    updated_manual = `Updated since UJ report -- manual confirmation`,
    deposit_after  = `deposit date > unjournal pub date`,
    research_area  = main_cause_cat_abbrev,
    nb_link        = notebookLM_link,
    pub_status     = publication_status,
    wp_date        = working_paper_release_date,
    uj_pub_date    = publication_date_unjournal
  )

adj_labels <- c(
  "Evidence of updating", "Stated intention to update",
  "Mixed / minor updating", "Unlikely to update", "Not yet assessed"
)

df <- df |>
  mutate(
    adj_status_clean = case_when(
      adj_status == "evidence of updating"          ~ "Evidence of updating",
      adj_status == "Stated intention to update"    ~ "Stated intention to update",
      adj_status == "mixed evidence/minor updating" ~ "Mixed / minor updating",
      adj_status == "Unlikely to update"            ~ "Unlikely to update",
      TRUE                                          ~ "Not yet assessed"
    ),
    adj_status_clean = factor(adj_status_clean, levels = adj_labels),
    author_response_clean = case_when(
      author_response == "Formal response" ~ "Formal response",
      author_response == "Informal"        ~ "Informal response",
      author_response == "None?"           ~ "No response",
      TRUE                                 ~ "Not yet coded"
    )
  )

theme_uj <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(panel.grid.minor = element_blank(),
          axis.text = element_text(size = base_size * 0.85),
          plot.caption = element_text(size = base_size * 0.75, color = "grey50"),
          legend.position = "right")
}
```

## Combined Evidence of Paper Updating {#combined-evidence}

Two evidence streams are shown together: manual classification for all 57 papers, and an exploratory Claude Opus 4.6 screen for eight papers where before/after PDFs were available and meaningful text changes were detected. The latter is a way to prioritize inspection, not a causal estimate.

::: {.callout-important}
## Current validation work (July 2026)

The April figures below remain the historical snapshot. A current source refresh has captured 30 public current-paper versions, created nine before/after comparison records, and placed five records in a queue for agent-assisted triage after preliminary document, identity, and timing checks. These are coverage figures for a workflow, not findings that an evaluation caused a change.

For each candidate, the revised workflow preserves PDF snapshots and hashes, checks that the documents are the same paper and in the right order, records page- and section-level changes, and requires human review before any public claim of evaluation-linked updating. A newer PDF, a textual difference, or a model suggestion is not enough on its own.
:::

::: {.callout-note collapse="true"}
## Pipeline and caveats

**Current scripts**: `scripts/paper_response_evidence.py` for source snapshots, validation, and comparison cards; `scripts/run_paper_response_agents.py` for bounded, review-only triage.

- **Historical screen shown below**: before version = PDF sent to evaluators; after version = then-current NBER/arXiv PDF, fetched April 2026. Eight pairs had meaningful text changes and a matched evaluation file.
- **Current validation**: the refresh keeps immutable PDF snapshots; verifies paper identity and chronology; produces page- and section-anchored change cards; and puts ambiguous cases into human review.
- **Models**: the historical Opus screen used bounded extracted text, so it could not reliably assess figures, tables, or all document context. The current agent triage is also deliberately non-decisive: it can identify evidence to inspect, but cannot label causal influence.
- **Interpretation**: a newer version does not imply that changes were evaluation-driven. Some papers are already being revised, and a substantial revision can have unrelated causes.
:::

```{r}
#| label: llm-attribution-load
#| code-summary: "Load and merge LLM + manual data"

nn <- function(x, default = NA) if (!is.null(x)) x else default

llm_results_path <- "data/paper_change_llm_results.json"

# Signal tier labels used throughout — define once
tier_levels <- c(
  "Confirmed, LLM >=40%",
  "Confirmed, LLM <40%",
  "LLM >=30% (not confirmed)",
  "LLM 10-29%",
  "LLM <10%"
)

tier_colors_llm <- c(
  "Confirmed, LLM >=40%"      = "#1a7a47",
  "Confirmed, LLM <40%"       = "#2D9D5E",
  "LLM >=30% (not confirmed)" = "#5AE08A",
  "LLM 10-29%"                = "#a8d5b5",
  "LLM <10%"                  = "#d0e8d8"
)

if (file.exists(llm_results_path)) {
  llm_raw <- fromJSON(llm_results_path, simplifyVector = FALSE)

  llm_df <- lapply(llm_raw, function(r) {
    oa <- r$overall_assessment
    data.frame(
      paper_title      = nn(r$paper_title, ""),
      adj_status_csv   = nn(r$adj_status, ""),
      deposit_after_uj = isTRUE(r$deposit_after_uj),
      text_chg_pct     = nn(r$text_length_change_pct, NA_real_),
      additions        = nn(r$additions_count, NA_integer_),
      deletions        = nn(r$deletions_count, NA_integer_),
      n_major_changes  = length(nn(r$major_changes, list())),
      n_suggestions    = length(nn(r$evaluator_suggestions, list())),
      pct_influenced   = if (!is.null(oa)) nn(oa$pct_likely_influenced, NA_real_) else NA_real_,
      attr_confidence  = if (!is.null(oa)) nn(oa$confidence, NA_real_) else NA_real_,
      narrative        = if (!is.null(oa)) nn(oa$narrative, "") else "",
      eval_found       = !is.null(r$eval_file) && nchar(nn(r$eval_file, "")) > 0,
      skipped          = !is.null(r$skipped_reason),
      stringsAsFactors = FALSE
    )
  }) |> bind_rows()

  llm_analyzed <- llm_df |>
    filter(!skipped & eval_found & n_major_changes > 0 & !is.na(pct_influenced))

  llm_enriched <- llm_analyzed |>
    left_join(df |> select(paper_title, research_area, pubpub_link), by = "paper_title") |>
    mutate(
      manual_tier = case_when(
        adj_status_csv == "evidence of updating"          ~ "Confirmed update",
        adj_status_csv == "Stated intention to update"    ~ "Stated intention",
        adj_status_csv == "mixed evidence/minor updating" ~ "Mixed / minor",
        adj_status_csv == "Unlikely to update"            ~ "Unlikely",
        TRUE                                              ~ "Unclassified"
      ),
      combined_tier = case_when(
        manual_tier == "Confirmed update" & pct_influenced >= 40 ~ tier_levels[1],
        manual_tier == "Confirmed update"                        ~ tier_levels[2],
        pct_influenced >= 30                                     ~ tier_levels[3],
        pct_influenced >= 10                                     ~ tier_levels[4],
        TRUE                                                     ~ tier_levels[5]
      ),
      combined_tier = factor(combined_tier, levels = tier_levels)
    )

  n_analyzed      <- nrow(llm_enriched)
  n_total_fetched <- nrow(llm_df)
  med_pct         <- median(llm_enriched$pct_influenced, na.rm = TRUE)
  n_high          <- sum(llm_enriched$pct_influenced >= 30, na.rm = TRUE)
} else {
  llm_df <- llm_analyzed <- llm_enriched <- data.frame()
  n_analyzed <- n_total_fetched <- n_high <- 0
  med_pct <- NA_real_
}
```

```{r}
#| label: llm-attribution-summary
#| results: asis
#| code-summary: "Summary"

if (nrow(llm_enriched) > 0) {
  n_confirmed <- sum(llm_enriched$manual_tier == "Confirmed update")
  n_llm_high  <- sum(llm_enriched$manual_tier == "Unclassified" &
                       llm_enriched$pct_influenced >= 30)
  cat(glue::glue(
    "Of {n_total_fetched} papers with fetchable DOIs, {n_analyzed} had meaningful text ",
    "changes and a matched evaluation. {n_confirmed} are manually confirmed updates; ",
    "of the remaining {n_analyzed - n_confirmed}, the LLM finds {n_llm_high} with at least 30% ",
    "of changes likely driven by evaluator feedback (median across all: {round(med_pct)}%)."
  ))
}
```

```{r}
#| label: fig-combined-waffle
#| fig-cap: "All 57 tracked papers by combined evidence tier (each square = 1 paper). Green tones: papers where LLM analysis was run, shaded by attribution score and manual confirmation status. Blue: manually confirmed updates without LLM analysis. Orange/grey: stated intention, weak signal, or not yet assessed."
#| fig-height: 4.5

if (nrow(llm_enriched) > 0) {
  # Build one row per paper with tier assignment
  waffle_llm <- llm_enriched |>
    select(paper_title, combined_tier) |>
    mutate(group = as.character(combined_tier))

  waffle_manual <- df |>
    filter(!paper_title %in% llm_enriched$paper_title) |>
    mutate(group = case_when(
      adj_status_clean == "Evidence of updating"       ~ "Confirmed (manual only)",
      adj_status_clean == "Stated intention to update" ~ "Stated intention",
      adj_status_clean == "Mixed / minor updating"     ~ "Mixed / minor",
      adj_status_clean == "Unlikely to update"         ~ "Unlikely",
      TRUE                                             ~ "Not yet assessed"
    )) |>
    select(paper_title, group)

  waffle_all_groups <- c(tier_levels,
    "Confirmed (manual only)", "Stated intention",
    "Mixed / minor", "Unlikely", "Not yet assessed")

  waffle_colors <- c(
    tier_colors_llm,
    "Confirmed (manual only)" = "#2B7CE9",
    "Stated intention"        = "#7EB8F7",
    "Mixed / minor"           = "#E8722A",
    "Unlikely"                = "#C45A1E",
    "Not yet assessed"        = "#CBD5E1"
  )

  waffle_data <- bind_rows(waffle_llm, waffle_manual) |>
    mutate(group = factor(group, levels = waffle_all_groups)) |>
    arrange(group) |>
    mutate(rank = row_number(),
           x = (rank - 1) %% 9,
           y = -((rank - 1) %/% 9))

  ggplot(waffle_data, aes(x = x, y = y, fill = group)) +
    geom_tile(color = "white", linewidth = 1.5) +
    coord_equal() +
    scale_fill_manual(values = waffle_colors, name = NULL,
                      drop = FALSE) +
    theme_void() +
    theme(legend.position = "right",
          legend.text = element_text(size = 9),
          plot.title = element_text(size = 11, face = "plain",
                                    margin = margin(b = 6)),
          plot.caption = element_text(size = 8, color = "grey40",
                                      hjust = 0, margin = margin(t = 8))) +
    labs(title = "All 57 tracked papers — each square = 1 paper",
         caption = "LLM % = share of major post-evaluation changes attributed to evaluator feedback (Claude Opus 4.6)")
}
```

```{r}
#| label: fig-combined-dot
#| fig-cap: "LLM attribution for the 8 papers with pre/post versions compared. x-axis: estimated % of major changes attributed to evaluator feedback (Claude Opus 4.6, conservative). Point size scales with text change magnitude. Green = manually confirmed update; blue = not yet manually classified. Dashed line at 30%."
#| fig-height: 3.8

if (nrow(llm_enriched) > 0) {
  dot_colors <- c("Confirmed update" = "#2D9D5E", "Unclassified" = "#2B7CE9")

  llm_enriched |>
    mutate(paper_short = str_trunc(paper_title, 46)) |>
    ggplot(aes(x = pct_influenced, y = reorder(paper_short, pct_influenced),
               color = manual_tier, size = text_chg_pct)) +
    geom_vline(xintercept = 30, linetype = "dashed", color = "grey55", linewidth = 0.5) +
    geom_point(alpha = 0.85) +
    scale_color_manual(values = dot_colors, name = "Manual classification") +
    scale_size_continuous(range = c(3, 9), name = "Text change %") +
    scale_x_continuous(limits = c(0, 72), labels = \(x) paste0(x, "%"),
                       breaks = c(0, 15, 30, 50, 70)) +
    labs(x = "% of major changes attributed to evaluator feedback", y = NULL) +
    theme_uj() +
    theme(legend.position = "bottom", legend.box = "vertical",
          legend.margin = margin(0, 0, 0, 0))
}
```

```{r}
#| label: tbl-llm-attribution
#| tbl-cap: "Papers with LLM attribution results, sorted by % of changes attributed to evaluator feedback. 'Confirmed' = manually verified by Unjournal staff. Signal tier: 'Confirmed, LLM >=40%' = manually confirmed AND LLM finds strong corroboration; 'Confirmed, LLM <40%' = confirmed but weaker LLM signal; 'LLM >=30% (not confirmed)' = unconfirmed but LLM finds likely influence."
#| code-summary: "Combined evidence table"
#| eval: !expr knitr::is_html_output()

if (nrow(llm_enriched) > 0) {
  tier_bg <- c(
    "Confirmed, LLM \u226540%"      = "#c8e6c9",
    "Confirmed, LLM <40%"           = "#d4edda",
    "LLM \u226530% (not confirmed)" = "#e8f5e9",
    "LLM 10\u201329%"               = "#fff3e0",
    "LLM <10%"                      = "#f8fafc"
  )

  tbl_llm <- llm_enriched |>
    arrange(desc(pct_influenced)) |>
    mutate(
      Title = ifelse(
        !is.na(pubpub_link) & nchar(pubpub_link) > 0,
        paste0('<a href="', pubpub_link, '" target="_blank">',
               str_trunc(paper_title, 60), '</a>'),
        str_trunc(paper_title, 60)
      ),
      `Manual`      = manual_tier,
      `LLM %`       = paste0(round(pct_influenced), "%"),
      `Conf.`       = paste0(attr_confidence, "/5"),
      `Signal tier` = combined_tier,
      `LLM narrative` = str_trunc(narrative, 350)
    ) |>
    select(Title, Manual, `LLM %`, Conf., `Signal tier`, `LLM narrative`)

  datatable(
    tbl_llm,
    escape   = FALSE,
    rownames = FALSE,
    options  = list(
      pageLength = 5,
      dom        = "tip",
      scrollX    = TRUE,
      columnDefs = list(
        list(width = "18%", targets = 0),
        list(width = "11%", targets = 1),
        list(width = "6%",  targets = 2),
        list(width = "5%",  targets = 3),
        list(width = "16%", targets = 4),
        list(width = "44%", targets = 5)
      )
    )
  ) |>
  formatStyle("Signal tier",
    backgroundColor = styleEqual(names(tier_bg), unname(tier_bg)))
} else {
  cat("*No LLM attribution results available yet.*")
}
```

## Summary Statistics

```{r}
#| label: summary-numbers
#| code-summary: "Compute headline statistics"

n_total      <- nrow(df)
n_assessed   <- df |> filter(adj_status_clean != "Not yet assessed") |> nrow()
n_evidence   <- df |> filter(adj_status_clean == "Evidence of updating") |> nrow()
n_intention  <- df |> filter(adj_status_clean == "Stated intention to update") |> nrow()
n_any_positive <- df |>
  filter(adj_status_clean %in% c("Evidence of updating",
                                  "Stated intention to update",
                                  "Mixed / minor updating")) |> nrow()
n_formal  <- df |> filter(author_response_clean == "Formal response") |> nrow()
n_any_resp <- df |>
  filter(author_response_clean %in% c("Formal response", "Informal response")) |> nrow()
```

Of `r n_total` tracked evaluations, `r n_assessed` have been manually assessed. Among these: `r n_evidence` show clear evidence of substantive updating, `r n_intention` authors stated intention to update, and `r n_any_positive` show at least some positive signal. On engagement: `r n_any_resp` papers received a formal or informal author response, of which `r n_formal` were formal written responses.

```{r}
#| label: fig-response-bar
#| fig-cap: "Type of author response to Unjournal evaluation. 'Not yet coded' indicates the response status has not been assessed for those papers (37 blanks in data); this is distinct from confirmed non-response (only 1 paper is explicitly coded as no response)."
#| fig-height: 3.2

resp_colors <- c(
  "Formal response"   = "#2B7CE9",
  "Informal response" = "#5DA3F5",
  "No response"       = "#CBD5E1",
  "Not yet coded"     = "#E2E8F0"
)

df |>
  count(author_response_clean) |>
  ggplot(aes(x = reorder(author_response_clean, -n), y = n,
             fill = author_response_clean)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = resp_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Number of papers",
       title = "Type of author response to evaluation") +
  theme_uj() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

## Full Tabulation

::: {.callout-note collapse="true"}
## All 57 tracked evaluations

Click column headers to sort; use the search box to filter. Adjustment status is color-coded: green = positive signal, orange = minor/uncertain, grey = unlikely or not yet assessed.

```{r}
#| label: tbl-full
#| code-summary: "Build interactive DT table"
#| eval: !expr knitr::is_html_output()

tbl <- df |>
  mutate(
    title_linked = if_else(
      !is.na(pubpub_link) & pubpub_link != "",
      paste0('<a href="', pubpub_link, '" target="_blank">',
             str_trunc(paper_title, 72), '</a>'),
      str_trunc(paper_title, 72)
    ),
    authors_short = str_trunc(authors, 40),
    nb_icon = if_else(
      !is.na(nb_link) & nb_link != "",
      paste0('<a href="', nb_link, '" target="_blank" title="NotebookLM">&#128214;</a>'),
      ""
    )
  ) |>
  select(Paper = title_linked, Authors = authors_short,
         Area = research_area, Response = author_response_clean,
         Adjustment = adj_status_clean, `Pub.` = pub_status, NB = nb_icon)

adj_bg <- c(
  "Evidence of updating"        = "#d4edda",
  "Stated intention to update"  = "#e8f5e9",
  "Mixed / minor updating"      = "#fff3e0",
  "Unlikely to update"          = "#fce4d4",
  "Not yet assessed"            = "#f8fafc"
)

datatable(tbl, escape = FALSE, rownames = FALSE, filter = "top",
  options = list(
    pageLength = 8, autoWidth = FALSE, scrollX = TRUE, dom = "lfrtip",
    columnDefs = list(
      list(width = "36%", targets = 0),
      list(width = "18%", targets = 1),
      list(width = "8%",  targets = 2),
      list(width = "11%", targets = 3),
      list(width = "16%", targets = 4),
      list(width = "8%",  targets = 5),
      list(width = "3%",  targets = 6)
    )
  ),
  caption = "All tracked Unjournal evaluations. NB = NotebookLM analysis link."
) |>
  formatStyle("Adjustment",
    backgroundColor = styleEqual(names(adj_bg), unname(adj_bg)))
```

:::
