---
engine: knitr
---

# Results

```{r}
#| label: setup-results
#| code-summary: "Setup and libraries"
#| code-fold: true
#| message: false
#| warning: false

source("setup_params.R")

library("tidyverse")
library("janitor")
library("stringr")
library("here")
library("knitr")
library("kableExtra")
library("ggrepel")
library("scales")
library("jsonlite")
library("purrr")
library("tibble")

# Theme and colors — crisp palette (high saturation, maximum separation)
UJ_ORANGE <- "#E8722A"   # vivid saffron orange
UJ_GREEN  <- "#2D9D5E"   # rich emerald green
UJ_BLUE   <- "#2B7CE9"   # clear azure blue

MODEL_COLORS <- c(
  "GPT-5 Pro"        = "#E8722A",   # saffron orange (focal)
  "GPT-5.2 Pro"      = "#D62839",   # bright crimson
  "GPT-4o-mini"      = "#17B890",   # vivid teal
  "Claude Sonnet 4"  = "#A855F7",   # vivid purple
  "Claude Opus 4.6"  = "#7C3AED",   # deep violet
  "Gemini 2.0 Flash" = "#2B7CE9",   # clear azure
  "Human"            = "#2D9D5E"    # emerald green
)

theme_uj <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor   = element_blank(),
      panel.grid.major   = element_line(linewidth = 0.3, color = "grey88"),
      plot.title          = element_text(face = "bold", size = rel(1.1)),
      plot.title.position = "plot",
      plot.subtitle       = element_text(color = "grey40", size = rel(0.9)),
      plot.caption        = element_text(color = "grey50", size = rel(0.8), hjust = 0),
      axis.title          = element_text(size = rel(0.95)),
      axis.text           = element_text(size = rel(0.88)),
      legend.position     = "bottom",
      legend.text         = element_text(size = rel(0.88)),
      legend.title        = element_text(size = rel(0.9), face = "bold"),
      strip.text          = element_text(face = "bold", size = rel(0.95))
    )
}

canon_metric <- function(x) dplyr::recode(
  x,
  "advancing_knowledge" = "adv_knowledge",
  "open_science"        = "open_sci",
  "logic_communication" = "logic_comms",
  "global_relevance"    = "gp_relevance",
  "claims_evidence"     = "claims",
  .default = x
)

`%||%` <- function(x, y) if (!is.null(x)) x else y
```

```{r}
#| label: load-human-data-results
#| code-fold: true
#| code-summary: "Load human evaluation data"
#| message: false

UJmap <- read_delim("data/UJ_map.csv", delim = ";", show_col_types = FALSE) |>
  mutate(label_paper_title = research, label_paper = paper) |>
  select(c("label_paper_title", "label_paper"))

rsx <- read_csv("data/rsx_evalr_rating.csv", show_col_types = FALSE) |>
  clean_names() |>
  mutate(label_paper_title = research) |>
  select(-c("research"))

research <- read_csv("data/research.csv", show_col_types = FALSE) |>
  clean_names() |>
  filter(status == "50_published evaluations (on PubPub, by Unjournal)") |>
  left_join(UJmap, by = c("label_paper_title")) |>
  mutate(doi = str_trim(doi)) |>
  mutate(label_paper = case_when(
    doi == "https://doi.org/10.3386/w31162" ~ "Walker et al. 2023",
    doi == "doi.org/10.3386/w32728" ~ "Hahn et al. 2025",
    doi == "https://doi.org/10.3386/w30011" ~ "Bhat et al. 2022",
    doi == "10.1093/wbro/lkae010" ~ "Crawfurd et al. 2023",
    TRUE ~ label_paper
  )) |>
  left_join(rsx, by = c("label_paper_title"))

key_map <- research |>
  transmute(label_paper_title = str_trim(label_paper_title), label_paper = label_paper) |>
  filter(!is.na(label_paper_title)) |>
  distinct(label_paper_title, label_paper) |>
  group_by(label_paper_title) |>
  slice(1) |>
  ungroup()

rsx_research <- rsx |>
  mutate(label_paper_title = str_trim(label_paper_title)) |>
  left_join(key_map, by = "label_paper_title", relationship = "many-to-one")

metrics_human <- rsx_research |>
  mutate(criteria = canon_metric(criteria)) |>
  filter(criteria %in% c("overall", "claims", "methods", "adv_knowledge", "logic_comms", "open_sci", "gp_relevance")) |>
  transmute(
    paper = label_paper, criteria, evaluator, model = "Human",
    mid = as.numeric(middle_rating),
    lo = suppressWarnings(as.numeric(lower_ci)),
    hi = suppressWarnings(as.numeric(upper_ci))
  ) |>
  filter(!is.na(paper), !is.na(mid)) |>
  mutate(
    lo = ifelse(is.finite(lo), pmax(0, pmin(100, lo)), NA_real_),
    hi = ifelse(is.finite(hi), pmax(0, pmin(100, hi)), NA_real_)
  ) |>
  mutate(across(c(mid, lo, hi), ~ round(.x, 4))) |>
  distinct(paper, criteria, model, evaluator, mid, lo, hi)

human_avg <- metrics_human |>
  filter(criteria == "overall") |>
  group_by(paper) |>
  summarise(
    human_mid = mean(mid, na.rm = TRUE),
    human_lo = mean(lo, na.rm = TRUE),
    human_hi = mean(hi, na.rm = TRUE),
    n_human = n(),
    .groups = "drop"
  )

n_human_papers <- n_distinct(metrics_human$paper)
n_human_evaluators <- n_distinct(metrics_human$evaluator)
```

```{r}
#| label: load-llm-data-results
#| code-fold: true
#| code-summary: "Load LLM evaluation data (all models)"
#| message: false

model_dirs <- list(
  "gpt5_pro_updated_jan2026" = "GPT-5 Pro",
  "gpt52_pro_focal_jan2026" = "GPT-5.2 Pro",
  "gpt_4o_mini_2024_07_18" = "GPT-4o-mini",
  "claude_sonnet_4_20250514" = "Claude Sonnet 4",
  "claude_opus_4_6" = "Claude Opus 4.6",
  "gemini_2.0_flash" = "Gemini 2.0 Flash"
)

parse_response <- function(path, model_name) {
  tryCatch({
    r <- jsonlite::fromJSON(path, simplifyVector = FALSE)
    paper <- basename(path) |>
      str_replace("\\.response\\.json$", "") |>
      str_replace_all("_", " ")

    parsed <- NULL
    if (!is.null(r$parsed) && length(r$parsed) > 0) {
      parsed <- r$parsed
    } else if (!is.null(r$output_text) && nchar(r$output_text) > 0) {
      txt <- r$output_text
      txt <- sub("^\\s*```[a-z]*\\s*\n?", "", txt)
      txt <- sub("\\s*```\\s*$", "", txt)
      parsed <- jsonlite::fromJSON(txt, simplifyVector = TRUE)
    } else if (!is.null(r$output)) {
      msg <- purrr::detect(r$output, ~ .x$type == "message", .default = NULL)
      if (!is.null(msg) && length(msg$content) > 0) {
        parsed <- jsonlite::fromJSON(msg$content[[1]]$text, simplifyVector = TRUE)
      }
    }

    if (is.null(parsed)) return(NULL)

    metrics <- parsed$metrics
    metric_rows <- list()
    tier_rows <- list()
    tier_names <- c("tier_should", "tier_will", "journal_should", "journal_will")

    for (nm in names(metrics)) {
      if (nm %in% tier_names) {
        tier_kind <- sub("^journal_", "tier_", nm)
        tier_rows[[length(tier_rows) + 1]] <- tibble(
          paper = paper, model = model_name, tier_kind = tier_kind,
          score = metrics[[nm]]$score,
          ci_lower = metrics[[nm]]$ci_lower,
          ci_upper = metrics[[nm]]$ci_upper
        )
      } else {
        metric_rows[[length(metric_rows) + 1]] <- tibble(
          paper = paper, model = model_name, metric = nm,
          midpoint = metrics[[nm]]$midpoint,
          lower_bound = metrics[[nm]]$lower_bound,
          upper_bound = metrics[[nm]]$upper_bound
        )
      }
    }

    input_tok <- r$usage$input_tokens %||% r$input_tokens
    output_tok <- r$usage$output_tokens %||% r$output_tokens

    list(
      metrics = bind_rows(metric_rows),
      tiers = bind_rows(tier_rows),
      tokens = tibble(
        paper = paper, model = model_name,
        input_tokens = input_tok %||% NA_integer_,
        output_tokens = output_tok %||% NA_integer_
      )
    )
  }, error = function(e) NULL)
}

load_all_llm <- function() {
  all_metrics <- list()
  all_tiers <- list()
  all_tokens <- list()

  for (dir_name in names(model_dirs)) {
    model_name <- model_dirs[[dir_name]]
    json_dir <- here("results", dir_name, "json")

    if (dir.exists(json_dir)) {
      files <- list.files(json_dir, pattern = "\\.response\\.json$", full.names = TRUE)

      for (f in files) {
        result <- parse_response(f, model_name)
        if (!is.null(result)) {
          all_metrics[[length(all_metrics) + 1]] <- result$metrics
          all_tiers[[length(all_tiers) + 1]] <- result$tiers
          all_tokens[[length(all_tokens) + 1]] <- result$tokens
        }
      }
    }
  }

  list(
    metrics = bind_rows(all_metrics) |> mutate(criteria = canon_metric(metric)),
    tiers = bind_rows(all_tiers),
    tokens = bind_rows(all_tokens)
  )
}

llm_data <- load_all_llm()
llm_metrics <- llm_data$metrics
llm_tiers <- llm_data$tiers
llm_tokens <- llm_data$tokens

n_llm_models <- n_distinct(llm_metrics$model)
n_llm_papers <- n_distinct(llm_metrics$paper)

matched_papers <- intersect(
  unique(metrics_human$paper),
  unique(llm_metrics$paper)
)
n_matched <- length(matched_papers)

primary_model <- if ("GPT-5 Pro" %in% unique(llm_metrics$model)) "GPT-5 Pro" else unique(llm_metrics$model)[1]
```

We evaluate `r n_llm_models` frontier LLMs against human expert reviews from [The Unjournal](https://unjournal.pubpub.org) for `r n_matched` papers with both human and LLM evaluations. Each model receives the same PDF, system prompt mirroring The Unjournal rubric, and JSON schema requiring a diagnostic summary plus numeric midpoints and 90% credible intervals for every metric. Full methodological details appear in [Data and Methods](methods.qmd).

**Per-paper overview.** @fig-forest-overall displays individual human evaluator ratings (green circles with 90% CIs) alongside the GPT-5 Pro LLM rating (orange diamond with CI) for each paper's overall score. Papers are ordered by human mean rating. Dotted lines mark grand means.

```{r}
#| label: fig-forest-overall
#| fig-cap: "Per-paper overall ratings (0\u2013100 percentile). Green circles show individual human evaluator midpoints with 90% credible intervals; the orange diamond shows the GPT-5 Pro rating with its CI. Papers are sorted left-to-right by descending human mean. Dotted lines mark grand means for each evaluator type."
#| fig-width: 14
#| fig-height: 7
#| code-fold: true

H_ind <- metrics_human |>
  filter(criteria == "overall", paper %in% matched_papers) |>
  mutate(
    lo = ifelse(is.finite(lo), pmax(0, lo), NA_real_),
    hi = ifelse(is.finite(hi), pmin(100, hi), NA_real_)
  )

ord <- H_ind |>
  group_by(paper) |>
  summarise(h_mean = mean(mid, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(h_mean)) |>
  mutate(pos = row_number())

H_plot <- H_ind |>
  inner_join(ord, by = "paper") |>
  group_by(paper) |>
  mutate(
    off = (row_number() - (n() + 1) / 2) * 0.18,
    x   = pos + off
  ) |>
  ungroup()

L_c <- llm_metrics |>
  filter(criteria == "overall", model == primary_model, paper %in% matched_papers) |>
  group_by(paper) |>
  summarise(
    mid = mean(midpoint, na.rm = TRUE),
    lo  = suppressWarnings(min(coalesce(lower_bound, midpoint), na.rm = TRUE)),
    hi  = suppressWarnings(max(coalesce(upper_bound, midpoint), na.rm = TRUE)),
    .groups = "drop"
  ) |>
  inner_join(ord, by = "paper") |>
  mutate(x = pos)

hbar <- mean(ord$h_mean, na.rm = TRUE)
lbar <- mean(L_c$mid, na.rm = TRUE)

if (nrow(L_c) > 0) {
  ggplot() +
    geom_vline(data = ord, aes(xintercept = pos), color = "grey92", linewidth = 0.3) +
    geom_hline(yintercept = hbar, color = UJ_GREEN, linetype = "dotted", linewidth = 0.8) +
    geom_hline(yintercept = lbar, color = UJ_ORANGE, linetype = "dotted", linewidth = 0.8) +
    geom_errorbar(
      data = subset(H_plot, is.finite(lo) & is.finite(hi)),
      aes(x = x, ymin = lo, ymax = hi),
      width = 0, linewidth = 1, alpha = 0.5, color = UJ_GREEN
    ) +
    geom_point(data = H_plot, aes(x = x, y = mid), size = 3.0, alpha = 0.9, color = UJ_GREEN) +
    geom_errorbar(
      data = subset(L_c, is.finite(lo) & is.finite(hi)),
      aes(x = x, ymin = lo, ymax = hi),
      width = 0, linewidth = 1.0, color = UJ_ORANGE
    ) +
    geom_point(data = L_c, aes(x = x, y = mid), size = 3.6, shape = 18, color = UJ_ORANGE) +
    scale_x_continuous(
      breaks = ord$pos, labels = ord$paper,
      expand = expansion(mult = c(0.01, 0.03))
    ) +
    coord_cartesian(ylim = c(0, 100), clip = "off") +
    labs(x = NULL, y = "Percentile (0\u2013100)") +
    theme_uj() +
    annotate("text", x = 4, y = 40,
             label = sprintf("Means \u2014 Human: %.1f   LLM: %.1f", hbar, lbar),
             hjust = 0, size = 4.2, fontface = "italic") +
    theme(
      axis.text.x = element_text(angle = 55, hjust = 1, vjust = 1, size = 9),
      axis.title.y = element_text(size = 12),
      panel.grid.major.x = element_blank(),
      plot.margin = margin(5, 40, 5, 5)
    )
} else {
  cat("Insufficient data for forest plot.\n")
}
```

The spread of green circles within each paper column illustrates substantial human inter-rater variability---CIs often span 20--40 points. In most cases the LLM diamond sits comfortably within the range of human opinions, though several papers show the LLM falling outside the human consensus.

**Overall ratings.** @fig-scatter-focal compares overall ratings from GPT-5 Pro and Claude Opus 4.6---the two highest-performing reasoning models---against human mean ratings. Points on the diagonal indicate perfect agreement; distance from the line reflects disagreement.

```{r}
#| label: fig-scatter-focal
#| fig-cap: "Human mean overall rating (x-axis) vs LLM overall rating (y-axis) for GPT-5 Pro and Claude Opus 4.6. Each point is one paper; the dashed diagonal is the identity line (perfect agreement). Points above the line indicate the LLM rated the paper higher than humans; points below indicate harsher LLM ratings. Paper labels are truncated for readability."
#| fig-width: 11
#| fig-height: 6

focal_models <- c("GPT-5 Pro", "Claude Opus 4.6")

scatter_data <- llm_metrics |>
  filter(criteria == "overall", model %in% focal_models) |>
  inner_join(human_avg, by = "paper") |>
  mutate(
    diff = midpoint - human_mid,
    paper_short = str_trunc(paper, 25),
    human_lo = coalesce(human_lo, human_mid),
    human_hi = coalesce(human_hi, human_mid),
    lower_bound = coalesce(lower_bound, midpoint),
    upper_bound = coalesce(upper_bound, midpoint)
  ) |>
  filter(!is.na(human_mid), !is.na(midpoint))

if (nrow(scatter_data) > 0) {
  ggplot(scatter_data, aes(x = human_mid, y = midpoint, color = model)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
    geom_point(size = 3.5, alpha = 0.7) +
    ggrepel::geom_text_repel(aes(label = paper_short), size = 2.4, max.overlaps = 8, show.legend = FALSE) +
    scale_color_manual(values = MODEL_COLORS) +
    coord_cartesian(xlim = c(0, 100), ylim = c(0, 100)) +
    facet_wrap(~model, ncol = 2) +
    labs(x = "Human mean rating (0\u2013100)", y = "LLM rating (0\u2013100)", color = NULL) +
    theme_uj() +
    theme(strip.text = element_text(size = 13, face = "bold"))
} else {
  cat("No matching data for scatter plot.\n")
}
```

Both models cluster around the identity line in the 40--80 range but diverge more at the extremes. LLMs tend to compress ratings toward the centre of the scale relative to humans, producing narrower spread---a pattern consistent with alignment training that discourages extreme outputs. Where humans rate a paper very highly or very harshly, the LLM typically pulls toward the middle. Full agreement metrics including Pearson *r*, Spearman *ρ*, mean bias, and RMSE for all six models appear in @tbl-agreement.

**Rank comparison.** @fig-rankslope-main places human rankings in the centre column with GPT-5 Pro on the left and Claude Opus 4.6 on the right. Each paper appears as a pair of S-curves connecting its human rank to its rank under each LLM. Horizontal curves indicate agreement; curves that climb toward the top indicate the LLM ranked the paper higher (better) than humans did. Colour encodes the rank shift: green when humans rank it higher, orange/purple when the LLM does. The triptych layout makes it easy to spot papers where both LLMs agree with each other but disagree with humans (both curves slope the same way), versus papers where the two LLMs diverge.

```{r}
#| label: fig-rankslope-main
#| fig-cap: "Triptych slopegraph of paper rankings (overall criterion). Centre column shows human mean ranking; left and right columns show GPT-5 Pro and Claude Opus 4.6 rankings respectively. Each paper is connected by S-shaped curves; colour intensity encodes the magnitude of rank shift (\u0394rank). Green = humans rank it higher; warm colour = LLM ranks it higher. Horizontal curves indicate agreement."
#| fig-height: 12
#| fig-width: 14
#| code-fold: true

library("ggforce")

# --- Prepare ranking data for all three sources ---
rank_human <- human_avg |>
  filter(paper %in% matched_papers) |>
  arrange(desc(human_mid)) |>
  mutate(rank_h = row_number(), label_use = str_trunc(paper, 28)) |>
  select(paper, label_use, human_mid, rank_h)

rank_gpt <- llm_metrics |>
  filter(criteria == "overall", model == "GPT-5 Pro", paper %in% matched_papers) |>
  arrange(desc(midpoint)) |>
  mutate(rank_gpt = row_number()) |>
  select(paper, rank_gpt)

rank_claude <- llm_metrics |>
  filter(criteria == "overall", model == "Claude Opus 4.6", paper %in% matched_papers) |>
  arrange(desc(midpoint)) |>
  mutate(rank_claude = row_number()) |>
  select(paper, rank_claude)

rank_all <- rank_human |>
  inner_join(rank_gpt, by = "paper") |>
  inner_join(rank_claude, by = "paper") |>
  mutate(
    d_left  = rank_h - rank_gpt,    # positive = GPT ranks it higher
    d_right = rank_h - rank_claude   # positive = Claude ranks it higher
  )

n_papers_slope <- nrow(rank_all)

# --- Build bezier curves: Human (center, x=0.5) <-> GPT-5 Pro (left, x=0) ---
left_curves <- purrr::map_dfr(seq_len(n_papers_slope), function(i) {
  tibble(
    group = paste0("L", i),
    side  = "GPT-5 Pro",
    x = c(0.5, 0.35, 0.15, 0),
    y = c(rank_all$rank_h[i], rank_all$rank_h[i],
          rank_all$rank_gpt[i], rank_all$rank_gpt[i]),
    dr  = rank_all$d_left[i],
    mag = abs(rank_all$d_left[i])
  )
})

# --- Build bezier curves: Human (center, x=0.5) <-> Claude (right, x=1) ---
right_curves <- purrr::map_dfr(seq_len(n_papers_slope), function(i) {
  tibble(
    group = paste0("R", i),
    side  = "Claude Opus 4.6",
    x = c(0.5, 0.65, 0.85, 1),
    y = c(rank_all$rank_h[i], rank_all$rank_h[i],
          rank_all$rank_claude[i], rank_all$rank_claude[i]),
    dr  = rank_all$d_right[i],
    mag = abs(rank_all$d_right[i])
  )
})

all_curves <- bind_rows(left_curves, right_curves) |>
  mutate(alpha_val = pmin(mag / 5, 1))

if (n_papers_slope > 0) {
  ggplot() +
    # --- Bezier curves ---
    ggforce::geom_bezier(
      data = all_curves,
      aes(x = x, y = y, group = group, colour = dr, alpha = alpha_val),
      linewidth = 0.85
    ) +
    scale_color_gradient2(
      low = UJ_GREEN, mid = "grey88", high = UJ_ORANGE, midpoint = 0,
      name = "\u0394 rank  (Human \u2212 LLM)",
      breaks = c(-10, -5, 0, 5, 10),
      labels = c("\u221210", "\u22125", "0", "+5", "+10")
    ) +
    scale_alpha(range = c(0.30, 1), guide = "none") +

    # --- Column dots ---
    geom_point(data = rank_all, aes(x = 0,   y = rank_gpt),
               color = MODEL_COLORS["GPT-5 Pro"], size = 2.2) +
    geom_point(data = rank_all, aes(x = 0.5, y = rank_h),
               color = MODEL_COLORS["Human"], size = 2.6, shape = 16) +
    geom_point(data = rank_all, aes(x = 1,   y = rank_claude),
               color = MODEL_COLORS["Claude Opus 4.6"], size = 2.2) +

    # --- Paper labels (left side, next to GPT column) ---
    geom_text(data = rank_all,
              aes(x = 0, y = rank_gpt, label = label_use),
              hjust = 1.08, size = 2.6, color = "grey25") +

    # --- Rank numbers (right side, next to Claude column) ---
    geom_text(data = rank_all,
              aes(x = 1, y = rank_claude,
                  label = sprintf("H #%d  |  \u0394G %+d  |  \u0394C %+d",
                                  rank_h, d_left, d_right)),
              hjust = -0.06, size = 2.4, color = "grey35") +

    # --- Axes ---
    scale_x_continuous(
      breaks = c(0, 0.5, 1),
      labels = c("GPT-5 Pro", "Human", "Claude Opus 4.6"),
      limits = c(-0.65, 1.55),
      expand = expansion(mult = 0)
    ) +
    scale_y_reverse(expand = expansion(mult = c(0.02, 0.04))) +
    coord_cartesian(clip = "off") +
    labs(x = NULL, y = NULL) +
    theme_uj() +
    theme(
      axis.text.x      = element_text(size = 13, face = "bold"),
      axis.text.y       = element_blank(),
      axis.ticks.y      = element_blank(),
      panel.grid.major.y = element_blank(),
      panel.grid.major.x = element_line(linewidth = 0.35, color = "grey82"),
      plot.margin       = margin(t = 8, r = 180, b = 8, l = 140)
    ) +
    guides(colour = guide_colourbar(
      title.position = "top", barwidth = 14, barheight = 0.6
    ))
} else {
  cat("Insufficient data for triptych slopegraph.\\n")
}
```

Papers where both curves slope the same direction indicate systematic human--LLM disagreement; papers where the two curves diverge indicate model-specific "taste" differences. The criteria-level Krippendorff's alpha analysis in @tbl-human-baseline (a chance-corrected reliability coefficient that generalises across varying numbers of raters, missing data, and any measurement level; see [Methods](methods.qmd) for a formal definition) quantifies where human--LLM agreement approaches the ceiling set by human inter-rater variability.

**Rating differences by paper and criterion.** While the scatter plots and slopegraph summarise agreement on a single dimension, @fig-gap-heatmap-main unpacks Human − LLM differences across all seven criteria simultaneously. Each column is a paper, each row a criterion, and tile colour encodes the signed difference (human mean minus LLM midpoint). Green tiles indicate the human panel rated the paper higher; orange tiles indicate the LLM was more generous. Papers are sorted by GPT-5 Pro's overall difference so that both panels share the same column order, making model-specific patterns immediately visible.

```{r}
#| label: fig-gap-heatmap-main
#| fig-cap: "Human minus LLM rating difference for every paper (columns) and criterion (rows), comparing GPT-5 Pro (left panel) and Claude Opus 4.6 (right panel). Green tiles indicate the human mean was higher; orange tiles indicate the LLM rated the paper higher. Colour is clamped to \u00b130 percentile points. Papers are sorted by GPT-5 Pro overall difference so the two panels are directly comparable."
#| fig-width: 16
#| fig-height: 8
#| code-fold: true

library("patchwork")

metric_order <- c("overall", "claims", "methods", "adv_knowledge",
                  "logic_comms", "open_sci", "gp_relevance")
metric_lab <- c(
  overall = "Overall",
  claims = "Claims & Evidence",
  methods = "Methods",
  adv_knowledge = "Adv. Knowledge",
  logic_comms = "Logic & Comms",
  open_sci = "Open Science",
  gp_relevance = "Global Relevance"
)

H_mean <- metrics_human |>
  filter(criteria %in% metric_order, paper %in% matched_papers) |>
  group_by(paper, criteria) |>
  summarise(h = mean(mid, na.rm = TRUE), .groups = "drop")

# Fixed paper order based on GPT-5 Pro overall difference
L_mean_primary <- llm_metrics |>
  filter(criteria %in% metric_order, model == primary_model, paper %in% matched_papers) |>
  group_by(paper, criteria) |>
  summarise(l = mean(midpoint, na.rm = TRUE), .groups = "drop")

Ddiff_primary <- inner_join(H_mean, L_mean_primary, by = c("paper", "criteria")) |>
  mutate(diff = h - l)

ord_p <- Ddiff_primary |>
  filter(criteria == "overall") |>
  arrange(desc(diff)) |>
  pull(paper)

build_gap_panel <- function(model_name, show_y_labels = TRUE) {
  L_mean_m <- llm_metrics |>
    filter(criteria %in% metric_order, model == model_name, paper %in% matched_papers) |>
    group_by(paper, criteria) |>
    summarise(l = mean(midpoint, na.rm = TRUE), .groups = "drop")

  Ddiff_m <- inner_join(H_mean, L_mean_m, by = c("paper", "criteria")) |>
    mutate(
      diff = h - l,
      crit = factor(criteria, levels = metric_order, labels = metric_lab[metric_order])
    )

  p <- ggplot(Ddiff_m, aes(x = factor(paper, levels = ord_p), y = crit, fill = diff)) +
    geom_tile(color = "white", linewidth = 0.25) +
    scale_fill_gradient2(
      low = UJ_ORANGE, mid = "grey95", high = UJ_GREEN, midpoint = 0,
      name = "Human \u2212 LLM",
      limits = c(-30, 30),
      oob = scales::squish
    ) +
    labs(x = NULL, y = NULL, title = model_name) +
    theme_uj() +
    theme(
      axis.text.x = element_text(angle = 55, hjust = 1, vjust = 1, size = 8),
      panel.grid = element_blank()
    )

  if (show_y_labels) {
    p <- p + theme(axis.text.y = element_text(size = 11))
  } else {
    p <- p + theme(axis.text.y = element_blank())
  }

  p
}

if (length(ord_p) > 0) {
  p_left  <- build_gap_panel("GPT-5 Pro", show_y_labels = TRUE)
  p_right <- build_gap_panel("Claude Opus 4.6", show_y_labels = FALSE)

  p_left + p_right + plot_layout(guides = "collect") & theme(legend.position = "bottom")
} else {
  cat("Insufficient data for gap heatmap.\n")
}
```

Rows with uniformly green or orange tiles indicate papers where humans and the LLM disagree systematically across all criteria, not just on one dimension. Columns with consistent colour suggest criteria-level biases---for instance, if "Open Science" is green across nearly all papers, humans may systematically reward data-sharing practices more than the LLM does. Comparing the two panels reveals whether disagreement patterns are model-specific or shared across frontier LLMs; tiles that change colour between panels highlight papers where the models diverge from each other, not just from humans.

**Qualitative critique comparison.** Beyond numeric ratings, we compare the substantive critiques each model raises against the consensus issues identified by human experts. @fig-coverage-precision-main plots coverage---the fraction of human-identified concerns that the LLM also raised in some form---against precision---the fraction of LLM-raised issues that correspond to a substantive human concern. These metrics are assessed by GPT-5.2 Pro acting as an independent judge (see [Methods](methods.qmd) for the judging protocol).

```{r}
#| label: load-critique-data
#| code-fold: true
#| code-summary: "Load critique comparison data"
#| message: false

comparison_results_file <- c(
  "results/key_issue_comp_results.json",
  "results/key_issues_comparison_results.json"
)
comparison_results_file <- comparison_results_file[file.exists(comparison_results_file)][1]

has_critique_data <- FALSE

if (!is.na(comparison_results_file)) {
  comparison_file <- "results/key_issues_comparison.json"
  if (file.exists(comparison_file)) {
    comparison_data <- fromJSON(comparison_file)
    llm_results_raw <- fromJSON(comparison_results_file)
    llm_results <- llm_results_raw |>
      as_tibble() |>
      unnest_wider(comparison) |>
      select(
        gpt_paper,
        coverage_pct,
        precision_pct,
        any_of(c("matched_pairs", "unmatched_human", "unmatched_llm")),
        any_of(c("missed_issues", "extra_issues")),
        overall_rating,
        overall_justification,
        detailed_notes
      )
    comparison_data <- comparison_data |>
      left_join(llm_results, by = "gpt_paper")
    has_critique_data <- nrow(comparison_data) > 0 &&
      "coverage_pct" %in% names(comparison_data) &&
      any(!is.na(comparison_data$coverage_pct))
  }
}
```

```{r}
#| label: fig-coverage-precision-main
#| fig-cap: "Coverage (% of human-identified issues that the LLM also raised) vs Precision (% of LLM-raised issues that match a substantive human concern) for each paper, as assessed by GPT-5.2 Pro acting as judge. Dashed lines mark the cross-paper mean for each axis. Papers in the upper-right quadrant show strong alignment between human and LLM critiques."
#| fig-width: 10
#| fig-height: 7

if (has_critique_data) {
  crit_results <- comparison_data |>
    filter(!is.na(coverage_pct) & !is.na(precision_pct)) |>
    mutate(paper_short = str_trunc(gpt_paper, 25))

  ggplot(crit_results, aes(x = coverage_pct, y = precision_pct)) +
    geom_point(size = 4.5, color = UJ_ORANGE, alpha = 0.85) +
    geom_text(aes(label = paper_short), hjust = -0.1, vjust = 0.5, size = 2.6, check_overlap = TRUE) +
    geom_vline(xintercept = mean(crit_results$coverage_pct), linetype = "dashed", color = UJ_BLUE, linewidth = 0.6) +
    geom_hline(yintercept = mean(crit_results$precision_pct), linetype = "dashed", color = UJ_BLUE, linewidth = 0.6) +
    scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
    scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
    labs(
      x = "Coverage (%) \u2014 Human issues captured by LLM",
      y = "Precision (%) \u2014 LLM issues that are substantive"
    ) +
    theme_uj()
}
```

Coverage and precision vary substantially across papers. Some papers achieve high scores on both dimensions, indicating strong alignment between human and LLM critiques, while others reveal the LLM missing key human concerns or raising issues absent from the expert consensus. Because these metrics are themselves LLM-assessed, they should be interpreted with the caveat that an LLM judge may systematically over- or under-credit matches relative to a human annotator; manual validation through our annotation tool is underway and preliminary results are consistent with the automated scores.

Detailed paper-by-paper comparisons---including matched issue pairs with severity labels, structural difference tables, and per-evaluator breakdowns---appear in [Appendix B: Critiques & Key Issues](results_critiques.qmd). Extended quantitative analysis, including per-criterion correlations, bootstrap confidence intervals, mutual information, tier prediction accuracy, cost-quality trade-offs across all six models, and top disagreement cases, is reported in [Appendix A: Results Ratings](results_ratings.qmd). The full LLM reasoning traces and assessment summaries are available in [Appendix C: LLM Traces](appendix_llm_traces.qmd).
