---
title: "Results: Critiques & Key Issues"
format:
  html: default
engine: knitr
---

```{r}
#| label: setup-critiques
#| code-summary: "Setup and libraries"
#| code-fold: true
#| message: false
#| warning: false

library("tidyverse")
library("jsonlite")
library("knitr")
library("kableExtra")
library("DT")
library("reticulate")

# Theme colors (matching results_ratings.qmd)
UJ_ORANGE <- "#f19e4b"
UJ_GREEN  <- "#99bb66"
UJ_BLUE   <- "#4e79a7"

# Severity colors for badges
SEV_NECESSARY <- "#e74c3c"   # Red
SEV_OPTIONAL  <- "#f39c12"   # Orange
SEV_UNSURE    <- "#95a5a6"   # Gray

theme_uj <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot",
      legend.position = "bottom"
    )
}

# Function to create severity badge HTML
severity_badge <- function(severity) {
  sev <- tolower(trimws(severity))
  if (grepl("necessary", sev)) {
    return('<span style="background-color:#e74c3c;color:white;padding:2px 6px;border-radius:3px;font-size:0.8em;">🔴 Necessary</span>')
  } else if (grepl("optional", sev)) {
    return('<span style="background-color:#f39c12;color:white;padding:2px 6px;border-radius:3px;font-size:0.8em;">🟠 Optional</span>')
  } else if (sev != "") {
    return('<span style="background-color:#95a5a6;color:white;padding:2px 6px;border-radius:3px;font-size:0.8em;">⚪ Unsure</span>')
  }
  return('<span style="background-color:#bdc3c7;color:white;padding:2px 6px;border-radius:3px;font-size:0.8em;">—</span>')
}

# Function to truncate text for table display
truncate_text <- function(text, max_chars = 80) {
  if (nchar(text) > max_chars) {
    paste0(substr(text, 1, max_chars), "...")
  } else {
    text
  }
}
```

This chapter compares *qualitative critiques*: the key methodological and interpretive issues identified by the LLM model — against human expert critiques documented by Unjournal evaluation managers in Coda.

We first assess alignment using LLM-based comparison: GPT-5.2 Pro evaluates coverage (what proportion of human concerns GPT identified) and precision (whether GPT issues are substantive). Human-checks on this alignment are underway. See [LLM-Based Assessment](#sec-llm-assessment) for aggregate metrics.

## Data Sources

**GPT-5.2 Pro Key Issues**: Structured output from the focal evaluation run (January 2026), where the model was prompted to identify 5-15 key issues per paper, ordered from most to least important.

**Human Expert Critiques**: Curated content from the "Key critiques & issues with paper" column in The Unjournal's internal tracking database (Coda), written by evaluation managers synthesizing evaluator feedback. These use severity labels ("Necessary", "Optional but important", "Unsure") and cite specific evaluator comments.

```{r}
#| label: load-comparison-data
#| code-fold: true
#| code-summary: "Load comparison data"

# Load the matched comparison data
comparison_file <- "results/key_issues_comparison.json"
comparison_results_file <- c(
  "results/key_issue_comp_results.json",
  "results/key_issues_comparison_results.json"
)
comparison_results_file <- comparison_results_file[file.exists(comparison_results_file)][1]

if (file.exists(comparison_file)) {
  comparison_data <- fromJSON(comparison_file)
  n_papers <- nrow(comparison_data)
} else {
  comparison_data <- NULL
  n_papers <- 0
}

# Load LLM-based comparison results (coverage/precision) if available
if (!is.null(comparison_data) && !is.na(comparison_results_file)) {
  llm_results_raw <- fromJSON(comparison_results_file)
  llm_results <- llm_results_raw |>
    as_tibble() |>
    unnest_wider(comparison) |>
    select(
      gpt_paper,
      coverage_pct,
      precision_pct,
      # New format uses matched_pairs, unmatched_human, unmatched_llm
      any_of(c("matched_pairs", "unmatched_human", "unmatched_llm")),
      # Old format used missed_issues, extra_issues (keep for backward compat)
      any_of(c("missed_issues", "extra_issues")),
      overall_rating,
      overall_justification,
      detailed_notes
    )

  comparison_data <- comparison_data |>
    left_join(llm_results, by = "gpt_paper")
}

# Update paper count after loading
if (!is.null(comparison_data)) {
  n_papers <- nrow(comparison_data)
}

# Load human and LLM ratings for context tables
# Criterion name mapping: human -> standardized
criterion_map_human <- c(
  "overall" = "overall",
  "claims" = "claims_evidence",
  "methods" = "methods",
  "adv_knowledge" = "advancing_knowledge",
  "logic_comms" = "logic_communication",
  "open_sci" = "open_science",
  "gp_relevance" = "global_relevance",
  "real_world" = "real_world"
)

# Display names for nice labels
criterion_display <- c(
  "overall" = "Overall assessment",
  "claims_evidence" = "Claims & evidence",
  "methods" = "Methods",
  "advancing_knowledge" = "Advancing knowledge",
  "logic_communication" = "Logic & communication",
  "open_science" = "Open science",
  "global_relevance" = "Global relevance",
  "real_world" = "Real-world relevance"
)

# Normalize paper names for matching (remove underscores, periods, lowercase)
normalize_paper_name <- function(x) {
  x <- gsub("[_.]", " ", x)    # Replace underscores and periods with spaces
  x <- gsub("\\s+", " ", x)    # Collapse multiple spaces
  trimws(tolower(x))           # Lowercase and trim
}

human_ratings <- if (file.exists("data/all_ratings.rds")) {
  readRDS("data/all_ratings.rds") |>
    filter(criteria != "merits_journal", criteria != "journal_predict") |>
    mutate(
      criterion_std = criterion_map_human[as.character(criteria)],
      paper_norm = normalize_paper_name(label_paper)
    ) |>
    filter(!is.na(criterion_std)) |>
    group_by(paper = label_paper, paper_norm, criterion = criterion_std) |>
    summarise(human_mean = round(mean(middle_rating, na.rm = TRUE), 0), .groups = "drop")
} else {
  NULL
}

llm_ratings <- if (file.exists("data/metrics_long_gpt5_pro_jan2026.csv")) {
  read_csv("data/metrics_long_gpt5_pro_jan2026.csv", show_col_types = FALSE) |>
    mutate(paper_norm = normalize_paper_name(paper)) |>
    select(paper_llm = paper, paper_norm, criterion = metric, llm_rating = midpoint)
} else if (file.exists("data/metrics_long.csv")) {
  read_csv("data/metrics_long.csv", show_col_types = FALSE) |>
    mutate(paper_norm = normalize_paper_name(paper)) |>
    select(paper_llm = paper, paper_norm, criterion = metric, llm_rating = midpoint)
} else {
  NULL
}

# Combine ratings if both available - join on normalized paper name AND criterion
if (!is.null(human_ratings) && !is.null(llm_ratings)) {
  combined_ratings <- human_ratings |>
    full_join(llm_ratings, by = c("paper_norm", "criterion")) |>
    mutate(
      display_name = criterion_display[criterion],
      paper_final = coalesce(paper, paper_llm)
    ) |>
    select(paper = paper_final, paper_norm, criterion, display_name, human_mean, llm_rating)
} else {
  combined_ratings <- NULL
}
```

```{python}
#| label: parse-human-issues
#| code-fold: true
#| code-summary: "Parse human critiques into individual issues"

import json
import re
from pathlib import Path

# Severity parsing regex patterns
SEVERITY_HEADER_RE = re.compile(
    r"^(necessary|optional|optional but important|unsure|less important|possibly relevant|"
    r"probably important|probably correct|probably|note|notes|optional/desirable|"
    r"probably useful|maybe|possibly|unsure if correct|limitations)(\s|:|$)",
    re.IGNORECASE,
)
ENUM_RE = re.compile(r"^\s*(\d+)[\.\)]\s+")
EVALUATOR_RE = re.compile(r"^\[?(E\d+|Evaluator|DR|NotebookLM)\]?[\s:\[]", re.IGNORECASE)

def normalize_severity(header):
    h = header.lower()
    if "necessary" in h:
        return "necessary"
    if "optional" in h:
        return "optional"
    if any(word in h for word in ["unsure", "possibly", "probably", "less important", "maybe", "limitations"]):
        return "unsure"
    return ""

def extract_short_label(text, max_words=12):
    """Extract a descriptive short label from issue text."""
    original = text.strip()

    # Meta-commentary patterns to skip
    META_SKIP_PATTERNS = [
        r'^I chose',
        r'^I selected',
        r'^I think',
        r'^I would',
        r'^I believe',
        r'^These are',
        r'^This is (a|the|my)',
        r'^Note[s]?\s*[:)]',
        r'^DR\s*[:)]',
        r'^NotebookLM\s*[:)]',
        r'^The following',
        r'^Summary\s*[:)]',
        r'^Quote from',
        r'^See also',
        r'^As noted',
        r'^and that',  # Fragments starting with conjunctions
        r'^and it',
        r"^and that it's",
        r'^or that',
        r'^but that',
        r'^but it',
        r'^which ',
        r'^that ',
        r'^to be ',
        r'^it is ',
        r'^it would ',
    ]

    # First clean: Remove leading evaluator tags (DR —, [E1], etc.)
    cleaned_for_meta_check = re.sub(
        r'^[\s\[]*(?:E\d+|Evaluator\s*\d*|DR|NotebookLM|Treich|Triech|Capitan|Reiley|Manheim|Espinosa|Seinkmane|Cloud|Tsai)[\]:\s—\-]+',
        '', original, flags=re.IGNORECASE
    ).strip()

    # Check for meta-commentary patterns AFTER removing evaluator prefix
    for pattern in META_SKIP_PATTERNS:
        if re.match(pattern, cleaned_for_meta_check, re.IGNORECASE):
            # Try to find actual content after the meta-commentary
            # Look for a quoted phrase or substantive content after colon/period
            quote_in_meta = re.search(r'["""\']([^"""\'\n]{15,80})["""\']', cleaned_for_meta_check)
            if quote_in_meta:
                quoted = quote_in_meta.group(1).strip()
                if not re.match(r'^(the |a |an |this )', quoted.lower()):
                    return quoted
            # Look for content after a colon
            after_meta = re.split(r'[:\n]', cleaned_for_meta_check, maxsplit=1)
            if len(after_meta) > 1 and len(after_meta[1].strip()) > 20:
                original = after_meta[1].strip()
                break
            else:
                # Return a generic placeholder if no content found
                return "[Meta-commentary - see full text]"

    # FIRST: Check if text starts with "Label [Evaluator] more text" pattern
    # Extract just the label part before the evaluator tag
    # Match common evaluator patterns: [E1], [Treich], [Triech], [Reiley], [Capitan], etc.
    label_before_eval = re.match(r'^([A-Z][^[\n]{10,80}?)\s*\[(?:E\d+|[A-Z][a-z]+)\]', original)
    if label_before_eval:
        label = label_before_eval.group(1).strip()
        label = re.sub(r'[\s:,;—\-]+$', '', label)
        if 3 <= len(label.split()) <= max_words:
            return label

    # Also check for "Label E2: more text" pattern (evaluator without brackets)
    label_before_eval2 = re.match(r'^([A-Z][^\n]{10,80}?)\s+(?:E\d+|DR)\s*:', original)
    if label_before_eval2:
        label = label_before_eval2.group(1).strip()
        label = re.sub(r'[\s:,;—\-]+$', '', label)
        if 3 <= len(label.split()) <= max_words:
            return label

    # Remove evaluator tags, names, and quotes at start
    clean = re.sub(r'^\[?(?:E\d+|Evaluator\s*\d*|DR|NotebookLM|Treich|Triech|Capitan|Reiley|Manheim|Espinosa|Seinkmane|Cloud|Tsai)\]?[\s:\[]*', '', original, flags=re.IGNORECASE)
    clean = re.sub(r'^[""\'\[\(]', '', clean)
    clean = re.sub(r'^Author[\'s]*\s*response:?\s*', '', clean, flags=re.IGNORECASE)

    # Skip if remaining text starts with a name (single capitalized word followed by colon)
    name_check = re.match(r'^([A-Z][a-z]+)\s*:', clean)
    if name_check and len(name_check.group(1)) < 15:
        clean = clean[name_check.end():].strip()

    # Look for quoted phrases that are key terms (handle both straight and curly quotes)
    quote_match = re.search(r'["""\']([^"""\'\n]{15,80})["""\']', clean[:250])
    if quote_match:
        quoted = quote_match.group(1).strip()
        if not re.match(r'^(the |a |an |this )', quoted.lower()):
            return quoted

    # Look for "Label: description" or "Label — description" patterns
    # Use (?<!\w) to avoid matching hyphens in compound words like "pre-registered"
    label_match = re.match(r'^([^:\n]{10,80}?)(?:\s*[:\—]\s*|\s+—\s+)', clean)
    if label_match:
        label = label_match.group(1).strip()
        label = re.sub(r'[""\']+$', '', label)
        # Skip if it ends with a preposition or conjunction (incomplete phrase)
        if not re.search(r'\s+(or|and|of|for|to|in|on|by|with|the|a|an)$', label.lower()):
            if not re.match(r'^[A-Z][a-z]+$', label) and len(label.split()) <= max_words:
                return label

    # Look for descriptive noun phrases (allow hyphens in words)
    patterns = [
        r'((?:Lack of |Missing |Unclear |Weak |Strong |Potential )?[A-Z][a-z]+(?:[-\s]+[a-z-]+){1,5}(?:\s+(?:effects?|bias|issues?|concerns?|limitations?|assumptions?|analysis|validity|problems?|risks?|errors?|size|needs?|checking)))',
        r'((?:Small |Large |Limited |Significant )[a-z-]+(?:\s+[a-z-]+){0,5})',
        r'([A-Z][a-z]+\s+(?:sample|data|model|method|approach|design|analysis|claims?|effects?)(?:\s+[a-z-]+){0,3})',
        r'(Consider\s+[a-z-]+(?:[-\s]+[a-z-]+){0,5})',
        r'(Deviations?\s+from\s+(?:the\s+)?[a-z-]+(?:[-\s]+[a-z-]+){0,5})',
    ]
    for pattern in patterns:
        match = re.search(pattern, clean[:250])
        if match:
            label = match.group(1).strip()
            # Remove trailing preposition phrases like "or degrees of", "and more", etc.
            while True:
                new_label = re.sub(r'\s+(or|and|with|for|to|of|in|on|by|including)(\s+\w+)*\s*$', '', label)
                if new_label == label:
                    break
                label = new_label
            if 10 < len(label) < 120 and len(label.split()) <= max_words:
                return label

    # Take first sentence or clause
    first_clause = re.split(r'(?<=[.!?])\s+|\n', clean)[0].strip()
    first_clause = re.sub(r'^(The |A |An )', '', first_clause)
    first_clause = re.sub(r'\s+(including|or|and|with|for|to|of|in|on|by)\s*$', '', first_clause)

    # Remove evaluator tags from middle/end of text
    first_clause = re.sub(r'\s*\[(?:E\d+|[A-Z][a-z]+)\]\s*', ' ', first_clause)
    first_clause = re.sub(r'\s+(?:E\d+|DR)\s*:.*$', '', first_clause)
    first_clause = re.sub(r'\s+', ' ', first_clause).strip()

    words = first_clause.split()
    if 3 <= len(words) <= max_words:
        return first_clause
    elif len(words) > max_words:
        text = ' '.join(words[:max_words])
        last_break = max(text.rfind(' or '), text.rfind(' and '), text.rfind(', '), text.rfind(' — '))
        if last_break > len(text) // 2:
            return text[:last_break].strip(' ,—')
        return text + '...'

    # Fallback
    key_terms = re.findall(r'[A-Z][a-z]+(?:\s+[a-z]+){0,2}', clean[:150])
    if key_terms:
        return ' / '.join(key_terms[:3])

    words = clean.split()[:max_words]
    return ' '.join(words) + ('...' if len(clean.split()) > max_words else '')

def extract_llm_label(text, max_words=10):
    """Extract a short label from LLM issue text."""
    # Clean up numbering
    clean = re.sub(r'^[\d\.\)]+\s*', '', text.strip())

    # LLM issues often have format "Topic: detailed explanation"
    colon_match = re.match(r'^([^:]{8,60}):\s*', clean)
    if colon_match:
        label = colon_match.group(1).strip()
        # Make sure it's descriptive enough
        if len(label.split()) >= 2:
            return label

    # Take first clause
    first_part = re.split(r'[;.]', clean)[0].strip()
    words = first_part.split()
    if len(words) <= max_words:
        return first_part
    return ' '.join(words[:max_words]) + '...'

def split_human_issues(text):
    """Parse human critique text into individual issues with severity labels."""
    if not text:
        return []

    issues = []
    current_severity = ""
    current_issue_lines = []

    def flush_issue():
        nonlocal current_issue_lines
        if not current_issue_lines:
            return
        full_text = "\n".join(current_issue_lines).strip()
        # Clean leading punctuation/numbering
        cleaned = re.sub(r"^[\-\*\d\.\)\s]+", "", full_text).strip()
        if not cleaned or len(cleaned) < 10:
            current_issue_lines = []
            return
        # Skip if it's just a severity label
        if SEVERITY_HEADER_RE.match(cleaned):
            current_issue_lines = []
            return

        short_label = extract_short_label(cleaned)
        # Skip if this is meta-commentary (the label extraction will return placeholder)
        if short_label == "[Meta-commentary - see full text]":
            current_issue_lines = []
            return

        issues.append({
            "text": cleaned,
            "severity": current_severity,
            "short_label": short_label
        })
        current_issue_lines = []

    lines = text.splitlines()
    i = 0
    while i < len(lines):
        line = lines[i].rstrip()
        raw = line.strip()

        # Check for severity header
        if raw and SEVERITY_HEADER_RE.match(raw):
            flush_issue()
            current_severity = normalize_severity(raw.split(':')[0].split()[0])
            i += 1
            continue

        # Check for numbered issue start
        if ENUM_RE.match(raw):
            flush_issue()
            # Remove the number prefix
            issue_text = re.sub(r"^\s*\d+[\.\)]\s*", "", raw)
            current_issue_lines = [issue_text] if issue_text else []
            i += 1
            continue

        # Check for issue label line (short line followed by longer explanation)
        # e.g., "Ceiling effects with Likert scale measures"
        if raw and not EVALUATOR_RE.match(raw) and len(raw) < 80:
            # Look ahead - if next non-empty line starts with [E or is indented, this is a label
            next_idx = i + 1
            while next_idx < len(lines) and not lines[next_idx].strip():
                next_idx += 1
            if next_idx < len(lines):
                next_line = lines[next_idx].strip()
                if (EVALUATOR_RE.match(next_line) or
                    lines[next_idx].startswith(' ') or
                    next_line.startswith('"') or
                    next_line.startswith("'")):
                    # This short line is likely a label for a new issue
                    flush_issue()
                    current_issue_lines = [raw]
                    i += 1
                    continue

        # Empty line - might signal new issue in some formats
        if not raw:
            # Don't flush on every empty line - only if we have significant content
            if current_issue_lines and len("\n".join(current_issue_lines)) > 100:
                # Check if next content looks like new issue
                next_idx = i + 1
                while next_idx < len(lines) and not lines[next_idx].strip():
                    next_idx += 1
                if next_idx < len(lines):
                    next_line = lines[next_idx].strip()
                    if ENUM_RE.match(next_line) or (len(next_line) < 80 and not EVALUATOR_RE.match(next_line)):
                        flush_issue()
            i += 1
            continue

        # Regular content line - add to current issue
        if current_issue_lines:
            current_issue_lines.append(raw)
        else:
            current_issue_lines = [raw]
        i += 1

    flush_issue()
    return issues

# Load comparison data and results
comparison_file = Path("results/key_issues_comparison.json")
results_file = Path("results/key_issues_comparison_results.json")

parsed_issues = {}
comparison_results = {}

if comparison_file.exists():
    comparison = json.loads(comparison_file.read_text())
    for item in comparison:
        paper_id = item.get("gpt_paper", "")
        coda_critique = item.get("coda_critique", "")
        parsed_issues[paper_id] = split_human_issues(coda_critique)

if results_file.exists():
    results = json.loads(results_file.read_text())
    for r in results:
        paper_id = r.get("gpt_paper", "")
        comparison_results[paper_id] = r.get("comparison", {})
```

We matched **`r n_papers` papers** with both GPT-5.2 Pro key issues and human expert critiques.

## Overview Statistics

```{r}
#| label: tbl-overview-stats
#| tbl-cap: "Summary of key issues comparison data"

if (!is.null(comparison_data) && n_papers > 0) {
  overview <- comparison_data |>
    mutate(
      num_gpt_issues = sapply(gpt_key_issues, length)
    ) |>
    summarise(
      `Papers Compared` = n(),
      `Avg GPT Issues per Paper` = round(mean(num_gpt_issues), 1),
      `Min GPT Issues` = min(num_gpt_issues),
      `Max GPT Issues` = max(num_gpt_issues),
      `Avg Coda Critique Length (chars)` = round(mean(coda_critique_length)),
      `Min Coda Length` = min(coda_critique_length),
      `Max Coda Length` = max(coda_critique_length)
    ) |>
    pivot_longer(everything(), names_to = "Metric", values_to = "Value")

  kable(overview, align = c("l", "r"))
}
```

## Paper-by-Paper Comparison {#sec-paper-comparison}

Each paper section shows:

-   **Assessment badge**: Coverage and precision metrics with overall rating
-   **LLM Assessment Summary**: GPT-5.2 Pro's justification of the comparison
-   **Missed Issues**: Human expert issues that GPT failed to capture
-   **Extra Issues**: GPT issues not present in human critique
-   **Full texts**: Expandable sections with complete human critique and GPT issues

```{r}
#| label: render-comparisons
#| results: asis
#| code-fold: true
#| code-summary: "Render paper comparisons"

# Access parsed issues and comparison results from Python
parsed_issues_list <- py$parsed_issues
comparison_results_list <- py$comparison_results

if (!is.null(comparison_data) && n_papers > 0) {

  for (i in 1:nrow(comparison_data)) {
    row <- comparison_data[i, ]
    paper_id <- row$gpt_paper

    # Paper header
    cat(sprintf("\n### %s\n\n", paper_id))
    cat(sprintf("**Coda title:** %s\n\n", row$coda_title))

    # Get comparison results for this paper
    paper_comparison <- comparison_results_list[[paper_id]]
    gpt_issues <- row$gpt_key_issues[[1]]

    # Show LLM assessment badge
    if (!is.na(row$coverage_pct)) {
      coverage_val <- as.numeric(row$coverage_pct)
      precision_val <- as.numeric(row$precision_pct)
      avg_score <- (coverage_val + precision_val) / 2
      rating <- if (!is.null(row$overall_rating)) row$overall_rating else "Unknown"

      # Calculate weighted coverage: mean of match_quality across all human issues (0 for unmatched)
      matched_pairs_data <- if (!is.null(paper_comparison$matched_pairs)) paper_comparison$matched_pairs else list()
      unmatched_human_data <- if (!is.null(paper_comparison$unmatched_human)) paper_comparison$unmatched_human else list()
      n_matched <- length(matched_pairs_data)
      n_unmatched <- length(unmatched_human_data)
      n_total_human <- n_matched + n_unmatched

      if (n_total_human > 0) {
        # Sum of match_quality for matched issues + 0 for each unmatched issue
        match_qualities <- sapply(matched_pairs_data, function(x) as.numeric(x$match_quality))
        weighted_coverage <- round(sum(match_qualities) / n_total_human, 0)
      } else {
        weighted_coverage <- NA
      }

      # Determine badge color based on rating
      if (rating %in% c("Good", "Excellent")) {
        badge_class <- "callout-tip"
      } else if (rating == "Moderate") {
        badge_class <- "callout-note"
      } else {
        badge_class <- "callout-warning"
      }

      cat(sprintf("::: {.%s appearance=\"simple\"}\n", badge_class))
      if (!is.na(weighted_coverage)) {
        cat(sprintf("**Coverage:** %s%% | **Weighted Coverage:** %s%% | **Precision:** %s%%\n",
                    row$coverage_pct, weighted_coverage, row$precision_pct))
      } else {
        cat(sprintf("**Coverage:** %s%% | **Precision:** %s%%\n",
                    row$coverage_pct, row$precision_pct))
      }
      cat(":::\n\n")
    }

    # ---- RATINGS CONTEXT TABLE (collapsible) ----
    if (!is.null(combined_ratings)) {
      # Match paper using normalized name for exact matching
      paper_id_norm <- normalize_paper_name(paper_id)
      paper_ratings <- combined_ratings |>
        filter(paper_norm == paper_id_norm) |>
        filter(!is.na(display_name)) |>  # Only rows with valid display names
        arrange(criterion != "overall", display_name)  # Overall first, then alphabetical

      if (nrow(paper_ratings) > 0) {
        cat('<details style="margin-bottom:1em;">\n')
        cat('<summary style="font-size:0.9em;color:#666;cursor:pointer;">📊 Human vs LLM ratings for this paper</summary>\n')
        cat('<div style="margin-top:0.5em;overflow-x:auto;">\n')
        cat('<table style="font-size:0.85em;border-collapse:collapse;width:100%;">\n')
        cat('<tr style="background:#f5f5f5;"><th style="padding:6px 12px;text-align:left;">Criterion</th><th style="padding:6px 12px;text-align:center;">Human (avg)</th><th style="padding:6px 12px;text-align:center;">LLM (mid)</th></tr>\n')

        for (j in 1:nrow(paper_ratings)) {
          r <- paper_ratings[j, ]
          h_val <- if (!is.na(r$human_mean)) as.character(r$human_mean) else "—"
          l_val <- if (!is.na(r$llm_rating)) as.character(round(r$llm_rating, 0)) else "—"
          crit_name <- r$display_name
          cat(sprintf('<tr><td style="padding:6px 12px;">%s</td><td style="padding:6px 12px;text-align:center;">%s</td><td style="padding:6px 12px;text-align:center;">%s</td></tr>\n',
                      crit_name, h_val, l_val))
        }
        cat('</table>\n')
        cat('<div style="font-size:0.75em;color:#888;margin-top:0.25em;">Human = avg of evaluator midpoints; LLM = model midpoint rating</div>\n')
        cat('</div>\n')
        cat('</details>\n\n')
      }
    }

    # ---- LLM ASSESSMENT SUMMARY ----
    if (!is.null(row$overall_justification) && nchar(row$overall_justification) > 0) {
      cat("#### LLM Assessment Summary\n\n")
      cat(row$overall_justification)
      cat("\n\n")
    }

    # ---- DETAILED NOTES (expandable) ----
    if (!is.null(row$detailed_notes) && nchar(row$detailed_notes) > 0) {
      cat("::: {.callout-note collapse='true'}\n")
      cat("## Detailed Comparison Notes\n\n")
      cat(row$detailed_notes)
      cat("\n:::\n\n")
    }

    # ---- MATCHED PAIRS (New format from enhanced LLM comparison) ----
    matched_pairs <- if (!is.null(paper_comparison$matched_pairs)) paper_comparison$matched_pairs else list()

    if (length(matched_pairs) > 0) {
      cat(sprintf("#### ✅ Matched Issues (%d human issues with LLM coverage)\n\n", length(matched_pairs)))

      for (j in seq_along(matched_pairs)) {
        pair <- matched_pairs[[j]]
        label <- if (!is.null(pair$label)) pair$label else "Unnamed issue"
        match_quality <- if (!is.null(pair$match_quality)) pair$match_quality else "?"
        match_explanation <- if (!is.null(pair$match_explanation)) pair$match_explanation else ""
        detailed_discussion <- if (!is.null(pair$detailed_discussion)) pair$detailed_discussion else ""
        human_idx <- if (!is.null(pair$human_issue_index)) pair$human_issue_index else j
        llm_indices <- if (!is.null(pair$llm_issue_indices)) pair$llm_issue_indices else c()

        # Match quality color
        if (is.numeric(match_quality)) {
          if (match_quality >= 70) {
            quality_color <- "#27ae60"  # Green
          } else if (match_quality >= 40) {
            quality_color <- "#f39c12"  # Orange
          } else {
            quality_color <- "#e74c3c"  # Red
          }
        } else {
          quality_color <- "#95a5a6"  # Gray
        }

        # Each matched issue is collapsible to reduce clutter
        cat(sprintf('<details style="margin:0.75em 0;padding:0;background:#f0fff0;border-left:4px solid %s;border-radius:4px;">\n', quality_color))
        llm_tags <- if (length(llm_indices) > 0) paste(sapply(llm_indices, function(x) sprintf("L%d", x)), collapse = ", ") else ""
        cat(sprintf('<summary style="padding:0.75em;cursor:pointer;display:flex;justify-content:space-between;align-items:center;">\n'))
        cat(sprintf('<span><strong style="font-size:1.05em;">H%d: %s</strong> <span style="font-size:0.85em;color:#666;">↔ %s</span></span>\n',
                    human_idx, gsub("<", "&lt;", gsub(">", "&gt;", label)), llm_tags))
        cat(sprintf('<span style="background-color:%s;color:white;padding:3px 8px;border-radius:12px;font-weight:bold;font-size:0.9em;">%s%%</span>\n', quality_color, match_quality))
        cat('</summary>\n')
        cat('<div style="padding:0 0.75em 0.75em 0.75em;">\n')

        if (nchar(match_explanation) > 0) {
          cat(sprintf('<div style="font-size:0.9em;margin-bottom:0.5em;">%s</div>\n', gsub("<", "&lt;", gsub(">", "&gt;", match_explanation))))
        }

        # ---- SHOW ACTUAL HUMAN ISSUE TEXT ----
        # First check if human_issue_text is directly available in data
        human_text <- if (!is.null(pair$human_issue_text)) pair$human_issue_text else NULL

        # If not, try to extract from coda_critique using label keywords
        if (is.null(human_text) || nchar(human_text) == 0) {
          # Extract keywords from label (split on / and space, take significant words)
          keywords <- unlist(strsplit(tolower(label), "[/ ]+"))
          keywords <- keywords[nchar(keywords) > 4]  # Only words > 4 chars

          # Search coda_critique for lines containing these keywords
          coda_lines <- unlist(strsplit(row$coda_critique, "\n"))
          for (kw in keywords[1:min(3, length(keywords))]) {
            # Use fixed matching to avoid regex issues with special chars
            matches <- coda_lines[grepl(kw, tolower(coda_lines), fixed = TRUE)]
            if (length(matches) > 0) {
              human_text <- paste(matches[1:min(2, length(matches))], collapse = " ")
              break
            }
          }
        }

        if (!is.null(human_text) && nchar(human_text) > 10) {
          cat('<details style="margin-top:0.75em;background:#fff3e0;padding:0.5em;border-radius:4px;" open>\n')
          cat('<summary style="font-size:0.9em;color:#e65100;cursor:pointer;font-weight:500;">👤 Human expert critique text</summary>\n')
          cat(sprintf('<div style="margin-top:0.5em;padding:0.5em;background:white;border-left:3px solid #e65100;font-size:0.85em;">%s</div>\n',
                      gsub("<", "&lt;", gsub(">", "&gt;", human_text))))
          cat('</details>\n')
        }

        # ---- SHOW ACTUAL LLM ISSUE TEXT(S) ----
        if (length(llm_indices) > 0 && length(gpt_issues) > 0) {
          cat('<details style="margin-top:0.75em;background:#e8f4fd;padding:0.5em;border-radius:4px;" open>\n')
          cat('<summary style="font-size:0.9em;color:#1565c0;cursor:pointer;font-weight:500;">🤖 Matched LLM issue text</summary>\n')
          cat('<div style="margin-top:0.5em;">\n')
          for (idx in llm_indices) {
            if (idx <= length(gpt_issues)) {
              llm_text <- gpt_issues[idx]
              cat(sprintf('<div style="margin:0.5em 0;padding:0.5em;background:white;border-left:3px solid #1565c0;font-size:0.85em;">\n'))
              cat(sprintf('<strong style="color:#1565c0;">L%d:</strong> %s\n', idx, gsub("<", "&lt;", gsub(">", "&gt;", llm_text))))
              cat('</div>\n')
            }
          }
          cat('</div>\n')
          cat('</details>\n')
        }

        if (nchar(detailed_discussion) > 0) {
          cat('<details style="margin-top:0.5em;">\n')
          cat('<summary style="font-size:0.85em;color:#666;cursor:pointer;">Show detailed discussion</summary>\n')
          cat(sprintf('<div style="margin-top:0.5em;padding:0.5em;background:#fafafa;font-size:0.85em;border-radius:4px;">%s</div>\n',
                      gsub("<", "&lt;", gsub(">", "&gt;", detailed_discussion))))
          cat('</details>\n')
        }

        cat('</div>\n')  # Close content wrapper
        cat('</details>\n')  # Close collapsible matched issue
      }
      cat('\n')
    }

    # ---- UNMATCHED HUMAN ISSUES (New format, collapsible) ----
    unmatched_human <- if (!is.null(paper_comparison$unmatched_human)) paper_comparison$unmatched_human else list()

    if (length(unmatched_human) > 0) {
      cat('<details style="margin-bottom:1em;">\n')
      cat(sprintf('<summary style="color:#c62828;font-weight:bold;cursor:pointer;">⚠️ Unmatched Human Issues (%d not captured by LLM)</summary>\n', length(unmatched_human)))
      cat('<div style="margin-top:0.5em;">\n')

      for (j in seq_along(unmatched_human)) {
        item <- unmatched_human[[j]]
        idx <- if (!is.null(item$index)) item$index else j
        brief <- if (!is.null(item$brief_description)) item$brief_description else "No description"
        why_missed <- if (!is.null(item$why_missed)) item$why_missed else ""

        cat(sprintf('<div style="margin:0.5em 0;padding:0.75em;background:#ffebee;border-left:4px solid #c62828;font-size:0.9em;">\n'))
        cat(sprintf('<strong>H%d:</strong> %s\n', idx, gsub("<", "&lt;", gsub(">", "&gt;", brief))))
        if (nchar(why_missed) > 0) {
          cat(sprintf('<div style="margin-top:0.5em;font-size:0.85em;color:#666;"><em>Why missed:</em> %s</div>\n',
                      gsub("<", "&lt;", gsub(">", "&gt;", why_missed))))
        }
        cat('</div>\n')
      }
      cat('</div>\n')
      cat('</details>\n\n')
    }

    # ---- UNMATCHED LLM ISSUES (New format) ----
    unmatched_llm <- if (!is.null(paper_comparison$unmatched_llm)) paper_comparison$unmatched_llm else list()

    if (length(unmatched_llm) > 0) {
      cat('<details style="margin-bottom:1.5em;">\n')
      cat(sprintf('<summary style="color:#1565c0;font-weight:bold;cursor:pointer;">📋 Unmatched LLM Issues (%d not in human critique)</summary>\n', length(unmatched_llm)))
      cat('<div style="margin-top:0.5em;">\n')

      for (j in seq_along(unmatched_llm)) {
        item <- unmatched_llm[[j]]
        idx <- if (!is.null(item$index)) item$index else j
        brief <- if (!is.null(item$brief_description)) item$brief_description else "No description"
        why_extra <- if (!is.null(item$why_extra)) item$why_extra else ""

        cat(sprintf('<div style="margin:0.5em 0;padding:0.5em;background:#e3f2fd;border-left:4px solid #1565c0;font-size:0.9em;">\n'))
        cat(sprintf('<strong>L%d:</strong> %s\n', idx, gsub("<", "&lt;", gsub(">", "&gt;", brief))))
        if (nchar(why_extra) > 0) {
          cat(sprintf('<div style="margin-top:0.5em;font-size:0.85em;color:#666;"><em>Why extra:</em> %s</div>\n',
                      gsub("<", "&lt;", gsub(">", "&gt;", why_extra))))
        }
        cat('</div>\n')
      }
      cat('</div>\n')
      cat('</details>\n\n')
    }

    # ---- FALLBACK: Old format missed/extra issues (if new format not available) ----
    if (length(matched_pairs) == 0 && length(unmatched_human) == 0) {
      # Use old format
      missed_issues <- if (!is.null(paper_comparison$missed_issues)) paper_comparison$missed_issues else character(0)

      if (length(missed_issues) > 0) {
        cat(sprintf("#### ⚠️ Issues Human Experts Raised That GPT Missed (%d)\n\n", length(missed_issues)))
        for (j in seq_along(missed_issues)) {
          issue_text <- missed_issues[[j]]
          issue_escaped <- gsub("<", "&lt;", gsub(">", "&gt;", issue_text))
          cat(sprintf('<div style="margin:0.5em 0;padding:0.75em;background:#ffebee;border-left:4px solid #c62828;font-size:0.9em;">\n'))
          cat(sprintf('%s\n', issue_escaped))
          cat('</div>\n')
        }
        cat('\n')
      }

      extra_issues <- if (!is.null(paper_comparison$extra_issues)) paper_comparison$extra_issues else character(0)

      if (length(extra_issues) > 0) {
        cat('<details style="margin-bottom:1.5em;">\n')
        cat(sprintf('<summary style="color:#1565c0;font-weight:bold;cursor:pointer;">📋 Extra Issues GPT Identified (%d not in human critique)</summary>\n', length(extra_issues)))
        cat('<div style="margin-top:0.5em;">\n')
        for (j in seq_along(extra_issues)) {
          issue_text <- extra_issues[[j]]
          issue_escaped <- gsub("<", "&lt;", gsub(">", "&gt;", issue_text))
          cat(sprintf('<div style="margin:0.5em 0;padding:0.5em;background:#e3f2fd;border-left:4px solid #1565c0;font-size:0.9em;">\n'))
          cat(sprintf('%s\n', issue_escaped))
          cat('</div>\n')
        }
        cat('</div>\n')
        cat('</details>\n\n')
      }
    }

    # ---- FULL HUMAN CRITIQUE (expandable) ----
    cat('<details style="margin-bottom:1em;">\n')
    cat('<summary style="color:#e65100;cursor:pointer;">📄 View full human critique</summary>\n')
    cat('<div style="margin-top:0.5em;padding:0.75em;background:#fffbf0;border:1px solid #ffe0b2;border-radius:4px;font-size:0.9em;white-space:pre-wrap;">\n')
    cat(gsub("<", "&lt;", gsub(">", "&gt;", row$coda_critique)))
    cat('\n</div>\n')
    cat('</details>\n\n')

    # ---- FULL GPT ISSUES (expandable) ----
    if (length(gpt_issues) > 0) {
      cat('<details style="margin-bottom:1em;">\n')
      cat(sprintf('<summary style="color:#1565c0;cursor:pointer;">🤖 View all GPT key issues (%d)</summary>\n', length(gpt_issues)))
      cat('<div style="margin-top:0.5em;">\n')
      for (j in seq_along(gpt_issues)) {
        issue <- gpt_issues[j]
        issue_clean <- gsub("^[0-9]+[)\\.]\\s*", "", issue)
        issue_clean <- gsub("^\\. ", "", issue_clean)
        issue_escaped <- gsub("<", "&lt;", gsub(">", "&gt;", issue_clean))
        cat(sprintf('<div style="margin:0.5em 0;padding:0.5em;background:#f0f7ff;border-left:4px solid #4e79a7;font-size:0.9em;">\n'))
        cat(sprintf('<strong>%d.</strong> %s\n', j, issue_escaped))
        cat('</div>\n')
      }
      cat('</div>\n')
      cat('</details>\n\n')
    }

    cat("---\n\n")
  }
}
```

## LLM-Based Assessment {#sec-llm-assessment}

The comparison between GPT key issues and human critiques was assessed using GPT-5.2 Pro, which evaluated coverage (what proportion of human concerns GPT identified) and precision (whether GPT issues are substantive).

-   **Coverage**: Proportion of *consensus* human issues that have any LLM match (match quality ≥ 30%)
-   **Weighted Coverage**: Mean match quality across all human issues (treating unmatched issues as 0%)
-   **Precision**: Proportion of LLM issues that match a human concern

**Note**: Any interpretation or narrative commentary in this section is written by an LLM (Codex). The values themselves come from the GPT-5.2 Pro comparison outputs.

::: callout-note
## Caveat on Precision

"Precision" only reflects whether LLM issues match the *consensus* human issues curated in Coda—those prioritized by evaluation managers as the most important concerns. Many LLM-identified issues may have been noted by one or more individual human evaluators but were not included in this curated set. A low precision score does not necessarily mean the LLM raised irrelevant concerns; it may simply reflect issues that weren't prioritized in the consensus summary.
:::

```{r}
#| label: check-llm-results
#| code-fold: true

# Check if LLM comparison results are available
has_llm_results <- !is.null(comparison_data) &&
                   n_papers > 0 &&
                   "coverage_pct" %in% names(comparison_data) &&
                   any(!is.na(comparison_data$coverage_pct))
```

```{r}
#| label: llm-assessment-summary
#| tbl-cap: "LLM assessment of GPT vs human critique alignment"
#| eval: !expr has_llm_results

if (has_llm_results) {
  # Filter to papers with valid LLM results
  llm_results <- comparison_data |>
    filter(!is.na(coverage_pct) & !is.na(precision_pct))

  summary_stats <- llm_results |>
    summarise(
      `Papers Assessed` = n(),
      `Mean Coverage (%)` = round(mean(coverage_pct, na.rm = TRUE), 1),
      `Mean Precision (%)` = round(mean(precision_pct, na.rm = TRUE), 1),
      `Coverage Range` = paste0(min(coverage_pct, na.rm = TRUE), "-", max(coverage_pct, na.rm = TRUE)),
      `Precision Range` = paste0(min(precision_pct, na.rm = TRUE), "-", max(precision_pct, na.rm = TRUE))
    ) |>
    mutate(across(everything(), as.character)) |>
    pivot_longer(everything(), names_to = "Metric", values_to = "Value")

  kable(summary_stats, align = c("l", "r"))
}
```

```{r}
#| label: fig-coverage-precision
#| fig-cap: "Coverage vs Precision across papers (LLM-assessed)"
#| fig-width: 9
#| fig-height: 7
#| eval: !expr has_llm_results

if (has_llm_results) {
  llm_results <- comparison_data |>
    filter(!is.na(coverage_pct) & !is.na(precision_pct)) |>
    mutate(paper_short = str_trunc(gpt_paper, 25))

  ggplot(llm_results, aes(x = coverage_pct, y = precision_pct)) +
    geom_point(size = 4, color = UJ_ORANGE, alpha = 0.8) +
    geom_text(aes(label = paper_short), hjust = -0.1, vjust = 0.5, size = 2.5, check_overlap = TRUE) +
    geom_vline(xintercept = mean(llm_results$coverage_pct), linetype = "dashed", color = UJ_BLUE) +
    geom_hline(yintercept = mean(llm_results$precision_pct), linetype = "dashed", color = UJ_BLUE) +
    scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
    scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
    labs(
      x = "Coverage (%): Human issues captured by GPT",
      y = "Precision (%): GPT issues that are substantive",
      subtitle = "Dashed lines = means"
    ) +
    theme_uj()
}
```

```{r}
#| label: tbl-paper-ratings
#| tbl-cap: "Per-paper LLM assessment results"
#| eval: !expr has_llm_results

if (has_llm_results) {
  ratings_table <- comparison_data |>
    filter(!is.na(coverage_pct)) |>
    mutate(
      Paper = str_trunc(gpt_paper, 35),
      `Coverage (%)` = coverage_pct,
      `Precision (%)` = precision_pct,
      Rating = overall_rating
    ) |>
    select(Paper, `Coverage (%)`, `Precision (%)`, Rating) |>
    arrange(desc(`Coverage (%)`))

  kable(ratings_table, align = c("l", "r", "r", "l"))
}
```

```{r}
#| label: fig-rating-distribution
#| fig-cap: "Distribution of overall ratings across papers"
#| fig-width: 8
#| fig-height: 4
#| eval: !expr has_llm_results

if (has_llm_results) {
  rating_counts <- comparison_data |>
    filter(!is.na(overall_rating)) |>
    mutate(
      Rating = factor(overall_rating, levels = c("Poor", "Moderate", "Good", "Excellent"))
    ) |>
    count(Rating, .drop = FALSE)

  # Define colors for ratings
  rating_colors <- c("Poor" = "#e74c3c", "Moderate" = "#f39c12", "Good" = "#27ae60", "Excellent" = "#2ecc71")

  ggplot(rating_counts, aes(x = Rating, y = n, fill = Rating)) +
    geom_col(alpha = 0.9) +
    geom_text(aes(label = n), vjust = -0.3, size = 5) +
    scale_fill_manual(values = rating_colors, drop = FALSE) +
    labs(
      x = "Overall Rating",
      y = "Number of Papers"
    ) +
    theme_uj() +
    theme(legend.position = "none")
}
```

::: {.callout-note collapse="true"}
## Interpretation Guide

-   **Coverage**: Percentage of human-identified issues that GPT also captured (in some form). Higher = GPT missed fewer human concerns.
-   **Precision**: Percentage of GPT issues that are substantive rather than generic or spurious. Higher = GPT's critiques are more targeted.
-   **Overall Rating**: Qualitative assessment (Excellent/Good/Moderate/Poor) based on both coverage and precision metrics.
:::

## Summary and Observations

```{r}
#| label: summary-analysis
#| code-fold: true

if (!is.null(comparison_data) && n_papers > 0) {

  # Calculate some summary statistics
  issue_counts <- comparison_data |>
    mutate(
      num_gpt_issues = sapply(gpt_key_issues, length),
      paper_short = str_trunc(gpt_paper, 30)
    ) |>
    select(paper_short, num_gpt_issues, coda_critique_length)

  # Calculate severity distribution from parsed issues
  all_human_issues <- do.call(rbind, lapply(names(parsed_issues_list), function(paper_id) {
    issues <- parsed_issues_list[[paper_id]]
    if (is.null(issues) || length(issues) == 0) return(NULL)
    data.frame(
      paper_id = paper_id,
      severity = sapply(issues, function(x) x$severity),
      stringsAsFactors = FALSE
    )
  }))

  if (!is.null(all_human_issues) && nrow(all_human_issues) > 0) {
    severity_dist <- all_human_issues |>
      mutate(
        severity = case_when(
          grepl("necessary", tolower(severity)) ~ "Necessary",
          grepl("optional", tolower(severity)) ~ "Optional",
          severity != "" ~ "Unsure",
          TRUE ~ "Unclassified"
        ),
        severity = factor(severity, levels = c("Necessary", "Optional", "Unsure", "Unclassified"))
      ) |>
      count(severity, .drop = FALSE)
  }
}
```

### Aggregate Statistics

```{r}
#| label: tbl-aggregate-stats
#| tbl-cap: "Aggregate statistics across all papers"
#| results: asis

if (!is.null(comparison_data) && n_papers > 0) {
  # Count human issues by severity
  total_human_issues <- if (!is.null(all_human_issues)) nrow(all_human_issues) else 0
  total_gpt_issues <- sum(issue_counts$num_gpt_issues)

  necessary_count <- if (!is.null(all_human_issues)) sum(grepl("necessary", tolower(all_human_issues$severity))) else 0
  optional_count <- if (!is.null(all_human_issues)) sum(grepl("optional", tolower(all_human_issues$severity))) else 0

  aggregate_stats <- data.frame(
    Metric = c(
      "Papers Compared",
      "Total Human Issues Parsed",
      "  - Necessary",
      "  - Optional",
      "  - Unsure/Other",
      "Total LLM Issues",
      "Avg Human Issues per Paper",
      "Avg LLM Issues per Paper"
    ),
    Value = c(
      n_papers,
      total_human_issues,
      necessary_count,
      optional_count,
      total_human_issues - necessary_count - optional_count,
      total_gpt_issues,
      round(total_human_issues / n_papers, 1),
      round(total_gpt_issues / n_papers, 1)
    )
  )

  kable(aggregate_stats, align = c("l", "r"))
}
```

```{r}
#| label: fig-severity-distribution
#| fig-cap: "Distribution of human issue severity labels"
#| fig-width: 8
#| fig-height: 4

if (!is.null(all_human_issues) && nrow(all_human_issues) > 0) {
  # Define severity colors
  sev_colors <- c("Necessary" = "#e74c3c", "Optional" = "#f39c12", "Unsure" = "#95a5a6", "Unclassified" = "#bdc3c7")

  ggplot(severity_dist, aes(x = severity, y = n, fill = severity)) +
    geom_col(alpha = 0.9) +
    geom_text(aes(label = n), vjust = -0.3, size = 5) +
    scale_fill_manual(values = sev_colors, drop = FALSE) +
    labs(
      x = "Severity Label",
      y = "Number of Issues"
    ) +
    theme_uj() +
    theme(legend.position = "none")
}
```

### Issue Count Distribution

```{r}
#| label: fig-issue-counts
#| fig-cap: "Number of key issues identified by GPT-5.2 Pro per paper"
#| fig-width: 10
#| fig-height: 5

if (!is.null(comparison_data) && n_papers > 0) {

  issue_counts |>
    ggplot(aes(x = reorder(paper_short, num_gpt_issues), y = num_gpt_issues)) +
    geom_col(fill = UJ_ORANGE, alpha = 0.8) +
    geom_hline(yintercept = mean(issue_counts$num_gpt_issues),
               linetype = "dashed", color = UJ_BLUE, linewidth = 1) +
    coord_flip() +
    labs(
      x = NULL,
      y = "Number of Key Issues",
      subtitle = paste0("Dashed line = mean (", round(mean(issue_counts$num_gpt_issues), 1), " issues)")
    ) +
    theme_uj()
}
```

### Coda Critique Length vs GPT Issue Count

```{r}
#| label: fig-length-scatter
#| fig-cap: "Relationship between human critique detail and GPT issue count"
#| fig-width: 8
#| fig-height: 6

if (!is.null(comparison_data) && n_papers > 0) {

  ggplot(issue_counts, aes(x = coda_critique_length, y = num_gpt_issues)) +
    geom_point(size = 3, color = UJ_ORANGE, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, color = UJ_BLUE, alpha = 0.2) +
    labs(
      x = "Human Critique Length (characters)",
      y = "GPT-5.2 Pro Issue Count"
    ) +
    theme_uj()
}
```

### Observable Structural Differences

The following are **verifiable structural differences** between the two data sources (not assessments of quality or coverage):

| Aspect | GPT-5.2 Pro | Human Expert (Coda) |
|------------------|---------------------|---------------------------------|
| Format | Numbered bullet points (array of strings) | Free-form prose |
| Structure | Ordered list (prompted: "most to least important") | Often uses severity labels ("Necessary", "Optional but important") |
| Source attribution | None (single model output) | Often cites specific evaluators (E1, E2, names) |
| Length | Constrained by prompt (\~10-12 issues) | Unconstrained, highly variable |

### Questions for Manual Review

The side-by-side comparisons above are provided for **manual expert assessment**. Key questions to investigate:

1.  **Coverage**: What proportion of human-identified issues appear in the GPT output (in some form, to some extent)? (This is the focal question).
2.  **Precision**: Are GPT issues substantive or does the model identify spurious/generic concerns?
3.  **Severity alignment**: Does GPT's importance ordering correlate with human severity labels?
4.  **Missed issues**: Are there critical human concerns that GPT systematically misses?
5.  **Novel issues**: Does GPT surface valid concerns that humans overlooked?

```{r}
#| label: assessment-status-callout
#| results: asis

if (has_llm_results) {
  cat('::: {.callout-tip}\n')
  cat('## LLM Assessment Complete\n\n')
  cat('The comparisons above have been assessed using GPT-5.2 Pro. See the [LLM-Based Assessment](#sec-llm-assessment) section for coverage and precision metrics.\n')
  cat(':::\n')
} else {
  cat('::: {.callout-warning}\n')
  cat('## No Automated Assessment Yet\n\n')
  cat('This page displays raw data for manual review. The claims about coverage, precision, and alignment **have not been assessed**—they require either:\n\n')
  cat('- Manual expert review of each paper\'s comparisons\n')
  cat('- LLM-based comparison (run the `key-issues-comparison` chunk in `methods.qmd`)\n\n')
  cat('Any conclusions about "high overlap" or "systematic differences" should emerge from that analysis, not be assumed.\n')
  cat(':::\n')
}
```

## Manual Annotation Tool

To systematically assess concordance between human and LLM critiques, use the **[Issue Annotation Tool](tools/issue_annotation_ui/index.html)**.

::: {.callout-note collapse="true"}
## For developers: Regenerating annotation data

To update the annotation data after changes to human critiques or LLM responses:

``` bash
# Generate annotation data (parses human critiques into individual issues)
python3 tools/build_issue_annotation_data.py
```
:::

**Annotation workflow:**

For each human-identified issue:

1.  **Match score (0-1)**: How well do LLM issues capture this concern?

    -   0 = Not addressed at all
    -   0.5 = Partially captured or tangentially related
    -   1 = Fully captured by one or more LLM issues

2.  **Confidence (0-1)**: How certain are you of this assessment?

3.  **Context flag**: Check this box if the human critique references information the LLM didn't have (e.g., appendix or preregistration materials not shared with the LLM at the point this evaluation was done)

4.  **Link to LLM issues**: Select which LLM issues (L1, L2, ...) correspond to this human issue

5.  **Discussion**: Explain your reasoning, note partial matches, or flag ambiguities

**Export** annotations as JSON/CSV for analysis. Annotations are auto-saved to browser localStorage.

::: callout-note
## Annotation Data Schema

The tool parses human critiques using heuristics:

-   **Severity labels** normalized to: `necessary`, `optional`, `unsure`
-   **Evaluator attributions** (E1, E2, DR) preserved in issue text
-   **Issue boundaries** detected via enumeration, sentence breaks, and section headers

Review parsed issues in the UI and edit/add/remove as needed before annotating.
:::
