---
title: "Results"
format:
  html: default
engine: knitr
---

```{r}
#| label: setup
#| code-summary: "Setup and libraries"
#| code-fold: true
#| message: false
#| warning: false

source("setup_params.R")

library("tidyverse")
library("janitor")
library("stringr")
library("here")
library("knitr")
library("kableExtra")
library("ggrepel")
library("scales")
library("jsonlite")
library("purrr")
library("tibble")
library("plotly")

# Theme and colors
UJ_ORANGE <- "#f19e4b"   # LLM primary
UJ_GREEN  <- "#99bb66"   # Human
UJ_BLUE   <- "#4e79a7"

MODEL_COLORS <- c(
  "GPT-5 Pro" = "#f19e4b",
  "GPT-5.2 Pro" = "#e57c23",
  "GPT-4o-mini" = "#74aa9c",
  "Claude Sonnet 4" = "#d4a574",
  "Gemini 2.0 Flash" = "#7986cb",
  "Human" = "#99bb66"
)

theme_uj <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot",
      legend.position = "bottom"
    )
}

canon_metric <- function(x) dplyr::recode(
  x,
  "advancing_knowledge" = "adv_knowledge",
  "open_science"        = "open_sci",
  "logic_communication" = "logic_comms",
  "global_relevance"    = "gp_relevance",
  "claims_evidence"     = "claims",
  .default = x
)

`%||%` <- function(x, y) if (!is.null(x)) x else y

# Pricing per 1M tokens (USD)
pricing <- tribble(
  ~model,              ~input_per_m, ~output_per_m,
  "GPT-5 Pro",         15.00,        60.00,
  "GPT-5.2 Pro",       15.00,        60.00,
  "GPT-4o-mini",       0.15,         0.60,
  "Claude Sonnet 4",   3.00,         15.00,
  "Gemini 2.0 Flash",  0.075,        0.30
)
```

```{r}
#| label: load-human-data
#| code-fold: true
#| code-summary: "Load human evaluation data"
#| message: false

# Load human ratings from Unjournal
UJmap <- read_delim("data/UJ_map.csv", delim = ";", show_col_types = FALSE) |>
  mutate(label_paper_title = research, label_paper = paper) |>
  select(c("label_paper_title", "label_paper"))

rsx <- read_csv("data/rsx_evalr_rating.csv", show_col_types = FALSE) |>
  clean_names() |>
  mutate(label_paper_title = research) |>
  select(-c("research"))

research <- read_csv("data/research.csv", show_col_types = FALSE) |>
  clean_names() |>
  filter(status == "50_published evaluations (on PubPub, by Unjournal)") |>
  left_join(UJmap, by = c("label_paper_title")) |>
  mutate(doi = str_trim(doi)) |>
  mutate(label_paper = case_when(
    doi == "https://doi.org/10.3386/w31162" ~ "Walker et al. 2023",
    doi == "doi.org/10.3386/w32728" ~ "Hahn et al. 2025",
    doi == "https://doi.org/10.3386/w30011" ~ "Bhat et al. 2022",
    doi == "10.1093/wbro/lkae010" ~ "Crawfurd et al. 2023",
    TRUE ~ label_paper
  )) |>
  left_join(rsx, by = c("label_paper_title"))

key_map <- research |>
  transmute(label_paper_title = str_trim(label_paper_title), label_paper = label_paper) |>
  filter(!is.na(label_paper_title)) |>
  distinct(label_paper_title, label_paper) |>
  group_by(label_paper_title) |>
  slice(1) |>
  ungroup()

rsx_research <- rsx |>
  mutate(label_paper_title = str_trim(label_paper_title)) |>
  left_join(key_map, by = "label_paper_title", relationship = "many-to-one")

metrics_human <- rsx_research |>
  mutate(criteria = canon_metric(criteria)) |>
  filter(criteria %in% c("overall", "claims", "methods", "adv_knowledge", "logic_comms", "open_sci", "gp_relevance")) |>
  transmute(
    paper = label_paper, criteria, evaluator, model = "Human",
    mid = as.numeric(middle_rating),
    lo = suppressWarnings(as.numeric(lower_ci)),
    hi = suppressWarnings(as.numeric(upper_ci))
  ) |>
  filter(!is.na(paper), !is.na(mid)) |>
  mutate(
    lo = ifelse(is.finite(lo), pmax(0, pmin(100, lo)), NA_real_),
    hi = ifelse(is.finite(hi), pmax(0, pmin(100, hi)), NA_real_)
  ) |>
  mutate(across(c(mid, lo, hi), ~ round(.x, 4))) |>
  distinct(paper, criteria, model, evaluator, mid, lo, hi)

human_avg <- metrics_human |>
  filter(criteria == "overall") |>
  group_by(paper) |>
  summarise(
    human_mid = mean(mid, na.rm = TRUE),
    human_lo = mean(lo, na.rm = TRUE),
    human_hi = mean(hi, na.rm = TRUE),
    n_human = n(),
    .groups = "drop"
  )

# Summary stats for inline use
n_human_papers <- n_distinct(metrics_human$paper)
n_human_evaluators <- n_distinct(metrics_human$evaluator)
```

```{r}
#| label: load-llm-data
#| code-fold: true
#| code-summary: "Load LLM evaluation data (all models)"
#| message: false

model_dirs <- list(
  "gpt5_pro_updated_jan2026" = "GPT-5 Pro",
  "gpt52_pro_focal_jan2026" = "GPT-5.2 Pro",
  "gpt_4o_mini_2024_07_18" = "GPT-4o-mini",
  "claude_sonnet_4_20250514" = "Claude Sonnet 4",
  "gemini_2.0_flash" = "Gemini 2.0 Flash"
)

parse_response <- function(path, model_name) {
  tryCatch({
    r <- jsonlite::fromJSON(path, simplifyVector = FALSE)
    paper <- basename(path) |>
      str_replace("\\.response\\.json$", "") |>
      str_replace_all("_", " ")

    parsed <- NULL
    if (!is.null(r$output_text) && nchar(r$output_text) > 0) {
      parsed <- jsonlite::fromJSON(r$output_text, simplifyVector = TRUE)
    } else if (!is.null(r$parsed)) {
      parsed <- r$parsed
    } else if (!is.null(r$output)) {
      msg <- purrr::detect(r$output, ~ .x$type == "message", .default = NULL)
      if (!is.null(msg) && length(msg$content) > 0) {
        parsed <- jsonlite::fromJSON(msg$content[[1]]$text, simplifyVector = TRUE)
      }
    }

    if (is.null(parsed)) return(NULL)

    metrics <- parsed$metrics
    metric_rows <- list()
    tier_rows <- list()

    for (nm in names(metrics)) {
      if (nm %in% c("tier_should", "tier_will")) {
        tier_rows[[length(tier_rows) + 1]] <- tibble(
          paper = paper, model = model_name, tier_kind = nm,
          score = metrics[[nm]]$score,
          ci_lower = metrics[[nm]]$ci_lower,
          ci_upper = metrics[[nm]]$ci_upper
        )
      } else {
        metric_rows[[length(metric_rows) + 1]] <- tibble(
          paper = paper, model = model_name, metric = nm,
          midpoint = metrics[[nm]]$midpoint,
          lower_bound = metrics[[nm]]$lower_bound,
          upper_bound = metrics[[nm]]$upper_bound
        )
      }
    }

    input_tok <- r$usage$input_tokens %||% r$input_tokens
    output_tok <- r$usage$output_tokens %||% r$output_tokens
    reasoning_tok <- r$usage$output_tokens_details$reasoning_tokens

    list(
      metrics = bind_rows(metric_rows),
      tiers = bind_rows(tier_rows),
      tokens = tibble(
        paper = paper, model = model_name,
        input_tokens = input_tok %||% NA_integer_,
        output_tokens = output_tok %||% NA_integer_,
        reasoning_tokens = reasoning_tok %||% NA_integer_
      ),
      summary = tibble(
        paper = paper, model = model_name,
        assessment_summary = parsed$assessment_summary
      )
    )
  }, error = function(e) NULL)
}

load_all_llm <- function() {
  all_metrics <- list()
  all_tiers <- list()
  all_tokens <- list()
  all_summaries <- list()

  for (dir_name in names(model_dirs)) {
    model_name <- model_dirs[[dir_name]]
    json_dir <- here("results", dir_name, "json")

    if (dir.exists(json_dir)) {
      files <- list.files(json_dir, pattern = "\\.response\\.json$", full.names = TRUE)

      for (f in files) {
        result <- parse_response(f, model_name)
        if (!is.null(result)) {
          all_metrics[[length(all_metrics) + 1]] <- result$metrics
          all_tiers[[length(all_tiers) + 1]] <- result$tiers
          all_tokens[[length(all_tokens) + 1]] <- result$tokens
          all_summaries[[length(all_summaries) + 1]] <- result$summary
        }
      }
    }
  }

  list(
    metrics = bind_rows(all_metrics) |> mutate(criteria = canon_metric(metric)),
    tiers = bind_rows(all_tiers),
    tokens = bind_rows(all_tokens),
    summaries = bind_rows(all_summaries)
  )
}

llm_data <- load_all_llm()
llm_metrics <- llm_data$metrics
llm_tiers <- llm_data$tiers
llm_tokens <- llm_data$tokens
llm_summaries <- llm_data$summaries

# Summary stats for inline use
n_llm_models <- n_distinct(llm_metrics$model)
n_llm_papers <- n_distinct(llm_metrics$paper)
llm_model_names <- unique(llm_metrics$model)
llm_model_list <- paste(llm_model_names, collapse = ", ")

# Papers with both human and LLM data
matched_papers <- intersect(
 unique(metrics_human$paper),
 unique(llm_metrics$paper)
)
n_matched <- length(matched_papers)

# Cost summary
cost_data <- llm_tokens |>
  left_join(pricing, by = "model") |>
  mutate(
    reasoning_tokens = coalesce(reasoning_tokens, 0L),
    cost_usd = (input_tokens * input_per_m + output_tokens * output_per_m) / 1e6
  )

total_cost <- sum(cost_data$cost_usd, na.rm = TRUE)
total_evaluations <- nrow(llm_tokens)

# Per-model summaries
model_summary <- cost_data |>
  group_by(model) |>
  summarise(
    n_papers = n(),
    avg_input = round(mean(input_tokens, na.rm = TRUE)),
    avg_output = round(mean(output_tokens, na.rm = TRUE)),
    avg_reasoning = round(mean(reasoning_tokens[reasoning_tokens > 0], na.rm = TRUE)),
    total_cost = sum(cost_usd, na.rm = TRUE),
    cost_per_paper = mean(cost_usd, na.rm = TRUE),
    .groups = "drop"
  )
```

We compare LLM-generated research evaluations against human expert reviews from [The Unjournal](https://unjournal.org). This analysis covers `r n_llm_papers` papers evaluated by `r n_llm_models` frontier LLMs (`r llm_model_list`), with `r n_matched` papers having both human and LLM evaluations for direct comparison. Human evaluations come from `r n_human_evaluators` expert reviewers across `r n_human_papers` papers in The Unjournal's database.

## Evaluation Overview

We submit each PDF through a single, schema-enforced Responses call. The system prompt mirrors the Unjournal rubric, blocks extrinsic priors (authors, venue, web context), and requires strict JSON: a ~1k-word diagnostic summary plus numeric midpoints and 90% credible intervals for every metric. We enforce well-formed intervals (`lower < midpoint < upper` for 0–100 metrics; `ci_lower < score < ci_upper` for 0–5 tiers). GPT-5 Pro runs with reasoning effort = "high", so traces are available for audit (see [Methods](methods.qmd)).

Metrics (0–100 unless noted): overall (global quality/impact), claims_evidence (claim clarity/support), methods (design/identification/robustness), advancing_knowledge (contribution to field/practice), logic_communication (internal consistency and exposition), open_science (reproducibility, data/code availability), global_relevance (decision and priority usefulness). Journal tiers (0–5, continuous) capture where the work should publish vs. will publish.

Each numeric field carries a midpoint (model’s 50% belief) and a 90% credible interval; wide intervals indicate acknowledged uncertainty.

::: {.callout-note}
## Future Models

We plan to extend this comparison to newer frontier models as they become available, including GPT-5.2 Pro, Claude Opus 4.5, Gemini 3, and Grok 3. 
:::

### Token Usage and Cost

Token costs are computed from API-reported input/output tokens (GPT-5 Pro runs with `reasoning` = high). Use the boxplot log scale to compare efficiency frontiers; reasoning tokens are reported when available.

```{r}
#| label: tbl-cost-overview
#| tbl-cap: "Token usage and estimated cost per model"

if (nrow(llm_tokens) > 0) {

  cost_table <- model_summary |>
    transmute(
      Model = model,
      Papers = n_papers,
      `Avg input` = scales::comma(avg_input),
      `Avg output` = scales::comma(avg_output),
      `Avg reasoning` = ifelse(is.finite(avg_reasoning) & avg_reasoning > 0,
                                scales::comma(avg_reasoning), "—"),
      `Total cost` = scales::dollar(total_cost, accuracy = 0.01),
      `Cost/paper` = scales::dollar(cost_per_paper, accuracy = 0.001)
    ) |>
    arrange(desc(as.numeric(gsub("[$,]", "", `Total cost`))))

  knitr::kable(cost_table, align = c("l", rep("r", 6)))
}
```

Total cost across all `r total_evaluations` LLM evaluations: **`r scales::dollar(total_cost, accuracy = 0.01)`**.

```{r}
#| label: fig-cost-box
#| fig-cap: "Cost per paper by model (log scale)"
#| fig-width: 9
#| fig-height: 5

if (nrow(cost_data) > 0) {

  ggplot(cost_data, aes(x = reorder(model, cost_usd, FUN = median), y = cost_usd, fill = model)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.5, size = 2) +
    scale_fill_manual(values = MODEL_COLORS, guide = "none") +
    scale_y_log10(labels = scales::dollar_format(accuracy = 0.001)) +
    coord_flip() +
    labs(x = NULL, y = "Cost per paper (USD, log scale)") +
    theme_uj()
}
```

---

## Interactive Forest Plot: Human vs LLM Ratings

The forest plot below shows **overall percentile ratings** for each of the `r n_matched` papers with matched human and LLM evaluations. Human evaluator ratings are shown in green (individual evaluators as small dots, mean as diamond). LLM ratings from different models are shown with distinct colors.

Interpretation: points are midpoints, horizontal bars are the model’s 90% credible interval.

Hover over any point to see details. Click legend items to show/hide models.

```{r}
#| label: fig-forest-plotly
#| fig-cap: "Interactive forest plot: Human and LLM overall ratings by paper"
#| fig-width: 12
#| fig-height: 10

forest_human <- metrics_human |>
  filter(criteria == "overall") |>
  transmute(paper, model = "Human", evaluator, mid, lo, hi, is_mean = FALSE)

forest_human_means <- human_avg |>
  transmute(paper, model = "Human", evaluator = "Mean",
            mid = human_mid, lo = human_lo, hi = human_hi, is_mean = TRUE)

forest_llm <- llm_metrics |>
  filter(criteria == "overall") |>
  transmute(paper, model, evaluator = model,
            mid = midpoint, lo = lower_bound, hi = upper_bound, is_mean = FALSE)

forest_data <- bind_rows(forest_human, forest_human_means, forest_llm) |>
  filter(paper %in% matched_papers)

if (nrow(forest_data) > 0) {

  paper_order <- forest_data |>
    filter(model == "Human", is_mean) |>
    arrange(desc(mid)) |>
    pull(paper)

  forest_data <- forest_data |>
    mutate(
      paper = factor(paper, levels = rev(paper_order)),
      paper_num = as.numeric(paper),
      hover_text = paste0(
        "<b>", paper, "</b><br>",
        model, ifelse(evaluator != model & !is_mean, paste0(" (", evaluator, ")"), ""), "<br>",
        "Rating: ", round(mid, 1),
        ifelse(!is.na(lo) & !is.na(hi), paste0(" [", round(lo, 1), "–", round(hi, 1), "]"), "")
      )
    )

  p <- plot_ly() |>
    add_segments(x = 25, xend = 25, y = 0.5, yend = length(paper_order) + 0.5,
                 line = list(color = "lightgray", dash = "dot"),
                 showlegend = FALSE, hoverinfo = "none") |>
    add_segments(x = 50, xend = 50, y = 0.5, yend = length(paper_order) + 0.5,
                 line = list(color = "gray", dash = "dot"),
                 showlegend = FALSE, hoverinfo = "none") |>
    add_segments(x = 75, xend = 75, y = 0.5, yend = length(paper_order) + 0.5,
                 line = list(color = "lightgray", dash = "dot"),
                 showlegend = FALSE, hoverinfo = "none")

  models_to_plot <- c("Human", setdiff(unique(forest_data$model), "Human"))

  for (m in models_to_plot) {
    model_data <- forest_data |> filter(model == m)

    if (m == "Human") {
      indiv <- model_data |> filter(!is_mean)
      if (nrow(indiv) > 0) {
        p <- p |> add_trace(
          data = indiv, type = "scatter", mode = "markers",
          x = ~mid, y = ~paper_num,
          marker = list(color = MODEL_COLORS[m], size = 6, opacity = 0.4),
          text = ~hover_text, hoverinfo = "text",
          name = "Human (individual)", legendgroup = "Human"
        )
      }
      means <- model_data |> filter(is_mean)
      if (nrow(means) > 0) {
        p <- p |> add_trace(
          data = means, type = "scatter", mode = "markers",
          x = ~mid, y = ~paper_num,
          error_x = list(type = "data", symmetric = FALSE,
                         arrayminus = ~(mid - lo), array = ~(hi - mid),
                         color = MODEL_COLORS[m], thickness = 1.5),
          marker = list(color = MODEL_COLORS[m], size = 12, symbol = "diamond"),
          text = ~hover_text, hoverinfo = "text",
          name = "Human (mean)", legendgroup = "Human"
        )
      }
    } else {
      p <- p |> add_trace(
        data = model_data, type = "scatter", mode = "markers",
        x = ~mid, y = ~paper_num,
        error_x = list(type = "data", symmetric = FALSE,
                       arrayminus = ~(mid - lo), array = ~(hi - mid),
                       color = MODEL_COLORS[m], thickness = 1.5),
        marker = list(color = MODEL_COLORS[m], size = 10),
        text = ~hover_text, hoverinfo = "text", name = m
      )
    }
  }

  p <- p |> layout(
    xaxis = list(title = "Overall percentile (0–100)", range = c(0, 105), zeroline = FALSE),
    yaxis = list(title = "", ticktext = paper_order, tickvals = seq_along(paper_order), tickmode = "array"),
    legend = list(orientation = "h", y = -0.1, x = 0.5, xanchor = "center"),
    hovermode = "closest",
    margin = list(l = 200)
  )

  p

} else {
  cat("No matched human-LLM data available for forest plot.\n")
}
```

## Human vs LLM Scatter Plot

Each point represents a paper, with human mean rating on the x-axis and LLM rating on the y-axis. Points on the diagonal indicate perfect agreement.

Note: human means carry their own variance; correlations here are bounded by human inter-rater noise.

```{r}
#| label: fig-scatter-interactive
#| fig-cap: "Interactive scatter: Human vs LLM overall ratings"
#| fig-width: 10
#| fig-height: 8

scatter_data <- llm_metrics |>
  filter(criteria == "overall") |>
  inner_join(human_avg, by = "paper") |>
  mutate(diff = midpoint - human_mid, paper_short = str_trunc(paper, 25))

if (nrow(scatter_data) > 0) {

  p <- ggplot(scatter_data, aes(
    x = human_mid, y = midpoint, color = model,
    text = paste0(
      "<b>", paper, "</b><br>",
      "Human: ", round(human_mid, 1), " [", round(human_lo, 1), "–", round(human_hi, 1), "]<br>",
      "LLM: ", round(midpoint, 1), " [", round(lower_bound, 1), "–", round(upper_bound, 1), "]<br>",
      "Diff: ", sprintf("%+.1f", diff)
    )
  )) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
    geom_point(size = 3.5, alpha = 0.7) +
    scale_color_manual(values = MODEL_COLORS) +
    coord_cartesian(xlim = c(0, 100), ylim = c(0, 100)) +
    labs(x = "Human mean rating (0–100)", y = "LLM rating (0–100)", color = "Model") +
    theme_uj()

  ggplotly(p, tooltip = "text") |>
    layout(
      legend = list(orientation = "h", y = -0.12, x = 0.5, xanchor = "center"),
      hoverlabel = list(bgcolor = "white")
    )
}
```

## Model Agreement Statistics

We assess agreement between LLM and human evaluators using multiple metrics:

- **Pearson r**: Linear correlation between ratings
- **Spearman ρ**: Rank-based correlation (robust to outliers)
- **Mean bias**: Average difference (LLM − Human); positive = LLM rates higher
- **RMSE/MAE**: Error magnitude measures

```{r}
#| label: tbl-agreement
#| tbl-cap: "Agreement between each LLM and human mean ratings (overall criterion)"

if (nrow(scatter_data) > 0) {

  agreement_tbl <- scatter_data |>
    group_by(Model = model) |>
    summarise(
      N = n(),
      `Pearson r` = round(cor(human_mid, midpoint, use = "complete.obs"), 3),
      `Spearman ρ` = round(cor(human_mid, midpoint, method = "spearman", use = "complete.obs"), 3),
      `Mean bias` = sprintf("%+.1f", mean(midpoint - human_mid, na.rm = TRUE)),
      RMSE = round(sqrt(mean((midpoint - human_mid)^2, na.rm = TRUE)), 1),
      MAE = round(mean(abs(midpoint - human_mid), na.rm = TRUE), 1),
      .groups = "drop"
    ) |>
    arrange(desc(`Pearson r`))

  knitr::kable(agreement_tbl, align = c("l", rep("r", 6)))
}
```

### Inter-Rater Agreement (Krippendorff's Alpha)

Krippendorff's alpha measures agreement among multiple raters, accounting for chance agreement. Values above 0.8 indicate good agreement; 0.67–0.8 is acceptable for exploratory research.
Ceiling effects apply because human evaluators are not perfectly aligned.

```{r}
#| label: tbl-krippendorff
#| tbl-cap: "Inter-rater agreement: Krippendorff's alpha by criterion"

# Install irr if needed: install.packages("irr")
if (requireNamespace("irr", quietly = TRUE)) {

  library(irr)

  # Prepare data: each row = paper, each column = rater (human mean + each LLM)
  compute_alpha <- function(criterion) {

    # Human mean per paper
    human_crit <- metrics_human |>
      filter(criteria == criterion, paper %in% matched_papers) |>
      group_by(paper) |>
      summarise(Human = mean(mid, na.rm = TRUE), .groups = "drop")

    # LLM ratings per paper
    llm_crit <- llm_metrics |>
      filter(criteria == criterion, paper %in% matched_papers) |>
      select(paper, model, midpoint) |>
      pivot_wider(names_from = model, values_from = midpoint)

    # Combine
    combined <- human_crit |>
      inner_join(llm_crit, by = "paper") |>
      select(-paper)

    if (nrow(combined) < 3 || ncol(combined) < 2) return(NA_real_)

    # Krippendorff's alpha (interval scale)
    tryCatch({
      ka <- kripp.alpha(t(as.matrix(combined)), method = "interval")
      ka$value
    }, error = function(e) NA_real_)
  }

  criteria_list <- c("overall", "claims", "methods", "adv_knowledge",
                     "logic_comms", "open_sci", "gp_relevance")

  alpha_results <- tibble(
    Criterion = c("Overall", "Claims", "Methods", "Adv. Knowledge",
                  "Logic & Comms", "Open Science", "Global Relevance"),
    criterion_key = criteria_list
  ) |>
    rowwise() |>
    mutate(
      `Krippendorff's α` = compute_alpha(criterion_key),
      `α (formatted)` = ifelse(is.na(`Krippendorff's α`), "—",
                                sprintf("%.3f", `Krippendorff's α`)),
      Interpretation = case_when(
        is.na(`Krippendorff's α`) ~ "Insufficient data",
        `Krippendorff's α` >= 0.8 ~ "Good",
        `Krippendorff's α` >= 0.67 ~ "Acceptable",
        `Krippendorff's α` >= 0.4 ~ "Fair",
        TRUE ~ "Poor"
      )
    ) |>
    ungroup() |>
    select(Criterion, `α (formatted)`, Interpretation) |>
    rename(`Krippendorff's α` = `α (formatted)`)

  knitr::kable(alpha_results, align = c("l", "r", "l"))

} else {
  cat("*Install the `irr` package for Krippendorff's alpha: `install.packages('irr')`*\n")
}
```

### Agreement by Evaluation Criterion

```{r}
#| label: tbl-agreement-by-criterion
#| tbl-cap: "Pearson and Spearman correlations by evaluation criterion"

# Compute correlations for each criterion and model
criterion_agreement <- llm_metrics |>
  filter(paper %in% matched_papers) |>
  inner_join(
    metrics_human |>
      filter(paper %in% matched_papers) |>
      group_by(paper, criteria) |>
      summarise(human_mid = mean(mid, na.rm = TRUE), .groups = "drop"),
    by = c("paper", "criteria")
  ) |>
  group_by(Model = model, Criterion = criteria) |>
  summarise(
    n = n(),
    r = cor(midpoint, human_mid, use = "complete.obs"),
    rho = cor(midpoint, human_mid, method = "spearman", use = "complete.obs"),
    .groups = "drop"
  ) |>
  filter(n >= 3)  # need at least 3 papers for meaningful correlation

if (nrow(criterion_agreement) > 0) {

  # Pivot for display
  criterion_labels <- c(
    overall = "Overall", claims = "Claims", methods = "Methods",
    adv_knowledge = "Adv. Knowledge", logic_comms = "Logic & Comms",
    open_sci = "Open Science", gp_relevance = "Global Relevance"
  )

  corr_table <- criterion_agreement |>
    mutate(
      Criterion = criterion_labels[Criterion],
      `r / ρ` = sprintf("%.2f / %.2f", r, rho)
    ) |>
    select(Model, Criterion, `r / ρ`) |>
    pivot_wider(names_from = Model, values_from = `r / ρ`, values_fill = "—")

  # Order criteria
  corr_table <- corr_table |>
    mutate(Criterion = factor(Criterion, levels = criterion_labels)) |>
    arrange(Criterion)

  knitr::kable(corr_table, align = c("l", rep("r", ncol(corr_table) - 1)))
}
```

## Rating Distribution by Model

```{r}
#| label: fig-rating-distribution
#| fig-cap: "Distribution of overall ratings by evaluator type"
#| fig-width: 10
#| fig-height: 6

all_overall <- bind_rows(
  metrics_human |> filter(criteria == "overall") |> transmute(paper, model, rating = mid),
  llm_metrics |> filter(criteria == "overall") |> transmute(paper, model, rating = midpoint)
) |>
  filter(paper %in% matched_papers) |>
  filter(model != "Claude Sonnet 4")

if (nrow(all_overall) > 0) {
  ggplot(all_overall, aes(x = rating, fill = model)) +
    geom_density(alpha = 0.5, adjust = 1.2) +
    scale_fill_manual(values = MODEL_COLORS) +
    labs(x = "Overall percentile (0–100)", y = "Density", fill = "Evaluator") +
    theme_uj() +
    theme(legend.position = "right")
}
```

## Criteria-Level Comparison

```{r}
#| label: fig-criteria-heatmap
#| fig-cap: "Mean rating by model and criterion"
#| fig-width: 11
#| fig-height: 6

metric_labels <- c(
  overall = "Overall", claims = "Claims", methods = "Methods",
  adv_knowledge = "Adv. Knowledge", logic_comms = "Logic & Comms",
  open_sci = "Open Science", gp_relevance = "Global Relevance"
)

criteria_means <- bind_rows(
  metrics_human |>
    filter(criteria %in% names(metric_labels), paper %in% matched_papers) |>
    group_by(criteria) |>
    summarise(mean_rating = mean(mid, na.rm = TRUE), .groups = "drop") |>
    mutate(model = "Human"),
  llm_metrics |>
    filter(criteria %in% names(metric_labels), paper %in% matched_papers) |>
    group_by(model, criteria) |>
    summarise(mean_rating = mean(midpoint, na.rm = TRUE), .groups = "drop")
) |>
  mutate(criteria_label = factor(metric_labels[criteria], levels = metric_labels))

if (nrow(criteria_means) > 0) {
  ggplot(criteria_means, aes(x = criteria_label, y = model, fill = mean_rating)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(aes(label = round(mean_rating, 0)), size = 4, color = "white", fontface = "bold") +
    scale_fill_gradient(low = "#f0f0f0", high = UJ_ORANGE, limits = c(40, 80), name = "Mean") +
    labs(x = NULL, y = NULL) +
    theme_uj() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 10), panel.grid = element_blank())
}
```

---

**Future extensions:** add calibration diagnostics (share of human means within model 90% CIs by criterion), interval-width profiles by model/criterion, and cost–agreement frontiers (correlation vs. cost per paper).

*Note: GPT-5 Pro evaluations include extended reasoning traces. See the [Appendix](appendix_llm_traces.qmd) for full assessment summaries and reasoning traces per paper.*

## Prompt Comparison: Legacy vs Updated GPT-5 Pro

We compare GPT-5 Pro evaluations between two prompt versions:

- **Legacy (Oct 2024):** Original combined schema without assessment summary
- **Updated (Jan 2026):** Modular prompt with explicit diagnostic assessment summary

```{r}
#| label: load-legacy-comparison
#| code-fold: true
#| code-summary: "Load legacy vs updated GPT-5 Pro data"
#| message: false

# Load legacy GPT-5 data
legacy_metrics <- read_csv("data/metrics_long_gpt-5.csv", show_col_types = FALSE) |>
  transmute(
    paper_raw = paper,
    paper = str_replace_all(paper, " ", "_") |> str_replace_all("\\.", ""),  # normalize
    metric = metric,
    legacy_mid = midpoint,
    legacy_lo = lower_bound,
    legacy_hi = upper_bound
  )

# Load updated GPT-5 Pro data
updated_metrics <- read_csv("data/metrics_long_gpt5_pro_jan2026.csv", show_col_types = FALSE) |>
  transmute(
    paper = str_replace_all(paper, "\\.", ""),  # remove dots for matching
    metric = metric,
    updated_mid = midpoint,
    updated_lo = lower_bound,
    updated_hi = upper_bound
  )

# Join by normalized paper name and metric
prompt_comparison <- legacy_metrics |>
  inner_join(updated_metrics, by = c("paper", "metric")) |>
  mutate(
    diff = updated_mid - legacy_mid,
    metric_label = case_when(
      metric == "overall" ~ "Overall",
      metric == "claims_evidence" ~ "Claims",
      metric == "methods" ~ "Methods",
      metric == "advancing_knowledge" ~ "Adv. Knowledge",
      metric == "logic_communication" ~ "Logic & Comms",
      metric == "open_science" ~ "Open Science",
      metric == "global_relevance" ~ "Global Relevance",
      TRUE ~ metric
    )
  )

n_matched_papers <- n_distinct(prompt_comparison$paper)
```

We matched `r n_matched_papers` papers across both runs.

```{r}
#| label: fig-prompt-scatter
#| fig-cap: "Legacy vs Updated GPT-5 Pro ratings (overall metric)"
#| fig-width: 8
#| fig-height: 7

overall_comp <- prompt_comparison |> filter(metric == "overall")

if (nrow(overall_comp) > 0) {
  cor_val <- cor(overall_comp$legacy_mid, overall_comp$updated_mid, use = "complete.obs")

  ggplot(overall_comp, aes(x = legacy_mid, y = updated_mid)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
    geom_point(alpha = 0.7, size = 2.5, color = UJ_ORANGE) +
    geom_smooth(method = "lm", se = TRUE, alpha = 0.2, color = UJ_BLUE) +
    annotate("text", x = 55, y = 95,
             label = paste0("r = ", round(cor_val, 2)),
             size = 5, fontface = "bold") +
    labs(
      x = "Legacy Prompt (Oct 2024)",
      y = "Updated Prompt (Jan 2026)",
      title = "GPT-5 Pro: Overall Rating Comparison",
      subtitle = "Dashed line = perfect agreement"
    ) +
    coord_fixed(xlim = c(40, 100), ylim = c(40, 100)) +
    theme_uj()
}
```

```{r}
#| label: fig-prompt-diff-by-metric
#| fig-cap: "Rating difference (Updated - Legacy) by criterion"
#| fig-width: 10
#| fig-height: 5

if (nrow(prompt_comparison) > 0) {
  metric_order <- prompt_comparison |>
    group_by(metric_label) |>
    summarise(mean_diff = mean(diff, na.rm = TRUE)) |>
    arrange(mean_diff) |>
    pull(metric_label)

  prompt_comparison |>
    mutate(metric_label = factor(metric_label, levels = metric_order)) |>
    ggplot(aes(x = metric_label, y = diff)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_boxplot(fill = UJ_ORANGE, alpha = 0.6, outlier.alpha = 0.3) +
    labs(
      x = NULL,
      y = "Rating Difference (Updated - Legacy)",
      title = "Prompt Version Effect by Criterion",
      subtitle = "Positive = updated prompt rates higher"
    ) +
    theme_uj() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}
```

```{r}
#| label: tbl-prompt-summary
#| tbl-cap: "Summary statistics: Legacy vs Updated prompt"

if (nrow(prompt_comparison) > 0) {
  prompt_comparison |>
    group_by(Criterion = metric_label) |>
    summarise(
      `Legacy Mean` = round(mean(legacy_mid, na.rm = TRUE), 1),
      `Updated Mean` = round(mean(updated_mid, na.rm = TRUE), 1),
      `Mean Diff` = round(mean(diff, na.rm = TRUE), 1),
      `SD Diff` = round(sd(diff, na.rm = TRUE), 1),
      .groups = "drop"
    ) |>
    kable() |>
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
}
```

The updated prompt with explicit diagnostic assessment appears to produce systematically different ratings. The correlation between prompt versions indicates the degree of consistency across the prompt change.
