---
title: "Results: Critiques & Key Issues"
format:
  html: default
engine: knitr
---

```{r}
#| label: setup-critiques
#| code-summary: "Setup and libraries"
#| code-fold: true
#| message: false
#| warning: false

library("tidyverse")
library("jsonlite")
library("knitr")
library("kableExtra")
library("DT")
library("reticulate")

# Theme colors (matching results_ratings.qmd)
UJ_ORANGE <- "#f19e4b"
UJ_GREEN  <- "#99bb66"
UJ_BLUE   <- "#4e79a7"

# Severity colors for badges
SEV_NECESSARY <- "#e74c3c"   # Red
SEV_OPTIONAL  <- "#f39c12"   # Orange
SEV_UNSURE    <- "#95a5a6"   # Gray

theme_uj <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot",
      legend.position = "bottom"
    )
}

# Function to create severity badge HTML
severity_badge <- function(severity) {
  sev <- tolower(trimws(severity))
  if (grepl("necessary", sev)) {
    return('<span style="background-color:#e74c3c;color:white;padding:2px 6px;border-radius:3px;font-size:0.8em;">🔴 Necessary</span>')
  } else if (grepl("optional", sev)) {
    return('<span style="background-color:#f39c12;color:white;padding:2px 6px;border-radius:3px;font-size:0.8em;">🟠 Optional</span>')
  } else if (sev != "") {
    return('<span style="background-color:#95a5a6;color:white;padding:2px 6px;border-radius:3px;font-size:0.8em;">⚪ Unsure</span>')
  }
  return('<span style="background-color:#bdc3c7;color:white;padding:2px 6px;border-radius:3px;font-size:0.8em;">—</span>')
}

# Function to truncate text for table display
truncate_text <- function(text, max_chars = 80) {
  if (nchar(text) > max_chars) {
    paste0(substr(text, 1, max_chars), "...")
  } else {
    text
  }
}
```

This chapter compares **qualitative critiques**: the key methodological and interpretive issues identified by GPT-5.2 Pro against human expert critiques documented by Unjournal evaluation managers in Coda.

We assess alignment using two approaches:[^approaches-note]

1. **LLM-based comparison**: GPT-5.2 Pro evaluates coverage and precision (see [LLM-Based Assessment](#sec-llm-assessment))
2. **Manual annotation**: Human labeling using the Issue Annotation Tool (see [Manual Annotation Tool](#manual-annotation-tool))

[^approaches-note]: Jan 8 2026: We are actively working on both approaches. Results shown here are preliminary.

## Data Sources

**GPT-5.2 Pro Key Issues**: Structured output from the focal evaluation run (January 2026), where the model was prompted to identify 5-15 key issues per paper, ordered from most to least important.

**Human Expert Critiques**: Curated content from the "Key critiques & issues with paper" column in The Unjournal's internal tracking database (Coda), written by evaluation managers synthesizing evaluator feedback. These use severity labels ("Necessary", "Optional but important", "Unsure") and cite specific evaluator comments.

```{r}
#| label: load-comparison-data
#| code-fold: true
#| code-summary: "Load comparison data"

# Load the matched comparison data
comparison_file <- "results/key_issues_comparison.json"
comparison_results_file <- c(
  "results/key_issue_comp_results.json",
  "results/key_issues_comparison_results.json"
)
comparison_results_file <- comparison_results_file[file.exists(comparison_results_file)][1]

if (file.exists(comparison_file)) {
  comparison_data <- fromJSON(comparison_file)
  n_papers <- nrow(comparison_data)
} else {
  comparison_data <- NULL
  n_papers <- 0
}

# Load LLM-based comparison results (coverage/precision) if available
if (!is.null(comparison_data) && !is.na(comparison_results_file)) {
  llm_results_raw <- fromJSON(comparison_results_file)
  llm_results <- llm_results_raw |>
    as_tibble() |>
    unnest_wider(comparison) |>
    select(
      gpt_paper,
      coverage_pct,
      precision_pct,
      missed_issues,
      extra_issues,
      overall_rating,
      overall_justification,
      detailed_notes
    )

  comparison_data <- comparison_data |>
    left_join(llm_results, by = "gpt_paper")
}
```

```{python}
#| label: parse-human-issues
#| code-fold: true
#| code-summary: "Parse human critiques into individual issues"

import json
import re
from pathlib import Path

# Severity parsing regex patterns
SEVERITY_HEADER_RE = re.compile(
    r"^(necessary|optional|optional but important|unsure|less important|possibly relevant|"
    r"probably important|probably correct|probably|note|notes|optional/desirable|"
    r"probably useful|maybe|possibly|unsure if correct|limitations)(\s|:|$)",
    re.IGNORECASE,
)
ENUM_RE = re.compile(r"^\s*(\d+)[\.\)]\s+")
EVALUATOR_RE = re.compile(r"^\[?(E\d+|Evaluator|DR|NotebookLM)\]?[\s:\[]", re.IGNORECASE)

def normalize_severity(header):
    h = header.lower()
    if "necessary" in h:
        return "necessary"
    if "optional" in h:
        return "optional"
    if any(word in h for word in ["unsure", "possibly", "probably", "less important", "maybe", "limitations"]):
        return "unsure"
    return ""

def extract_short_label(text, max_words=12):
    """Extract a descriptive short label from issue text."""
    original = text.strip()

    # Meta-commentary patterns to skip
    META_SKIP_PATTERNS = [
        r'^I chose',
        r'^I selected',
        r'^I think',
        r'^I would',
        r'^I believe',
        r'^These are',
        r'^This is (a|the|my)',
        r'^Note[s]?\s*[:)]',
        r'^DR\s*[:)]',
        r'^NotebookLM\s*[:)]',
        r'^The following',
        r'^Summary\s*[:)]',
        r'^Quote from',
        r'^See also',
        r'^As noted',
        r'^and that',  # Fragments starting with conjunctions
        r'^and it',
        r"^and that it's",
        r'^or that',
        r'^but that',
        r'^but it',
        r'^which ',
        r'^that ',
        r'^to be ',
        r'^it is ',
        r'^it would ',
    ]

    # First clean: Remove leading evaluator tags (DR —, [E1], etc.)
    cleaned_for_meta_check = re.sub(
        r'^[\s\[]*(?:E\d+|Evaluator\s*\d*|DR|NotebookLM|Treich|Triech|Capitan|Reiley|Manheim|Espinosa|Seinkmane|Cloud|Tsai)[\]:\s—\-]+',
        '', original, flags=re.IGNORECASE
    ).strip()

    # Check for meta-commentary patterns AFTER removing evaluator prefix
    for pattern in META_SKIP_PATTERNS:
        if re.match(pattern, cleaned_for_meta_check, re.IGNORECASE):
            # Try to find actual content after the meta-commentary
            # Look for a quoted phrase or substantive content after colon/period
            quote_in_meta = re.search(r'["""\']([^"""\'\n]{15,80})["""\']', cleaned_for_meta_check)
            if quote_in_meta:
                quoted = quote_in_meta.group(1).strip()
                if not re.match(r'^(the |a |an |this )', quoted.lower()):
                    return quoted
            # Look for content after a colon
            after_meta = re.split(r'[:\n]', cleaned_for_meta_check, maxsplit=1)
            if len(after_meta) > 1 and len(after_meta[1].strip()) > 20:
                original = after_meta[1].strip()
                break
            else:
                # Return a generic placeholder if no content found
                return "[Meta-commentary - see full text]"

    # FIRST: Check if text starts with "Label [Evaluator] more text" pattern
    # Extract just the label part before the evaluator tag
    # Match common evaluator patterns: [E1], [Treich], [Triech], [Reiley], [Capitan], etc.
    label_before_eval = re.match(r'^([A-Z][^[\n]{10,80}?)\s*\[(?:E\d+|[A-Z][a-z]+)\]', original)
    if label_before_eval:
        label = label_before_eval.group(1).strip()
        label = re.sub(r'[\s:,;—\-]+$', '', label)
        if 3 <= len(label.split()) <= max_words:
            return label

    # Also check for "Label E2: more text" pattern (evaluator without brackets)
    label_before_eval2 = re.match(r'^([A-Z][^\n]{10,80}?)\s+(?:E\d+|DR)\s*:', original)
    if label_before_eval2:
        label = label_before_eval2.group(1).strip()
        label = re.sub(r'[\s:,;—\-]+$', '', label)
        if 3 <= len(label.split()) <= max_words:
            return label

    # Remove evaluator tags, names, and quotes at start
    clean = re.sub(r'^\[?(?:E\d+|Evaluator\s*\d*|DR|NotebookLM|Treich|Triech|Capitan|Reiley|Manheim|Espinosa|Seinkmane|Cloud|Tsai)\]?[\s:\[]*', '', original, flags=re.IGNORECASE)
    clean = re.sub(r'^[""\'\[\(]', '', clean)
    clean = re.sub(r'^Author[\'s]*\s*response:?\s*', '', clean, flags=re.IGNORECASE)

    # Skip if remaining text starts with a name (single capitalized word followed by colon)
    name_check = re.match(r'^([A-Z][a-z]+)\s*:', clean)
    if name_check and len(name_check.group(1)) < 15:
        clean = clean[name_check.end():].strip()

    # Look for quoted phrases that are key terms (handle both straight and curly quotes)
    quote_match = re.search(r'["""\']([^"""\'\n]{15,80})["""\']', clean[:250])
    if quote_match:
        quoted = quote_match.group(1).strip()
        if not re.match(r'^(the |a |an |this )', quoted.lower()):
            return quoted

    # Look for "Label: description" or "Label — description" patterns
    # Use (?<!\w) to avoid matching hyphens in compound words like "pre-registered"
    label_match = re.match(r'^([^:\n]{10,80}?)(?:\s*[:\—]\s*|\s+—\s+)', clean)
    if label_match:
        label = label_match.group(1).strip()
        label = re.sub(r'[""\']+$', '', label)
        # Skip if it ends with a preposition or conjunction (incomplete phrase)
        if not re.search(r'\s+(or|and|of|for|to|in|on|by|with|the|a|an)$', label.lower()):
            if not re.match(r'^[A-Z][a-z]+$', label) and len(label.split()) <= max_words:
                return label

    # Look for descriptive noun phrases (allow hyphens in words)
    patterns = [
        r'((?:Lack of |Missing |Unclear |Weak |Strong |Potential )?[A-Z][a-z]+(?:[-\s]+[a-z-]+){1,5}(?:\s+(?:effects?|bias|issues?|concerns?|limitations?|assumptions?|analysis|validity|problems?|risks?|errors?|size|needs?|checking)))',
        r'((?:Small |Large |Limited |Significant )[a-z-]+(?:\s+[a-z-]+){0,5})',
        r'([A-Z][a-z]+\s+(?:sample|data|model|method|approach|design|analysis|claims?|effects?)(?:\s+[a-z-]+){0,3})',
        r'(Consider\s+[a-z-]+(?:[-\s]+[a-z-]+){0,5})',
        r'(Deviations?\s+from\s+(?:the\s+)?[a-z-]+(?:[-\s]+[a-z-]+){0,5})',
    ]
    for pattern in patterns:
        match = re.search(pattern, clean[:250])
        if match:
            label = match.group(1).strip()
            # Remove trailing preposition phrases like "or degrees of", "and more", etc.
            while True:
                new_label = re.sub(r'\s+(or|and|with|for|to|of|in|on|by|including)(\s+\w+)*\s*$', '', label)
                if new_label == label:
                    break
                label = new_label
            if 10 < len(label) < 120 and len(label.split()) <= max_words:
                return label

    # Take first sentence or clause
    first_clause = re.split(r'(?<=[.!?])\s+|\n', clean)[0].strip()
    first_clause = re.sub(r'^(The |A |An )', '', first_clause)
    first_clause = re.sub(r'\s+(including|or|and|with|for|to|of|in|on|by)\s*$', '', first_clause)

    # Remove evaluator tags from middle/end of text
    first_clause = re.sub(r'\s*\[(?:E\d+|[A-Z][a-z]+)\]\s*', ' ', first_clause)
    first_clause = re.sub(r'\s+(?:E\d+|DR)\s*:.*$', '', first_clause)
    first_clause = re.sub(r'\s+', ' ', first_clause).strip()

    words = first_clause.split()
    if 3 <= len(words) <= max_words:
        return first_clause
    elif len(words) > max_words:
        text = ' '.join(words[:max_words])
        last_break = max(text.rfind(' or '), text.rfind(' and '), text.rfind(', '), text.rfind(' — '))
        if last_break > len(text) // 2:
            return text[:last_break].strip(' ,—')
        return text + '...'

    # Fallback
    key_terms = re.findall(r'[A-Z][a-z]+(?:\s+[a-z]+){0,2}', clean[:150])
    if key_terms:
        return ' / '.join(key_terms[:3])

    words = clean.split()[:max_words]
    return ' '.join(words) + ('...' if len(clean.split()) > max_words else '')

def extract_llm_label(text, max_words=10):
    """Extract a short label from LLM issue text."""
    # Clean up numbering
    clean = re.sub(r'^[\d\.\)]+\s*', '', text.strip())

    # LLM issues often have format "Topic: detailed explanation"
    colon_match = re.match(r'^([^:]{8,60}):\s*', clean)
    if colon_match:
        label = colon_match.group(1).strip()
        # Make sure it's descriptive enough
        if len(label.split()) >= 2:
            return label

    # Take first clause
    first_part = re.split(r'[;.]', clean)[0].strip()
    words = first_part.split()
    if len(words) <= max_words:
        return first_part
    return ' '.join(words[:max_words]) + '...'

def split_human_issues(text):
    """Parse human critique text into individual issues with severity labels."""
    if not text:
        return []

    issues = []
    current_severity = ""
    current_issue_lines = []

    def flush_issue():
        nonlocal current_issue_lines
        if not current_issue_lines:
            return
        full_text = "\n".join(current_issue_lines).strip()
        # Clean leading punctuation/numbering
        cleaned = re.sub(r"^[\-\*\d\.\)\s]+", "", full_text).strip()
        if not cleaned or len(cleaned) < 10:
            current_issue_lines = []
            return
        # Skip if it's just a severity label
        if SEVERITY_HEADER_RE.match(cleaned):
            current_issue_lines = []
            return

        short_label = extract_short_label(cleaned)
        # Skip if this is meta-commentary (the label extraction will return placeholder)
        if short_label == "[Meta-commentary - see full text]":
            current_issue_lines = []
            return

        issues.append({
            "text": cleaned,
            "severity": current_severity,
            "short_label": short_label
        })
        current_issue_lines = []

    lines = text.splitlines()
    i = 0
    while i < len(lines):
        line = lines[i].rstrip()
        raw = line.strip()

        # Check for severity header
        if raw and SEVERITY_HEADER_RE.match(raw):
            flush_issue()
            current_severity = normalize_severity(raw.split(':')[0].split()[0])
            i += 1
            continue

        # Check for numbered issue start
        if ENUM_RE.match(raw):
            flush_issue()
            # Remove the number prefix
            issue_text = re.sub(r"^\s*\d+[\.\)]\s*", "", raw)
            current_issue_lines = [issue_text] if issue_text else []
            i += 1
            continue

        # Check for issue label line (short line followed by longer explanation)
        # e.g., "Ceiling effects with Likert scale measures"
        if raw and not EVALUATOR_RE.match(raw) and len(raw) < 80:
            # Look ahead - if next non-empty line starts with [E or is indented, this is a label
            next_idx = i + 1
            while next_idx < len(lines) and not lines[next_idx].strip():
                next_idx += 1
            if next_idx < len(lines):
                next_line = lines[next_idx].strip()
                if (EVALUATOR_RE.match(next_line) or
                    lines[next_idx].startswith(' ') or
                    next_line.startswith('"') or
                    next_line.startswith("'")):
                    # This short line is likely a label for a new issue
                    flush_issue()
                    current_issue_lines = [raw]
                    i += 1
                    continue

        # Empty line - might signal new issue in some formats
        if not raw:
            # Don't flush on every empty line - only if we have significant content
            if current_issue_lines and len("\n".join(current_issue_lines)) > 100:
                # Check if next content looks like new issue
                next_idx = i + 1
                while next_idx < len(lines) and not lines[next_idx].strip():
                    next_idx += 1
                if next_idx < len(lines):
                    next_line = lines[next_idx].strip()
                    if ENUM_RE.match(next_line) or (len(next_line) < 80 and not EVALUATOR_RE.match(next_line)):
                        flush_issue()
            i += 1
            continue

        # Regular content line - add to current issue
        if current_issue_lines:
            current_issue_lines.append(raw)
        else:
            current_issue_lines = [raw]
        i += 1

    flush_issue()
    return issues

# Load comparison data and results
comparison_file = Path("results/key_issues_comparison.json")
results_file = Path("results/key_issues_comparison_results.json")
matched_file = Path("results/key_issues_matched.json")

parsed_issues = {}
comparison_results = {}
matched_data = {}

if comparison_file.exists():
    comparison = json.loads(comparison_file.read_text())
    for item in comparison:
        paper_id = item.get("gpt_paper", "")
        coda_critique = item.get("coda_critique", "")
        parsed_issues[paper_id] = split_human_issues(coda_critique)

if results_file.exists():
    results = json.loads(results_file.read_text())
    for r in results:
        paper_id = r.get("gpt_paper", "")
        comparison_results[paper_id] = r.get("comparison", {})

# Load embedding-matched pairs
if matched_file.exists():
    matched_data = json.loads(matched_file.read_text())
```

We matched **`r n_papers` papers** with both GPT-5.2 Pro key issues and human expert critiques.

## Overview Statistics

```{r}
#| label: tbl-overview-stats
#| tbl-cap: "Summary of key issues comparison data"

if (!is.null(comparison_data) && n_papers > 0) {
  overview <- comparison_data |>
    mutate(
      num_gpt_issues = sapply(gpt_key_issues, length)
    ) |>
    summarise(
      `Papers Compared` = n(),
      `Avg GPT Issues per Paper` = round(mean(num_gpt_issues), 1),
      `Min GPT Issues` = min(num_gpt_issues),
      `Max GPT Issues` = max(num_gpt_issues),
      `Avg Coda Critique Length (chars)` = round(mean(coda_critique_length)),
      `Min Coda Length` = min(coda_critique_length),
      `Max Coda Length` = max(coda_critique_length)
    ) |>
    pivot_longer(everything(), names_to = "Metric", values_to = "Value")

  kable(overview, align = c("l", "r"))
}
```

## Paper-by-Paper Comparison {#sec-paper-comparison}

Each paper section shows:

- **Assessment badge**: Coverage and precision metrics (from LLM-based comparison)
- **Human Issues table**: Parsed individual issues with severity labels (🔴 Necessary, 🟠 Optional, ⚪ Unsure)
- **LLM Issues table**: GPT-5.2 Pro key issues (ordered by importance)
- **Full critique text**: Expandable section with complete human critique

```{r}
#| label: render-comparisons
#| results: asis
#| code-fold: true
#| code-summary: "Render paper comparisons"

# Access parsed issues from Python
parsed_issues_list <- py$parsed_issues

if (!is.null(comparison_data) && n_papers > 0) {

  # Check if LLM assessment fields exist
  has_assessment <- "coverage_pct" %in% names(comparison_data)

  for (i in 1:nrow(comparison_data)) {
    row <- comparison_data[i, ]
    paper_id <- row$gpt_paper

    # Paper header
    cat(sprintf("\n### %s\n\n", paper_id))
    cat(sprintf("**Coda title:** %s\n\n", row$coda_title))

    # Show LLM assessment badge if available
    if (has_assessment && !is.na(row$coverage_pct)) {
      # Determine badge color based on coverage/precision
      coverage_val <- as.numeric(row$coverage_pct)
      precision_val <- as.numeric(row$precision_pct)
      avg_score <- (coverage_val + precision_val) / 2
      if (avg_score >= 70) {
        badge_class <- "callout-tip"
      } else if (avg_score >= 40) {
        badge_class <- "callout-note"
      } else {
        badge_class <- "callout-warning"
      }
      cat(sprintf("::: {.%s appearance=\"simple\"}\n", badge_class))
      cat(sprintf("**Coverage:** %s%% | **Precision:** %s%%\n",
                  row$coverage_pct, row$precision_pct))
      cat(":::\n\n")
    }

    # Get parsed issues and comparison results for this paper
    human_issues <- parsed_issues_list[[paper_id]]
    paper_comparison <- py$comparison_results[[paper_id]]
    gpt_issues <- row$gpt_key_issues[[1]]
    coverage <- if (!is.null(paper_comparison$coverage_pct)) paper_comparison$coverage_pct else NA
    missed_issues <- if (!is.null(paper_comparison$missed_issues)) paper_comparison$missed_issues else list()

    # ---- MATCHED ISSUE COMPARISON ----
    cat("#### Issue Comparison\n\n")

    # Get matched data for this paper
    paper_matched <- py$matched_data[[paper_id]]

    if (!is.null(paper_matched) && length(paper_matched$matched_pairs) > 0) {
      matched_pairs <- paper_matched$matched_pairs
      human_issues_data <- paper_matched$human_issues
      llm_issues_data <- paper_matched$llm_issues
      unmatched_human <- paper_matched$unmatched_human
      unmatched_llm <- paper_matched$unmatched_llm

      # Helper: get LLM label from text
      get_llm_label <- function(text) {
        text_clean <- gsub("^[0-9]+[)\\.]\\s*", "", text)
        first_colon <- regexpr(":", text_clean)[1]
        if (first_colon > 0 && first_colon < 60) {
          return(substr(text_clean, 1, first_colon - 1))
        }
        return(truncate_text(text_clean, 50))
      }

      # ---- MATCHED ISSUES ----
      cat('<div style="margin-bottom:1.5em;">\n')
      cat('<h5 style="color:#2e7d32;">Matched Issues</h5>\n')
      cat('<p style="font-size:0.85em;color:#888;margin-bottom:1em;font-style:italic;"><strong>Note:</strong> Match quality is based on cosine similarity of sentence embeddings—a rough heuristic that may miss semantic matches or flag false positives. Use LLM-based comparison for more accurate matching.</p>\n')

      for (pair in matched_pairs) {
        h_idx <- pair$human_issue_index
        llm_indices <- pair$llm_issue_indices
        match_quality <- pair$match_quality

        if (length(llm_indices) > 0 && h_idx <= length(human_issues_data)) {
          h_issue <- human_issues_data[[h_idx]]
          h_text <- h_issue$text
          h_severity <- h_issue$severity
          badge <- severity_badge(h_severity)

          # Get human label
          h_label <- if (!is.null(human_issues) && h_idx <= length(human_issues)) {
            human_issues[[h_idx]]$short_label
          } else {
            truncate_text(h_text, 50)
          }
          h_label_escaped <- gsub("<", "&lt;", gsub(">", "&gt;", h_label))
          h_text_escaped <- gsub("<", "&lt;", gsub(">", "&gt;", h_text))
          border_color <- if (grepl("necessary", tolower(h_severity))) "#e74c3c" else if (grepl("optional", tolower(h_severity))) "#f39c12" else "#95a5a6"

          # Match quality badge color
          quality_color <- if (match_quality >= 70) "#2e7d32" else if (match_quality >= 50) "#f9a825" else "#c62828"

          cat('<div style="margin:0.75em 0;padding:0.75em;background:#f8fdf8;border:1px solid #c8e6c9;border-radius:5px;">\n')

          # Header with match quality
          cat(sprintf('<div style="display:flex;justify-content:space-between;align-items:center;margin-bottom:0.5em;">\n'))
          cat(sprintf('<span><strong>H%d</strong> %s <strong>%s</strong></span>\n', h_idx, badge, h_label_escaped))
          cat(sprintf('<span style="background:%s;color:white;padding:2px 8px;border-radius:10px;font-size:0.8em;">Match: %d%%</span>\n', quality_color, match_quality))
          cat('</div>\n')

          # Human text (expandable)
          cat('<details style="margin:0.5em 0;">\n')
          cat('<summary style="font-size:0.85em;color:#666;cursor:pointer;">📋 Human expert text</summary>\n')
          cat(sprintf('<div style="margin-top:0.5em;padding:0.5em;background:#fffbf0;border-left:3px solid %s;font-size:0.9em;">%s</div>\n', border_color, h_text_escaped))
          cat('</details>\n')

          # Matched LLM issues
          cat('<details style="margin:0.5em 0;">\n')
          llm_ids_str <- paste0("L", llm_indices, collapse=", ")
          cat(sprintf('<summary style="font-size:0.85em;color:#666;cursor:pointer;">🤖 LLM coverage (%s)</summary>\n', llm_ids_str))
          cat('<div style="margin-top:0.5em;">\n')
          for (l_idx in llm_indices) {
            if (l_idx <= length(llm_issues_data)) {
              l_text <- llm_issues_data[[l_idx]]
              l_label <- get_llm_label(l_text)
              l_label_escaped <- gsub("<", "&lt;", gsub(">", "&gt;", l_label))
              l_text_escaped <- gsub("<", "&lt;", gsub(">", "&gt;", l_text))
              cat(sprintf('<div style="padding:0.5em;background:#f0f7ff;border-left:3px solid #4e79a7;margin-bottom:0.3em;font-size:0.9em;"><strong>L%d:</strong> %s</div>\n', l_idx, l_text_escaped))
            }
          }
          cat('</div>\n')
          cat('</details>\n')

          cat('</div>\n')
        }
      }
      cat('</div>\n\n')

      # ---- UNMATCHED HUMAN ISSUES ----
      if (length(unmatched_human) > 0) {
        cat('<details style="margin-bottom:1.5em;">\n')
        cat(sprintf('<summary style="color:#c62828;font-weight:bold;cursor:pointer;">⚠️ Unmatched Human Issues (%d missed by LLM)</summary>\n', length(unmatched_human)))
        cat('<div style="margin-top:0.5em;">\n')
        for (h_idx in unmatched_human) {
          if (h_idx <= length(human_issues_data)) {
            h_issue <- human_issues_data[[h_idx]]
            h_text <- h_issue$text
            h_severity <- h_issue$severity
            badge <- severity_badge(h_severity)
            h_label <- if (!is.null(human_issues) && h_idx <= length(human_issues)) {
              human_issues[[h_idx]]$short_label
            } else {
              truncate_text(h_text, 50)
            }
            h_label_escaped <- gsub("<", "&lt;", gsub(">", "&gt;", h_label))
            h_text_escaped <- gsub("<", "&lt;", gsub(">", "&gt;", h_text))
            border_color <- if (grepl("necessary", tolower(h_severity))) "#e74c3c" else if (grepl("optional", tolower(h_severity))) "#f39c12" else "#95a5a6"

            cat(sprintf('<div style="margin:0.5em 0;padding:0.5em;background:#ffebee;border-left:4px solid #c62828;">\n'))
            cat(sprintf('<strong>H%d</strong> %s <strong>%s</strong>\n', h_idx, badge, h_label_escaped))
            cat('<details style="margin-top:0.3em;">\n')
            cat('<summary style="font-size:0.85em;color:#666;cursor:pointer;">Show full text</summary>\n')
            cat(sprintf('<div style="margin-top:0.5em;font-size:0.9em;color:#444;">%s</div>\n', h_text_escaped))
            cat('</details>\n')
            cat('</div>\n')
          }
        }
        cat('</div>\n')
        cat('</details>\n\n')
      }

      # ---- UNMATCHED LLM ISSUES ----
      if (length(unmatched_llm) > 0) {
        cat('<details style="margin-bottom:1.5em;">\n')
        cat(sprintf('<summary style="color:#1565c0;font-weight:bold;cursor:pointer;">📋 Extra LLM Issues (%d not in human critique)</summary>\n', length(unmatched_llm)))
        cat('<div style="margin-top:0.5em;">\n')
        for (l_idx in unmatched_llm) {
          if (l_idx <= length(llm_issues_data)) {
            l_text <- llm_issues_data[[l_idx]]
            l_label <- get_llm_label(l_text)
            l_label_escaped <- gsub("<", "&lt;", gsub(">", "&gt;", l_label))
            l_text_escaped <- gsub("<", "&lt;", gsub(">", "&gt;", l_text))

            cat('<div style="margin:0.5em 0;padding:0.5em;background:#e3f2fd;border-left:4px solid #1565c0;">\n')
            cat(sprintf('<strong>L%d</strong> <strong>%s</strong>\n', l_idx, l_label_escaped))
            cat('<details style="margin-top:0.3em;">\n')
            cat('<summary style="font-size:0.85em;color:#666;cursor:pointer;">Show full text</summary>\n')
            cat(sprintf('<div style="margin-top:0.5em;font-size:0.9em;color:#444;">%s</div>\n', l_text_escaped))
            cat('</details>\n')
            cat('</div>\n')
          }
        }
        cat('</div>\n')
        cat('</details>\n\n')
      }

      # ---- SUMMARY STATS ----
      n_matched <- sum(sapply(matched_pairs, function(p) length(p$llm_issue_indices) > 0))
      n_human <- length(human_issues_data)
      n_llm <- length(llm_issues_data)
      cat('<div style="background:#f5f5f5;padding:0.75em;border-radius:5px;font-size:0.9em;">\n')
      cat(sprintf('<strong>Summary:</strong> %d/%d human issues matched | %d/%d LLM issues matched | %d LLM extras\n',
                  n_matched, n_human, n_llm - length(unmatched_llm), n_llm, length(unmatched_llm)))
      cat('</div>\n\n')

    } else if (!is.null(human_issues) && length(human_issues) > 0) {
      # Fallback: no matched data, show separate lists
      cat('<p style="color:#666;font-style:italic;">Embedding-based matching not available for this paper.</p>\n\n')

      # Show human issues
      cat('<h5 style="color:#e65100;">Human Expert Issues</h5>\n')
      for (j in seq_along(human_issues)) {
        issue <- human_issues[[j]]
        badge <- severity_badge(issue$severity)
        label <- if (!is.null(issue$short_label)) issue$short_label else truncate_text(issue$text, 60)
        label_escaped <- gsub("<", "&lt;", gsub(">", "&gt;", label))
        text_escaped <- gsub("<", "&lt;", gsub(">", "&gt;", issue$text))
        border_color <- if (grepl("necessary", tolower(issue$severity))) "#e74c3c" else if (grepl("optional", tolower(issue$severity))) "#f39c12" else "#95a5a6"

        cat(sprintf('<div style="margin:0.5em 0;padding:0.5em;background:#fffbf0;border-left:4px solid %s;">\n', border_color))
        cat(sprintf('<strong>H%d</strong> %s <strong>%s</strong>\n', j, badge, label_escaped))
        cat('<details style="margin-top:0.3em;">\n')
        cat('<summary style="font-size:0.85em;color:#666;cursor:pointer;">Show full text</summary>\n')
        cat(sprintf('<div style="margin-top:0.5em;font-size:0.9em;color:#444;">%s</div>\n', text_escaped))
        cat('</details>\n')
        cat('</div>\n')
      }

      # Show LLM issues
      cat('\n<h5 style="color:#1565c0;">LLM Issues</h5>\n')
      for (j in seq_along(gpt_issues)) {
        issue <- gpt_issues[j]
        issue_clean <- gsub("^[0-9]+[)\\.]\\s*", "", issue)
        first_colon <- regexpr(":", issue_clean)[1]
        label <- if (first_colon > 0 && first_colon < 50) substr(issue_clean, 1, first_colon - 1) else truncate_text(issue_clean, 50)
        label_escaped <- gsub("<", "&lt;", gsub(">", "&gt;", label))
        text_escaped <- gsub("<", "&lt;", gsub(">", "&gt;", issue_clean))

        cat('<div style="margin:0.5em 0;padding:0.5em;background:#f0f7ff;border-left:4px solid #4e79a7;">\n')
        cat(sprintf('<strong>L%d</strong> <strong>%s</strong>\n', j, label_escaped))
        cat('<details style="margin-top:0.3em;">\n')
        cat('<summary style="font-size:0.85em;color:#666;cursor:pointer;">Show full text</summary>\n')
        cat(sprintf('<div style="margin-top:0.5em;font-size:0.9em;color:#444;">%s</div>\n', text_escaped))
        cat('</details>\n')
        cat('</div>\n')
      }
      cat('\n')

    } else {
      # No parsed human issues - show raw
      cat("*Human critique not parsed into individual issues.*\n\n")
      cat('<details><summary>View full human critique</summary>\n\n')
      cat(gsub("<", "&lt;", gsub(">", "&gt;", row$coda_critique)))
      cat('\n</details>\n\n')

      # Still show LLM issues
      if (length(gpt_issues) > 0) {
        cat("**LLM Issues:**\n\n")
        for (j in seq_along(gpt_issues)) {
          issue <- gpt_issues[j]
          issue_clean <- gsub("^[0-9]+[)\\.]\\s*", "", issue)
          issue_clean <- gsub("^\\. ", "", issue_clean)
          cat(sprintf("- **L%d:** %s\n", j, truncate_text(issue_clean, 100)))
        }
        cat("\n")
      }
    }

    # Show LLM justification if available
    if (has_assessment && !is.na(row$coverage_pct)) {
      if (!is.null(row$overall_justification) && nchar(row$overall_justification) > 0) {
        cat("::: {.callout-note collapse='true'}\n")
        cat("## LLM Comparison Notes\n\n")
        cat(row$overall_justification)
        cat("\n:::\n\n")
      }
    }

    cat("---\n\n")
  }
}
```

## LLM-Based Assessment {#sec-llm-assessment}

The comparison between GPT key issues and human critiques was assessed using GPT-5.2 Pro, which evaluated coverage (what proportion of human concerns GPT identified) and precision (whether GPT issues are substantive).

**Note**: Any interpretation or narrative commentary in this section is written by an LLM (Codex). The values themselves come from the GPT-5.2 Pro comparison outputs.

```{r}
#| label: check-llm-results
#| code-fold: true

# Check if LLM comparison results are available
has_llm_results <- !is.null(comparison_data) &&
                   n_papers > 0 &&
                   "coverage_pct" %in% names(comparison_data) &&
                   any(!is.na(comparison_data$coverage_pct))
```

```{r}
#| label: llm-assessment-summary
#| tbl-cap: "LLM assessment of GPT vs human critique alignment"
#| eval: !expr has_llm_results

if (has_llm_results) {
  # Filter to papers with valid LLM results
  llm_results <- comparison_data |>
    filter(!is.na(coverage_pct) & !is.na(precision_pct))

  summary_stats <- llm_results |>
    summarise(
      `Papers Assessed` = n(),
      `Mean Coverage (%)` = round(mean(coverage_pct, na.rm = TRUE), 1),
      `Mean Precision (%)` = round(mean(precision_pct, na.rm = TRUE), 1),
      `Coverage Range` = paste0(min(coverage_pct, na.rm = TRUE), "-", max(coverage_pct, na.rm = TRUE)),
      `Precision Range` = paste0(min(precision_pct, na.rm = TRUE), "-", max(precision_pct, na.rm = TRUE))
    ) |>
    mutate(across(everything(), as.character)) |>
    pivot_longer(everything(), names_to = "Metric", values_to = "Value")

  kable(summary_stats, align = c("l", "r"))
}
```

```{r}
#| label: fig-coverage-precision
#| fig-cap: "Coverage vs Precision across papers (LLM-assessed)"
#| fig-width: 9
#| fig-height: 7
#| eval: !expr has_llm_results

if (has_llm_results) {
  llm_results <- comparison_data |>
    filter(!is.na(coverage_pct) & !is.na(precision_pct)) |>
    mutate(paper_short = str_trunc(gpt_paper, 25))

  ggplot(llm_results, aes(x = coverage_pct, y = precision_pct)) +
    geom_point(size = 4, color = UJ_ORANGE, alpha = 0.8) +
    geom_text(aes(label = paper_short), hjust = -0.1, vjust = 0.5, size = 2.5, check_overlap = TRUE) +
    geom_vline(xintercept = mean(llm_results$coverage_pct), linetype = "dashed", color = UJ_BLUE) +
    geom_hline(yintercept = mean(llm_results$precision_pct), linetype = "dashed", color = UJ_BLUE) +
    scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
    scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
    labs(
      x = "Coverage (%): Human issues captured by GPT",
      y = "Precision (%): GPT issues that are substantive",
      subtitle = "Dashed lines = means"
    ) +
    theme_uj()
}
```

```{r}
#| label: tbl-paper-ratings
#| tbl-cap: "Per-paper LLM assessment results"
#| eval: !expr has_llm_results

if (has_llm_results) {
  ratings_table <- comparison_data |>
    filter(!is.na(coverage_pct)) |>
    mutate(
      Paper = str_trunc(gpt_paper, 35),
      `Coverage (%)` = coverage_pct,
      `Precision (%)` = precision_pct,
      Rating = overall_rating
    ) |>
    select(Paper, `Coverage (%)`, `Precision (%)`, Rating) |>
    arrange(desc(`Coverage (%)`))

  kable(ratings_table, align = c("l", "r", "r", "l"))
}
```

```{r}
#| label: fig-rating-distribution
#| fig-cap: "Distribution of overall ratings across papers"
#| fig-width: 8
#| fig-height: 4
#| eval: !expr has_llm_results

if (has_llm_results) {
  rating_counts <- comparison_data |>
    filter(!is.na(overall_rating)) |>
    mutate(
      Rating = factor(overall_rating, levels = c("Poor", "Fair", "Good", "Excellent"))
    ) |>
    count(Rating, .drop = FALSE)

  # Define colors for ratings
  rating_colors <- c("Poor" = "#e74c3c", "Fair" = "#f39c12", "Good" = "#27ae60", "Excellent" = "#2ecc71")

  ggplot(rating_counts, aes(x = Rating, y = n, fill = Rating)) +
    geom_col(alpha = 0.9) +
    geom_text(aes(label = n), vjust = -0.3, size = 5) +
    scale_fill_manual(values = rating_colors, drop = FALSE) +
    labs(
      x = "Overall Rating",
      y = "Number of Papers"
    ) +
    theme_uj() +
    theme(legend.position = "none")
}
```

::: {.callout-note collapse="true"}
## Interpretation Guide

-   **Coverage**: Percentage of human-identified issues that GPT also captured (in some form). Higher = GPT missed fewer human concerns.
-   **Precision**: Percentage of GPT issues that are substantive rather than generic or spurious. Higher = GPT's critiques are more targeted.
-   **Overall Rating**: Qualitative assessment (Excellent/Good/Fair/Poor) based on both metrics.
:::

## Summary and Observations

```{r}
#| label: summary-analysis
#| code-fold: true

if (!is.null(comparison_data) && n_papers > 0) {

  # Calculate some summary statistics
  issue_counts <- comparison_data |>
    mutate(
      num_gpt_issues = sapply(gpt_key_issues, length),
      paper_short = str_trunc(gpt_paper, 30)
    ) |>
    select(paper_short, num_gpt_issues, coda_critique_length)

  # Calculate severity distribution from parsed issues
  all_human_issues <- do.call(rbind, lapply(names(parsed_issues_list), function(paper_id) {
    issues <- parsed_issues_list[[paper_id]]
    if (is.null(issues) || length(issues) == 0) return(NULL)
    data.frame(
      paper_id = paper_id,
      severity = sapply(issues, function(x) x$severity),
      stringsAsFactors = FALSE
    )
  }))

  if (!is.null(all_human_issues) && nrow(all_human_issues) > 0) {
    severity_dist <- all_human_issues |>
      mutate(
        severity = case_when(
          grepl("necessary", tolower(severity)) ~ "Necessary",
          grepl("optional", tolower(severity)) ~ "Optional",
          severity != "" ~ "Unsure",
          TRUE ~ "Unclassified"
        ),
        severity = factor(severity, levels = c("Necessary", "Optional", "Unsure", "Unclassified"))
      ) |>
      count(severity, .drop = FALSE)
  }
}
```

### Aggregate Statistics

```{r}
#| label: tbl-aggregate-stats
#| tbl-cap: "Aggregate statistics across all papers"
#| results: asis

if (!is.null(comparison_data) && n_papers > 0) {
  # Count human issues by severity
  total_human_issues <- if (!is.null(all_human_issues)) nrow(all_human_issues) else 0
  total_gpt_issues <- sum(issue_counts$num_gpt_issues)

  necessary_count <- if (!is.null(all_human_issues)) sum(grepl("necessary", tolower(all_human_issues$severity))) else 0
  optional_count <- if (!is.null(all_human_issues)) sum(grepl("optional", tolower(all_human_issues$severity))) else 0

  aggregate_stats <- data.frame(
    Metric = c(
      "Papers Compared",
      "Total Human Issues Parsed",
      "  - Necessary",
      "  - Optional",
      "  - Unsure/Other",
      "Total LLM Issues",
      "Avg Human Issues per Paper",
      "Avg LLM Issues per Paper"
    ),
    Value = c(
      n_papers,
      total_human_issues,
      necessary_count,
      optional_count,
      total_human_issues - necessary_count - optional_count,
      total_gpt_issues,
      round(total_human_issues / n_papers, 1),
      round(total_gpt_issues / n_papers, 1)
    )
  )

  kable(aggregate_stats, align = c("l", "r"))
}
```

```{r}
#| label: fig-severity-distribution
#| fig-cap: "Distribution of human issue severity labels"
#| fig-width: 8
#| fig-height: 4

if (!is.null(all_human_issues) && nrow(all_human_issues) > 0) {
  # Define severity colors
  sev_colors <- c("Necessary" = "#e74c3c", "Optional" = "#f39c12", "Unsure" = "#95a5a6", "Unclassified" = "#bdc3c7")

  ggplot(severity_dist, aes(x = severity, y = n, fill = severity)) +
    geom_col(alpha = 0.9) +
    geom_text(aes(label = n), vjust = -0.3, size = 5) +
    scale_fill_manual(values = sev_colors, drop = FALSE) +
    labs(
      x = "Severity Label",
      y = "Number of Issues"
    ) +
    theme_uj() +
    theme(legend.position = "none")
}
```

### Issue Count Distribution

```{r}
#| label: fig-issue-counts
#| fig-cap: "Number of key issues identified by GPT-5.2 Pro per paper"
#| fig-width: 10
#| fig-height: 5

if (!is.null(comparison_data) && n_papers > 0) {

  issue_counts |>
    ggplot(aes(x = reorder(paper_short, num_gpt_issues), y = num_gpt_issues)) +
    geom_col(fill = UJ_ORANGE, alpha = 0.8) +
    geom_hline(yintercept = mean(issue_counts$num_gpt_issues),
               linetype = "dashed", color = UJ_BLUE, linewidth = 1) +
    coord_flip() +
    labs(
      x = NULL,
      y = "Number of Key Issues",
      subtitle = paste0("Dashed line = mean (", round(mean(issue_counts$num_gpt_issues), 1), " issues)")
    ) +
    theme_uj()
}
```

### Coda Critique Length vs GPT Issue Count

```{r}
#| label: fig-length-scatter
#| fig-cap: "Relationship between human critique detail and GPT issue count"
#| fig-width: 8
#| fig-height: 6

if (!is.null(comparison_data) && n_papers > 0) {

  ggplot(issue_counts, aes(x = coda_critique_length, y = num_gpt_issues)) +
    geom_point(size = 3, color = UJ_ORANGE, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, color = UJ_BLUE, alpha = 0.2) +
    labs(
      x = "Human Critique Length (characters)",
      y = "GPT-5.2 Pro Issue Count"
    ) +
    theme_uj()
}
```

### Observable Structural Differences

The following are **verifiable structural differences** between the two data sources (not assessments of quality or coverage):

| Aspect | GPT-5.2 Pro | Human Expert (Coda) |
|-----------------|---------------------|----------------------------------|
| Format | Numbered bullet points (array of strings) | Free-form prose |
| Structure | Ordered list (prompted: "most to least important") | Often uses severity labels ("Necessary", "Optional but important") |
| Source attribution | None (single model output) | Often cites specific evaluators (E1, E2, names) |
| Length | Constrained by prompt (\~10-12 issues) | Unconstrained, highly variable |

### Questions for Manual Review

The side-by-side comparisons above are provided for **manual expert assessment**. Key questions to investigate:

1.  **Coverage**: What proportion of human-identified issues appear in the GPT output (in some form, to some extent)? (This is the focal question).
2.  **Precision**: Are GPT issues substantive or does the model identify spurious/generic concerns?
3.  **Severity alignment**: Does GPT's importance ordering correlate with human severity labels?
4.  **Missed issues**: Are there critical human concerns that GPT systematically misses?
5.  **Novel issues**: Does GPT surface valid concerns that humans overlooked?

```{r}
#| label: assessment-status-callout
#| results: asis

if (has_llm_results) {
  cat('::: {.callout-tip}\n')
  cat('## LLM Assessment Complete\n\n')
  cat('The comparisons above have been assessed using GPT-5.2 Pro. See the [LLM-Based Assessment](#sec-llm-assessment) section for coverage and precision metrics.\n')
  cat(':::\n')
} else {
  cat('::: {.callout-warning}\n')
  cat('## No Automated Assessment Yet\n\n')
  cat('This page displays raw data for manual review. The claims about coverage, precision, and alignment **have not been assessed**—they require either:\n\n')
  cat('- Manual expert review of each paper\'s comparisons\n')
  cat('- LLM-based comparison (run the `key-issues-comparison` chunk in `methods.qmd`)\n\n')
  cat('Any conclusions about "high overlap" or "systematic differences" should emerge from that analysis, not be assumed.\n')
  cat(':::\n')
}
```

## Manual Annotation Tool

To systematically assess concordance between human and LLM critiques, use the **Issue Annotation Tool**. (Jan 8 2026: We are working on building a hosted tool, for now follow the instructions below.)

``` bash
# 1. Generate annotation data (parses human critiques into individual issues)
python3 tools/build_issue_annotation_data.py

# 2. Open the annotation UI in a browser
open tools/issue_annotation_ui/index.html
```

**Annotation workflow:**

For each human-identified issue:

1.  **Match score (0-1)**: How well do LLM issues capture this concern?

    -   0 = Not addressed at all
    -   0.5 = Partially captured or tangentially related
    -   1 = Fully captured by one or more LLM issues

2.  **Confidence (0-1)**: How certain are you of this assessment?

3.  **Context flag**: Check this box if the human critique references information the LLM didn't have (e.g., appendix or preregistration materials not shared with the LLM at the point this evaluation was done)

4.  **Link to LLM issues**: Select which LLM issues (L1, L2, ...) correspond to this human issue

5.  **Discussion**: Explain your reasoning, note partial matches, or flag ambiguities

**Export** annotations as JSON/CSV for analysis. Annotations are auto-saved to browser localStorage.

::: callout-note
## Annotation Data Schema

The tool parses human critiques using heuristics:

-   **Severity labels** normalized to: `necessary`, `optional`, `unsure`
-   **Evaluator attributions** (E1, E2, DR) preserved in issue text
-   **Issue boundaries** detected via enumeration, sentence breaks, and section headers

Review parsed issues in the UI and edit/add/remove as needed before annotating.
:::
