---
title: "Results"
format: 
  html: default
  # pdf: default
engine: knitr
--- 

```{r} 
#| label: setup call
#| code-summary: "Include global setup and parameters"

source("setup_params.R")
```

```{r}
#| label: setup-libs

library("tidyverse")
library("janitor")
library("stringr")
library("stringi") #probably redundant?
library("lubridate")
#library("readr") #redundant?
library("here") 
library("knitr") 
library("ggforce") 
library("ggrepel")
library("glue")
library("ggalluvial")
library("scales")
library("viridis")       # For colorblind-friendly palettes
library("ggbreak")


```

```{r}
#| label: setup-themes

 
UJ_ORANGE <- "#f19e4b"   # LLM
UJ_GREEN  <- "#99bb66"   # Human

theme_uj <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot",
      legend.position = "bottom"
    )
}

 
```



```{r}
#| label: functions-summaries

fix_bounds <- function(df, lo, hi) {
  lo2 <- suppressWarnings(as.numeric(df[[lo]]))
  hi2 <- suppressWarnings(as.numeric(df[[hi]]))
  swap <- !is.na(lo2) & !is.na(hi2) & (lo2 > hi2)
  df[swap, c(lo, hi)] <- df[swap, c(hi, lo)]
  df
}

safe_min <- function(ci, pts) {
  if (length(ci) == 0 || all(is.na(ci))) suppressWarnings(min(pts, na.rm = TRUE)) else suppressWarnings(min(ci, na.rm = TRUE))
}
safe_max <- function(ci, pts) {
  if (length(ci) == 0 || all(is.na(ci))) suppressWarnings(max(pts, na.rm = TRUE)) else suppressWarnings(max(ci, na.rm = TRUE))
}

# Quantile/equal binning shared by both raters (for κ)
bin_together <- function(a, b, n_bins = 5, strategy = c("quantile","equal")) {
  strategy <- match.arg(strategy)
  x <- c(a, b)
  x <- x[is.finite(x)]
  # discrete-ish fallback
  if (length(unique(x)) <= max(3, n_bins)) {
    u <- sort(unique(x))
    f <- function(v) match(v, u) - 1L
    return(list(a_bin = f(a), b_bin = f(b), k = length(u)))
  }
  if (strategy == "quantile") {
    qs <- unique(quantile(x, probs = seq(0, 1, length.out = n_bins + 1), na.rm = TRUE))
    if (length(qs) - 1L < 2L) {
      strategy <- "equal"
    } else {
      edges <- qs
    }
  }
  if (strategy == "equal") {
    lo <- min(x, na.rm = TRUE); hi <- max(x, na.rm = TRUE)
    edges <- seq(lo, hi, length.out = n_bins + 1)
  }
  # widen to include endpoints robustly
  edges[1] <- edges[1] - 1e-9
  edges[length(edges)] <- edges[length(edges)] + 1e-9
  a_bin <- cut(a, breaks = edges, include.lowest = TRUE, labels = FALSE) - 1L
  b_bin <- cut(b, breaks = edges, include.lowest = TRUE, labels = FALSE) - 1L
  k <- max(c(a_bin, b_bin), na.rm = TRUE) + 1L
  list(a_bin = a_bin, b_bin = b_bin, k = k)
}

# Cohen κ (unweighted, linear, quadratic)  

weighted_kappa <- function(a_bin, b_bin, k = NULL, weights = c("quadratic","linear","unweighted")) {
  weights <- match.arg(weights)
  a <- as.integer(a_bin); b <- as.integer(b_bin)
  keep <- is.finite(a) & is.finite(b)
  a <- a[keep]; b <- b[keep]
  if (!length(a)) return(NA_real_)
  if (is.null(k)) k <- max(c(a,b)) + 1L

  M <- matrix(0, nrow = k, ncol = k)
  for (i in seq_along(a)) M[a[i]+1L, b[i]+1L] <- M[a[i]+1L, b[i]+1L] + 1
  if (sum(M) == 0) return(NA_real_)
  M <- M / sum(M)
  r <- rowSums(M); csum <- colSums(M)
  E <- r %*% t(csum)

  I <- matrix(rep(0:(k-1), times = k), nrow = k)
  J <- t(I)
  if (weights == "quadratic") {
    W <- ((I - J)^2) / ((k - 1)^2)
  } else if (weights == "linear") {
    W <- abs(I - J) / (k - 1)
  } else {
    W <- 1 - diag(1, k)  # 1 off-diagonal, 0 on diagonal
  }
  num <- sum(W * M); den <- sum(W * E)
  if (den == 0) NA_real_ else 1 - num/den
}



metrics_meta <- readr::read_csv(here("data", "metrics_meta.csv"), show_col_types = FALSE) |>
  janitor::clean_names()

# Model label from meta 
models_run   <- metrics_meta |>
  dplyr::distinct(model) |>
  dplyr::pull(model) |> na.omit()
currentmodel <- if (length(models_run) == 1) models_run else paste(models_run, collapse = ", ")

# Token summary (input + output + reasoning tokens when available)
metrics_meta <- metrics_meta |>
  dplyr::mutate(total_tokens = dplyr::coalesce(input_tokens, 0) +
                                dplyr::coalesce(output_tokens, 0) +
                                dplyr::coalesce(reasoning_tokens, 0))

tok_sum <- metrics_meta |>
  dplyr::summarise(
    n_papers = dplyr::n_distinct(paper),
    median_tokens = stats::median(total_tokens, na.rm = TRUE),
    mean_tokens   = mean(total_tokens, na.rm = TRUE)
  )



reasoning_example = metrics_meta |>
  filter(!is.na(reasoning_summary), 
         paper == "Williams et al. 2024")


```

Here we present preliminary results, starting with a comparison of the LLM‑generated quantitative ratings (model: `r print(currentmodel)`, see the[(previous section](methods.qmd)) with human evaluations across [the Unjournal's criteria](https://globalimpact.gitbook.io/the-unjournal-project-and-communication-space/policies-projects-evaluation-workflow/evaluation/guidelines-for-evaluators#undefined-1). 



## Journal ranking tiers


```{r}
#| label: load-data-tiers


# paper_authors <- read_delim(here("data", "paper_authors.csv"), delim = ",")

# Mapping paper keys - short titles
UJmap <- read_delim(here("data", "UJ_map.csv"), delim = ";") |>
  mutate(label_paper_title = research,
         label_paper = paper) |>
  select(c("label_paper_title", "label_paper"))


# Unjournal ratings
rsx <- read_csv(here("data", "rsx_evalr_rating.csv"), show_col_types = FALSE) |> 
  clean_names()  |>
  mutate(label_paper_title = research) |>
  select(-c("research"))


# UJ evaluated research
research <- read_csv(here("data", "research.csv"), show_col_types = FALSE) |>
  clean_names() |>
  filter(status == "50_published evaluations (on PubPub, by Unjournal)") |>  
  left_join(UJmap, by = c("label_paper_title")) |>
  mutate(doi = str_trim(doi)) |>
  mutate(label_paper = if_else(doi == "https://doi.org/10.3386/w31162", "Walker et al. 2023", label_paper, missing = label_paper)) |>
  mutate(label_paper = if_else(doi == "doi.org/10.3386/w32728", "Hahn et al. 2025", label_paper, missing = label_paper))  |>
  mutate(label_paper = if_else(doi == "https://doi.org/10.3386/w30011", "Bhat et al. 2022", label_paper, missing = label_paper))  |>
  mutate(label_paper = if_else(doi == "10.1093/wbro/lkae010", "Crawfurd et al. 2023", label_paper, missing = label_paper))  |>
  left_join(rsx, by = c("label_paper_title"))
 

jtiers_llm <- read_csv(here("data", "journal_tiers_long.csv"), show_col_types = FALSE) |>
  mutate(middle_rating = score,
         lower_ci = ci_lower,
         upper_ci = ci_upper,
         criteria = if_else(tier_kind == "will", "journal_predict", "merits_journal"),
         evaluator = model,
         label_paper = paper
         ) |>
  select(c("label_paper", "evaluator", "middle_rating", "lower_ci", "upper_ci" , "criteria", "rationale"))


jtiers_uj <- research |>
  filter(criteria== "merits_journal" | criteria == "journal_predict") |>
  mutate(paper = label_paper,
         rationale = "")  |>
  select(c("label_paper", "evaluator", "middle_rating", "lower_ci", "upper_ci" , "criteria", "rationale"))


jtiers <- jtiers_uj |>
  rbind(jtiers_llm) |>
  mutate(human = if_else(evaluator == "o3", "Human", "o3"),
         lower_ci = if_else(lower_ci > 10, lower_ci/10, lower_ci))


# write_csv(all_ratings, here("data", "all_jtiers.csv"))
write_rds(
  jtiers,
  here("data", "all_jtiers.rds"),
  compress = "none"
  )


```



```{r}
#| label: fig-forest-tiers
#| fig-cap: "Journal tiers -- LLM vs individual Human ratings. Each rating on its own lane: orange LLM midpoint+CI centered; green human midpoints (CI when available) offset per rater."
#| fig-height: 13
#| fig-width: 8


jtiers <- jtiers |>
  mutate(
    who = if_else(evaluator == "o3", "LLM", "Human"),
    mid = middle_rating, lo = lower_ci, hi = upper_ci
  )  

# order rows within each facet: by LLM mid if present, else median human
ord_tbl <- jtiers %>%
  group_by(criteria, label_paper) %>%
  summarise(
    ord = if (any(who == "LLM")) mid[which(who == "LLM")[1]] else median(mid[who == "Human"], na.rm = TRUE),
    n_h = sum(who == "Human"),
    .groups = "drop"
  ) %>%
  group_by(criteria) %>% arrange(desc(ord), .by_group = TRUE) %>%
  mutate(level = paste(criteria, paste0(label_paper, " (n=", n_h, ")"), sep = "___"))

jplot <- jtiers %>%
  left_join(ord_tbl, by = c("criteria", "label_paper")) %>%
  mutate(level = paste(criteria, paste0(label_paper, " (n=", n_h, ")"), sep = "___"),
         paper_fac = factor(level, levels = ord_tbl$level))

lab_fun <- function(x) sub("^.*___", "", x)

set.seed(1)
ggplot(jplot, aes(x = mid, y = paper_fac, colour = who, shape = who)) +
  # humans: jittered lanes
  geom_errorbarh(
    data = subset(jplot, who == "Human" & is.finite(lo) & is.finite(hi)),
    aes(xmin = lo, xmax = hi),
    height = 0, alpha = 0.55, linewidth = 0.5,
    position = position_jitter(height = 0.22, width = 0)
  ) +
  geom_point(
    data = subset(jplot, who == "Human"),
    size = 1.9, alpha = 0.9,
    position = position_jitter(height = 0.22, width = 0)
  ) +
  # LLM: centered lane
  geom_errorbarh(
    data = subset(jplot, who == "LLM" & is.finite(lo) & is.finite(hi)),
    aes(xmin = lo, xmax = hi),
    height = 0, linewidth = 0.8
  ) +
  geom_point(
    data = subset(jplot, who == "LLM"),
    size = 2.4
  ) +
  facet_wrap(~criteria, ncol = 1, scales = "free_y",
             labeller = as_labeller(c(journal_predict = "Where will this paper be published?", merits_journal = "Where should this paper be published?"))) +
  scale_color_manual(values = c(LLM = UJ_ORANGE, Human = UJ_GREEN), name = NULL) +
  scale_shape_manual(values = c(LLM = 18, Human = 16), name = NULL) +
  scale_y_discrete(labels = lab_fun, expand = expansion(mult = c(0.02, 0.06))) +
  coord_cartesian(xlim = c(0, 5)) +
  labs(x = NULL, y = NULL) +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  theme_uj() +
  theme(panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.3),
        axis.text.y = element_text(hjust = 0.98),
        legend.position = "bottom")
```


### Journal tier correlations with quality metrics

Similar to [the analysis for human evaluations](https://unjournal.github.io/unjournaldata/posts/uj-data-update-2025/#fig-journal-quality-cors), we examine how both LLM and Human "where should this paper publish" ratings correlate with their respective other quality metrics. This reveals what dimensions each evaluator weighs most heavily when assessing journal placement.

```{r}
#| label: fig-tier-cors-comparison
#| fig-cap: "Comparison of journal tier correlations: LLM vs Human evaluators. Green dots show Human correlations, orange dots show LLM correlations. Lines connect the same metric."
#| fig-height: 6
#| fig-width: 9






# === LLM Correlations ===
llm_cors_data <- llm_raw |>
  select(label_paper, criteria, midpoint_llm) |>
  pivot_wider(names_from = criteria, values_from = midpoint_llm)

tier_should_llm <- jtiers_llm |>
  filter(criteria == "merits_journal") |>
  select(label_paper, tier_should = middle_rating)

llm_cors_data <- llm_cors_data |>
  left_join(tier_should_llm, by = "label_paper") |>
  filter(!is.na(tier_should))

metrics_to_cor <- c("overall", "claims", "methods", "adv_knowledge",
                    "logic_comms", "open_sci", "gp_relevance")

cors_llm <- tibble(
  metric = metrics_to_cor,
  correlation = map_dbl(metrics_to_cor, function(m) {
    if (m %in% names(llm_cors_data)) {
      cor(llm_cors_data[[m]], llm_cors_data$tier_should,
          use = "pairwise.complete.obs")
    } else {
      NA_real_
    }
  }),
  source = "LLM"
) |>
  filter(!is.na(correlation))

# === Human Correlations ===
human_cors_data <- human_raw |>
  select(label_paper, criteria, middle_rating) |>
  group_by(label_paper, criteria) |>
  summarise(midpoint_human = mean(middle_rating, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = criteria, values_from = midpoint_human)

tier_should_human <- jtiers_uj |>
  filter(criteria == "merits_journal") |>
  group_by(label_paper) |>
  summarise(tier_should = mean(middle_rating, na.rm = TRUE), .groups = "drop")

human_cors_data <- human_cors_data |>
  left_join(tier_should_human, by = "label_paper") |>
  filter(!is.na(tier_should))

cors_human <- tibble(
  metric = metrics_to_cor,
  correlation = map_dbl(metrics_to_cor, function(m) {
    if (m %in% names(human_cors_data)) {
      cor(human_cors_data[[m]], human_cors_data$tier_should,
          use = "pairwise.complete.obs")
    } else {
      NA_real_
    }
  }),
  source = "Human"
) |>
  filter(!is.na(correlation))

# === Combine and create dumbbell plot ===
cors_combined <- bind_rows(cors_llm, cors_human) |>
  mutate(
    metric_label = case_when(
      metric == "overall" ~ "Overall",
      metric == "claims" ~ "Claims & Evidence",
      metric == "methods" ~ "Methods",
      metric == "adv_knowledge" ~ "Advancing Knowledge",
      metric == "logic_comms" ~ "Logic & Communication",
      metric == "open_sci" ~ "Open Science",
      metric == "gp_relevance" ~ "Global Relevance",
      TRUE ~ metric
    )
  )

cors_wide <- cors_combined |>
  pivot_wider(names_from = source, values_from = correlation) |>
  mutate(
    avg_cor = (Human + LLM) / 2,
    metric_label = fct_reorder(metric_label, avg_cor)
  )

# Dumbbell plot
ggplot(cors_wide, aes(y = metric_label)) +
  geom_segment(aes(x = Human, xend = LLM, yend = metric_label),
               color = "gray50", linewidth = 1.2, alpha = 0.4) +
  geom_point(aes(x = Human), color = UJ_GREEN, size = 5, alpha = 0.9) +
  geom_point(aes(x = LLM), color = UJ_ORANGE, size = 5, alpha = 0.9) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40", alpha = 0.5) +
  geom_text(aes(x = Human, label = sprintf("%.2f", Human)),
            hjust = 1.4, size = 3.5, color = UJ_GREEN, fontface = "bold") +
  geom_text(aes(x = LLM, label = sprintf("%.2f", LLM)),
            hjust = -0.4, size = 3.5, color = UJ_ORANGE, fontface = "bold") +
  scale_x_continuous(limits = c(min(c(cors_wide$Human, cors_wide$LLM)) - 0.2,
                                 max(c(cors_wide$Human, cors_wide$LLM)) + 0.2)) +
  labs(
    x = "Correlation with 'Where should this publish?'",
    y = NULL,
    title = "How quality metrics predict journal tier: LLM vs Human evaluators",
    subtitle = paste0("Green = Human evaluators | Orange = ", currentmodel, " | Higher values = stronger weight in tier prediction")
  ) +
  theme_uj() +
  theme(
    panel.grid.major.x = element_line(color = "grey90"),
    panel.grid.major.y = element_blank(),
    plot.title.position = "plot"
  )
```

This visualization reveals interesting differences in how LLM and human evaluators weight different quality dimensions when predicting journal placement. For instance, if the LLM's orange dot is much higher than the human's green dot for a particular metric, it suggests the LLM relies more heavily on that dimension when making its "should publish" assessment.

## Quantitative ratings

We evaluated `r tok_sum$n_papers` papers using `r currentmodel`, reading each manuscript directly from the PDF and returning a strict JSON assessment. 

<!-- Before looking at the output, it is instructive to look at the reported reasoning summary of the LLM (only partially available).   -->
<!-- For example, this is what `r currentmodel` reports about its reasoning for @Williams2024: -->

<!-- r sprintf(reasoning_example$reasoning_summary) -->



### Overall ratings

We begin by comparing *overall* scores. 


```{r}
#| label: load-data-x

# Mapping paper keys - short titles
UJmap <- read_delim(here("data", "UJ_map.csv"), delim = ";") |>
  mutate(label_paper_title = research,
         label_paper = paper) |>
  select(c("label_paper_title", "label_paper"))



# LLM generated ratings
metrics <- read_csv(here("data", "metrics_long.csv"), show_col_types = FALSE)

metrics <- metrics |> clean_names() |>
  mutate(evaluator = currentmodel,
         label_paper = str_replace(paper, "et al ", "et al. "),
         middle_rating = midpoint,
         lower_ci = lower_bound,
         upper_ci = upper_bound,
         criteria = metric) |>
  mutate(criteria = if_else(criteria == "advancing_knowledge", "adv_knowledge", criteria)) |>
  mutate(criteria = if_else(criteria == "open_science", "open_sci", criteria)) |>
  mutate(criteria = if_else(criteria == "logic_communication", "logic_comms", criteria)) |>
  mutate(criteria = if_else(criteria == "global_relevance", "gp_relevance", criteria)) |>
  mutate(criteria = if_else(criteria == "claims_evidence", "claims", criteria)) |>
  # mutate(criteria = factor(criteria)) |>
  left_join(UJmap, by = c("label_paper")) |>
  select(c("label_paper", "label_paper_title", "evaluator", "criteria", "middle_rating", "lower_ci", "upper_ci", "rationale"))
  
  
# Unjournal ratings
rsx <- read_csv(here("data", "rsx_evalr_rating.csv"), show_col_types = FALSE) |> 
  clean_names()  |>
  mutate(label_paper_title = research) |>
  select(-c("research"))

# More on evaluated research
research <- read_csv(here("data", "research.csv"), show_col_types = FALSE) |>
  clean_names() |>
  filter(status == "50_published evaluations (on PubPub, by Unjournal)")

# rsx_collapsed = rsx |>
#   group_by(label_paper_title) |>
#   summarise(g_middle_rating = mean(middle_rating, na.rm = TRUE)) |>
#   left_join(UJmap, by = c("label_paper_title")) 

# More journal rating info
# jql70_raw <- read_csv(here("data", "jql70a.csv"), show_col_types = FALSE)
jql_enriched_raw <- read_csv(here("data", "jql-enriched.csv"), show_col_types = FALSE)


# Merge Unjournal data and keys
rsx_research <- rsx  |>
  left_join(research, by = c("label_paper_title")) |>  
  left_join(UJmap, by = c("label_paper_title")) |>
  select(c("label_paper", "label_paper_title", "evaluator","criteria","middle_rating","lower_ci","upper_ci")) |>
  mutate(rationale = "") |>
  mutate(
    label_paper = if_else(label_paper_title == "A Welfare Analysis of Policies Impacting Climate Change", "Hahn et al. 2025", label_paper),
    label_paper = if_else(label_paper_title == "Intergenerational Child Mortality Impacts of Deworming: Experimental Evidence from Two Decades of the Kenya Life Panel Survey
", "Walker et al. 2023", label_paper)
  )
 
# Merge Unjournal data and LLM metrics
all_ratings <- rbind(rsx_research, metrics) |>
  mutate(criteria = factor(criteria)) |>
  mutate(evaluator = factor(evaluator)) |>
  mutate(label_paper_title = factor(label_paper_title)) |>
  mutate(label_paper = factor(label_paper))

# clean up
# rm("metrics", "research", "rsx", "rsx_research", "UJmap")

# write_csv(all_ratings, here("data", "all_ratings.csv"))
write_rds(
  all_ratings,
  here("data", "all_ratings.rds"),
  compress = "none"
  )


# Ensure numeric + fix any swapped CI bounds
all_ratings <- all_ratings |>
  mutate(across(c(middle_rating, lower_ci, upper_ci), as.numeric)) |>
  fix_bounds("lower_ci","upper_ci")

# Split: LLM vs Human raters (LLM = evaluator == 'GPT-5')
human_raw <- all_ratings |>
  filter(evaluator != currentmodel) |>
  filter(!is.na(middle_rating))

llm_raw <- all_ratings |>
  filter(evaluator == currentmodel) |>
  transmute(
    label_paper, label_paper_title, criteria,
    midpoint_llm = middle_rating,
    lower_llm = lower_ci,
    upper_llm = upper_ci
  ) |>
  distinct()
 



human_use <- human_raw |>
  group_by(label_paper, label_paper_title, criteria) |>
  summarise(
    midpoint_human   = mean(middle_rating, na.rm = TRUE),
    # keep means of human CIs (used in one of the outlier tables)
    lower_human      = mean(lower_ci, na.rm = TRUE),
    upper_human      = mean(upper_ci, na.rm = TRUE),
    # CI union (fallback to min/max of points where CI missing)
    human_lo_union   = safe_min(lower_ci, middle_rating),
    human_hi_union   = safe_max(upper_ci, middle_rating),
    n_raters         = dplyr::n(),
    .groups = "drop"
  ) |>
  fix_bounds("human_lo_union","human_hi_union")

# Merge to LLM
merged <- llm_raw |>
  inner_join(human_use, by = c("label_paper","label_paper_title","criteria")) |>
  mutate(diff = midpoint_llm - midpoint_human)

# Long form for distributions
ratings_long <- merged |>
  select(label_paper, label_paper_title, criteria, midpoint_llm, midpoint_human) |>
  pivot_longer(starts_with("midpoint_"),
               names_to = "rater", values_to = "score") |>
  mutate(rater = recode(rater,
                        midpoint_llm   = "LLM",
                        midpoint_human = "Human"))

```

In @fig-forest, the orange diamond and line show the `r print(currentmodel)` model's midpoint and its 90% credible interval for each paper, while the green circles show each individual human evaluator’s midpoint (along with 90% CI when they provided these).

Looking at this figure, we can visually assess agreement. If the orange diamond lies among the green circles and lines for a paper, the AI’s overall assessment is in line with the human range. If the orange diamond is far to the right (higher score) or left (lower score) relative to the green points, it indicates a notable disagreement.



```{r}
#| label: fig-forest
#| fig-cap: "LLM vs individual Human ratings (overall). Each rating on its own lane: orange LLM midpoint+CI centered; green human midpoints (thin CI when available) offset per rater."
#| fig-height: 9
#| fig-width: 8

metric_name <- "overall"

# Helper: symmetric lane offsets (skip 0 so center lane is reserved for LLM)
lane_offsets <- function(m, gap = 0.18) { #DR @valentin -- we probably want a bigger gap between papers here
  if (m <= 0) return(numeric(0))
  k <- ceiling(m/2)
  cand <- c(-seq_len(k), seq_len(k)) * gap
  sort(cand)[seq_len(m)]
}

# Base ordering of papers (same logic as before) #DR @valentin  OK, I guess it's ordered by 'average human minus AI rating'? But we should then note this in the doc. Might also make the diagram dynamic if possible to let users sort in other ways, like  by human average rating, AI average rating, or alphabetically. 

D_base <- merged |>
  filter(criteria == metric_name) |>
  mutate(paper_order = fct_reorder(label_paper, diff, .desc = FALSE)) |>
  distinct(label_paper, label_paper_title, paper_order, n_raters) #DR @valentin -- maybe we don't need to put the number of raters here as the lines show that. At least as long as there's enough space to see that part of the diagram. 

# Pretty label with # of human raters
D_base <- D_base |>
  mutate(paper_lab = paste0(label_paper, " (n=", n_raters, ")")) |>
  arrange(paper_order) |>
  mutate(pos = row_number())

pos_map <- D_base |>
  select(label_paper, paper_lab, pos)

# --- Individual human ratings (each gets its own offset lane) --------------
H_indiv <- human_raw |>
  filter(criteria == metric_name) |>
  inner_join(pos_map, by = "label_paper") |>
  group_by(label_paper) |>
  arrange(evaluator, .by_group = TRUE) |>
  mutate(
    h_id = row_number(),
    n_h  = dplyr::n(),
    off  = purrr::map2_dbl(n_h, h_id, ~ lane_offsets(.x)[.y]),  # <-- scalarize here
    y    = pos + off
  ) |>
  ungroup() |>
  transmute(
    paper_lab,
    y,
    who = "Human",
    mid = as.numeric(middle_rating),
    lo  = suppressWarnings(as.numeric(lower_ci)),
    hi  = suppressWarnings(as.numeric(upper_ci))
  )

# --- LLM midpoint + CI (one centered lane per paper) -----------------------
L_llm <- llm_raw |>
  filter(criteria == metric_name) |>
  inner_join(pos_map, by = "label_paper") |>
  mutate(y = pos) |>
  transmute(
    paper_lab,
    y,
    who = "LLM",
    mid = as.numeric(midpoint_llm),
    lo  = as.numeric(lower_llm),
    hi  = as.numeric(upper_llm)
  )

Pts <- bind_rows(H_indiv, L_llm)

# Axis limits from all intervals
rng <- range(c(Pts$mid, Pts$lo, Pts$hi), na.rm = TRUE)
pad <- 0.04 * diff(rng)
xlim_use <- c(rng[1] - pad, rng[2] + pad)

# Light separators per paper row
row_lines <- pos_map$pos

ggplot() +
  geom_hline(yintercept = row_lines, color = "grey92", linewidth = 0.3) +
  # Human CIs (thin) + points on their own lanes
  geom_errorbarh(
    data = H_indiv |> filter(is.finite(lo), is.finite(hi)),
    aes(y = y, xmin = lo, xmax = hi, colour = who),
    height = 0, alpha = 0.5, linewidth = 0.5
  ) +
  geom_point(
    data = H_indiv,
    aes(x = mid, y = y, colour = who, shape = who),
    size = 2.1, alpha = 0.9
  ) +
  # LLM CI (thicker) + diamond at center lane
  geom_errorbarh(
    data = L_llm,
    aes(y = y, xmin = lo, xmax = hi, colour = who),
    height = 0, linewidth = 0.7
  ) +
  geom_point(
    data = L_llm,
    aes(x = mid, y = y, colour = who, shape = who),
    size = 2.6
  ) +
  scale_color_manual(values = c(LLM = UJ_ORANGE, Human = UJ_GREEN), name = NULL) +
  scale_shape_manual(values = c(LLM = 18, Human = 16), name = NULL) +
  scale_y_reverse(
    breaks = pos_map$pos,
    labels = pos_map$paper_lab,
    expand = expansion(mult = c(0.02, 0.06))
  ) +
  coord_cartesian(xlim = xlim_use) +
  labs(x = NULL, y = NULL
  ) +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  theme_uj() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(hjust = 0.98)
  )


```


```{r}
#| label: forest-improved
#| fig-height: 14
#| fig-cap: "Journal tiers -- LLM vs individual Human ratings. Each rating on its own lane: orange LLM midpoint+CI centered; green human midpoints (CI when available) offset per rater."

p <- ggplot() +
  # Background grid lines for each paper row
  geom_hline(yintercept = pos_map$pos, color = "grey90", linewidth = 0.4) +
  
  # Human intervals: dashed lines for distinction
  geom_errorbarh(
    data = H_indiv |> filter(is.finite(lo), is.finite(hi)),
    aes(y = y, xmin = lo, xmax = hi, colour = who),
    height = 0, alpha = 0.5, linewidth = 1.0, linetype = "dashed"
  ) +
  
  # Human points: larger, semi-transparent for visibility
  geom_point(
    data = H_indiv,
    aes(x = mid, y = y, colour = who, shape = who),
    size = 2.8, alpha = 0.9
  ) +
  
  # LLM intervals: thicker solid lines
  geom_errorbarh(
    data = L_llm,
    aes(y = y, xmin = lo, xmax = hi, colour = who),
    height = 0, linewidth = 1.2
  ) +
  
  # LLM points: diamond shape, largest
  geom_point(
    data = L_llm,
    aes(x = mid, y = y, colour = who, shape = who),
    size = 3.4, stroke = 1
  ) +
  
 scale_color_manual(
    values = c(LLM = UJ_ORANGE, Human = "#669933"),
    name = NULL,
    labels = c(LLM = "LLM", Human = "Human")
  ) +
  scale_shape_manual(
    values = c(LLM = 18, Human = 16),
    name = NULL,
    labels = c(LLM = "LLM", Human = "Human")
  ) +
  
  scale_y_reverse(
    breaks = pos_map$pos,
    labels = pos_map$paper_lab,
    expand = expansion(mult = c(0.02, 0.06))
  ) +
  coord_cartesian(xlim = xlim_use) +
  labs(x = NULL, y = NULL) +
  
  guides(
    colour = guide_legend(override.aes = list(alpha = 1, size = 3.4)),
    shape  = guide_legend(override.aes = list(size = 3.4))
  ) +
  
  theme_uj() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(hjust = 0.98, size = 12),
    legend.position = "bottom",
    plot.margin = margin(10, 30, 10, 30)
  ) 


# Optional: For interactive exploration in HTML output
 library(plotly)
 ggplotly(p)

```

Next, @fig-rankslope contrasts relative *ranks* of papers under human and LLM scoring. This visualization directly compares how the AI and the human reviewers rank each paper in terms of overall quality. Each paper is represented by a curve connecting two ranked lists: on the left, the papers are ordered top-to-bottom by the human overall score (rank 1 = highest rated by humans); on the right, the papers are ordered by the AI’s overall score (rank 1 = highest rated by AI). A paper that occupies the same rank in both lists would appear as a straight horizontal line. If the AI ranks a paper higher than the humans did, the line for that paper will slope upward from left to right (starting lower on the left and ending higher on the right). Those lines are drawn in orange, indicating “AI higher than human.” Conversely, if the AI ranks a paper lower, the line slopes downward (green line indicating “AI lower than human”).

 
```{r}
#| label: fig-rankslope
#| fig-cap: "Relative ranking (overall) by LLM and Human evaluators"
#| fig-height: 8
#| fig-width: 8


# Replace helper: make_s_bezier (now carries numeric Δrank info)
make_s_bezier <- function(D, dx_base = 0.33, dx_min = 0.08) {
  dy <- abs(D$pos_right - D$pos_left)
  k  <- max(2, stats::quantile(dy, 0.75, na.rm = TRUE))
  dx_i <- pmax(dx_min, dx_base * exp(-dy / k))

  purrr::map_dfr(seq_len(nrow(D)), function(i) {
    dr_i <- D$pos_left[i] - D$pos_right[i]   # Δrank = Human − LLM
    tibble::tibble(
      group = i,
      x = c(0, dx_i[i], 1 - dx_i[i], 1),
      y = c(D$pos_left[i], D$pos_left[i], D$pos_right[i], D$pos_right[i]),
      dr  = dr_i,                  # signed Δrank
      mag = abs(dr_i)              # |Δrank|
    )
  })
}


# --- replace build_rank_positions ------------------------------------------
build_rank_positions <- function(metric) {
  D <- merged |>
    filter(criteria == metric) |>
    # keep BOTH labels; create a display label preferring the short code
    mutate(
      label_use = dplyr::if_else(
        !is.na(label_paper) & nzchar(as.character(label_paper)),
        as.character(label_paper),
        as.character(label_paper_title)
      )
    ) |>
    select(label_use, label_paper, label_paper_title, midpoint_human, midpoint_llm)

  if (!nrow(D)) return(NULL)

  left  <- D |> arrange(desc(midpoint_human), label_use) |> mutate(pos_left  = row_number())
  right <- D |> arrange(desc(midpoint_llm),   label_use) |> mutate(pos_right = row_number())

  D |> 
    left_join(left  |> select(label_use, pos_left),  by = "label_use") |>
    left_join(right |> select(label_use, pos_right), by = "label_use") |>
    mutate(delta = midpoint_llm - midpoint_human)
}

plot_rank_slope_S <- function(metric,
                              D = NULL,
                              right_width = 0.28,
                              color_mode = c("gradient", "steps3"),
                              soft_thresh = 5) {
  color_mode <- match.arg(color_mode)
  if (is.null(D)) D <- build_rank_positions(metric)
  if (is.null(D) || !nrow(D)) return(ggplot() + theme_void())

  # Pretty metric for legend title
  lab_metric   <- stringr::str_to_title(gsub("_", " ", metric))
  legend_title <- paste0("\u0394rank (H \u2212 LLM) — ", lab_metric)

  # Ranks + right-side label
  D <- D |>
    mutate(
      rank_h   = pos_left,                  # Human rank (1 = highest)
      rank_l   = pos_right,                 # LLM   rank (1 = highest)
      d_rank   = rank_h - rank_l,           # Δrank = Human − LLM
      right_lab = sprintf("LLM #%d | H #%d | \u0394r=%+d", rank_l, rank_h, d_rank)
    )

  B <- make_s_bezier(D) |>
    mutate(alpha_by = pmin(mag / soft_thresh, 1))  # fade small |Δr|

  p <- ggplot()

  if (color_mode == "gradient") {
    p <- p +
      ggforce::geom_bezier(
        data = B,
        aes(x = x, y = y, group = group, colour = dr, alpha = alpha_by),
        size = 0.9
      ) +
      scale_color_gradient2(
        low = UJ_GREEN, mid = "grey90", high = UJ_ORANGE, midpoint = 0,
        name = legend_title
      ) +
      guides(colour = guide_colourbar(title.position = "top")) +
      scale_alpha(range = c(0.45, 1), guide = "none")
  } else { # "steps3": three colors, small diffs de-emphasized
    B <- B |>
      mutate(col3 = dplyr::case_when(
        mag <= soft_thresh ~ "Small (≤5)",
        dr > 0             ~ "LLM higher",
        TRUE               ~ "LLM lower"
      ))
    p <- p +
      ggforce::geom_bezier(
        data = B,
        aes(x = x, y = y, group = group, colour = col3),
        size = 0.9, alpha = 0.9
      ) +
      scale_color_manual(
        values = c("LLM higher" = UJ_ORANGE,
                   "LLM lower"  = UJ_GREEN,
                   "Small (≤5)" = "grey80"),
        name = legend_title
      ) +
      guides(colour = guide_legend(title.position = "top"))
  }

  p +
    geom_point(data = D, aes(x = 0, y = pos_left),
               color = UJ_GREEN, size = 2.2) +
    geom_point(data = D, aes(x = 1, y = pos_right),
               color = UJ_ORANGE, shape = 18, size = 2.4) +
    geom_text(data = D, aes(x = 0, y = pos_left, label = label_use),
              hjust = 1.05, size = 3, color = "grey20") +
    geom_text(data = D, aes(x = 1, y = pos_right, label = right_lab),
              hjust = -0.05, size = 3, color = "grey20") +
    scale_x_continuous(limits = c(-0.7, 1 + right_width),
                       breaks = c(0, 1), labels = c("Human", "LLM"),
                       expand = expansion(mult = 0)) +
    scale_y_reverse(expand = expansion(mult = c(0.02, 0.06))) +
    coord_cartesian(clip = "off") +
    labs(x = NULL, y = NULL) +          # no title; caption handles description
    theme_uj() +
    theme(
      axis.text.y  = element_blank(),   # hide y numbers
      axis.ticks.y = element_blank(),
      panel.grid.major.y = element_blank(),   # remove horizontal grid lines
      panel.grid.minor.y = element_blank(),   # (defensive)
      plot.margin = margin(t = 10, r = 190, b = 10, l = 140)
    )
}


plot_rank_slope_S(metric_name)



```

In our results, we see a mix: many papers lie fairly close to horizontal (especially in the middle of the pack), but there are several with pronounced slopes. For example, a few orange lines curve sharply upward – these are papers that the AI considered to be among the top-ranked, while humans had them in the middle or lower end. Papers with steep green downward curves are ones humans rated highly but the AI was less impressed by. One can identify specific papers by these lines; for instance, one orange line corresponds to @Williams2024, which the AI ranked much higher than the median human rank. On the other hand, we see a green line for a paper that humans ranked very highly but AI did not – for example, Aghion et al. 2017 was among the top few for human evaluators, but the LLM overall score put it notably lower relative to others, hence a downward green curve.

The overall pattern suggests that the AI and humans broadly agree on some top performers, but there are notable swaps in positions. It’s not the case that the AI simply gave everything high scores or low scores uniformly – it may have a distinct “taste,” elevating some work and devaluing other work differently than human referees. The question this raises is: on what basis are these differences happening? To explore that, we can look at the category-by-category differences next, followed by multidimensional modeling. However, we caution that the various observed characteristics of each paper are not determined independently from each other, and may also systematically relate to unobservable characteristics. Thus, unless we introduce exogenous variation (e.g., by altering the text of the papers shared, as in @Pataranutaporn2025), these results will be suggestive and exploratory, and not clearly causal or predictive. 

### Individual metrics

We then look closer at the specific ratings on the criterion level. @fig-heat displays LLM − Human differences by paper × metric. Each row is a paper (the rows here are ordered by the difference in overall score, with papers at the top being those the AI scored lower than humans overall, and at the bottom those the AI scored higher). Each column is one of the criteria: Claims & Evidence, Methods, Advancing Knowledge, Logic & Communication, Open Science, Global Relevance, and Overall. The cell color shows the AI score minus the human score for that paper on that criterion. Orange shades mean the AI’s score is higher; green shades mean the AI’s score is lower. White or light gray indicates the AI and human scores were about the same.

```{r}
#| label: fig-heat
#| fig-height: 7
#| fig-cap: "LLM − Human differences by paper × metric; sorted by 'greatest LLM relative preference'"

pair <- merged |>
  transmute(paper = label_paper, metric = criteria,
            diff = midpoint_llm - midpoint_human)

# Row order by signed difference on 'overall' (if present)
order_overall <- pair |>
  filter(metric == "overall") |>
  group_by(paper) |>
  summarise(d = mean(diff, na.rm = TRUE), .groups = "drop") |>
  arrange(d) |>
  pull(paper)

pair$paper <- factor(pair$paper, levels = unique(c(order_overall, pair$paper)))


ggplot(pair, aes(x = metric, y = paper, fill = diff)) +
  geom_tile(color = "white", size = 0.4) +
  scale_fill_gradient2(low = UJ_GREEN, mid = "grey95", high = UJ_ORANGE,
                       midpoint = 0,
                       name = "LLM − Human") +
  labs(x = NULL, y = NULL) +
  theme_uj() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))

```



This figure provides a more granular view of where AI and human evaluations diverge. A quick scan reveals a few systematic tendencies:
 
For some criteria,  the AI tended to score papers higher than humans.  For example, in the Logic & Communication column, we see many light-orange cells – the AI often thought papers were a bit clearer or better argued (by its judgment) than the human evaluators did.

In contrast, the Open Science column shows a notable amount of green. Here the AI frequently gave lower scores than the humans. This suggests that the AI was harsher about transparency/reproducibility issues. Indeed, GPT often noted lack of code or data sharing in papers and penalized for it, whereas some human reviewers may have been more forgiving or did not emphasize open-science practices as strongly, or may have had lower expectations, especially for pre-journal-publication work. As a result, for many papers the AI’s Open Science score is 5–10 points below the human average.

@tbl-agreement show agreement metrics across rating criteria. To quantify the agreements and differences observed, we calculated several statistics comparing LLM scores to the human scores, aggregated by criterion:

Correlation (Pearson’s r) between the AI’s and human scores across papers: This tells us, for example, if a paper that humans gave a high score also tended to get a high score from AI (regardless of absolute difference). For the Overall scores, Pearson r ≈ 0.30, indicating a weak-to-moderate positive correlation. In other words, there is some alignment – the AI does often rate the generally “better” papers higher – but there is considerable noise. Some criteria showed essentially no linear correlation (for Logic & Communication, r ~ 0, meaning the AI’s clarity ratings had no linear relationship with human ratings).

We also computed Spearman rank correlation, where for Overall it was a bit higher (around 0.40), suggesting the AI is moderately good at ranking papers in roughly the same order as humans even if the exact scores differ. Notably, the Methods criterion had one of the higher Spearman correlations (~0.43), implying AI and humans somewhat agree on which papers have stronger methods, whereas Global Relevance and Communication had very low or even slight negative correlation, implying essentially no agreement on those fronts.

Inter-rater reliability (Cohen’s κ): We treated the AI as one “reviewer” and the average human as another, and asked: how often do they effectively give similar ratings, beyond what chance alignment would predict? For overall scores, Cohen’s κ (with quadratic weighting for partial agreement) came out around 0.25–0.30. This would be characterized as “fair” agreement at best – a significant gap remains between AI and human judgments. For most specific criteria, the weighted κ was lower, often in the 0.1–0.2 range (and for one category, effectively 0). A κ of 0 would mean no more agreement than random chance (given the distribution of scores), so some categories are bordering on that. In contrast, typical inter-human agreement on these kinds of scoring tasks can also be low, but usually one would hope for κ in the 0.3–0.6 range among trained reviewers on well-defined criteria. Our finding of κ < 0.3 in all cases suggests that the AI’s ratings are not interchangeable with a human reviewer’s – at least not without further calibration.

```{r}
#| label: tbl-agreement
#| tbl-cap: "Overall agreement across all metrics"

# Replace: agreement_summary() and display
agreement_summary <- function(df, n_bins = 5, bin_strategy = "quantile") {
  df |>
    group_by(criteria) |>
    summarise(
      n             = sum(is.finite(midpoint_llm) & is.finite(midpoint_human)),
      pearson       = suppressWarnings(cor(midpoint_llm, midpoint_human, use = "pairwise.complete.obs", method = "pearson")),
      spearman      = suppressWarnings(cor(midpoint_llm, midpoint_human, use = "pairwise.complete.obs", method = "spearman")),
      # mean_bias     = mean(midpoint_llm - midpoint_human, na.rm = TRUE),
      # median_bias   = median(midpoint_llm - midpoint_human, na.rm = TRUE),
      # MAE           = mean(abs(midpoint_llm - midpoint_human), na.rm = TRUE),
      # MedAE         = median(abs(midpoint_llm - midpoint_human), na.rm = TRUE),
      # RMSE          = sqrt(mean((midpoint_llm - midpoint_human)^2, na.rm = TRUE)),
      kappa_unwtd   = { b <- bin_together(midpoint_llm, midpoint_human, n_bins, bin_strategy); weighted_kappa(b$a_bin, b$b_bin, k = b$k, weights = "unweighted") },
      kappa_quad    = { b <- bin_together(midpoint_llm, midpoint_human, n_bins, bin_strategy); weighted_kappa(b$a_bin, b$b_bin, k = b$k, weights = "quadratic") },
      .groups = "drop"
    ) |>
    arrange(criteria)
}

agree_view <- agreement_summary(merged, n_bins = 5, bin_strategy = "quantile") |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))

kable(agree_view)



```

### Rationale behind the largest differences

As a first shot at understanding sources of disagreement, we look at the reported rationale for those ratings that diverge the most from (average) human assessments. 


```{r}
#| label: rationale-outliers

rationale_outlier = all_ratings |>
  filter(evaluator == currentmodel, 
         label_paper == "Williams et al. 2024",
         criteria == "overall")

```


For @Williams2024 "overall", `r currentmodel` gives the following rationale: `r sprintf(rationale_outlier$rationale)`


## Statistical analyses of agreement (will move/integrate with Q&A)





Krippendorff’s alpha

## Claim identification


## Qualitative assessments









