---
title: "Results: Ratings"
engine: knitr
---

```{r}
#| label: setup
#| code-summary: "Setup and libraries"
#| code-fold: true
#| message: false
#| warning: false

source("setup_params.R")

library("tidyverse")
library("janitor")
library("stringr")
library("here")
library("knitr")
library("kableExtra")
library("ggrepel")
library("scales")
library("jsonlite")
library("purrr")
library("tibble")
# library("plotly")  # removed: all figures are now static ggplot

# Theme and colors — crisp palette (high saturation, maximum separation)
UJ_ORANGE <- "#E8722A"   # vivid saffron orange
UJ_GREEN  <- "#2D9D5E"   # rich emerald green
UJ_BLUE   <- "#2B7CE9"   # clear azure blue

MODEL_COLORS <- c(
  "GPT-5 Pro"        = "#E8722A",   # saffron orange (focal)
  "GPT-5.2 Pro"      = "#D62839",   # bright crimson
  "GPT-4o-mini"      = "#17B890",   # vivid teal
  "Claude Sonnet 4"  = "#A855F7",   # vivid purple
  "Claude Opus 4.6"  = "#7C3AED",   # deep violet
  "Gemini 2.0 Flash" = "#2B7CE9",   # clear azure
  "Human"            = "#2D9D5E"    # emerald green
)

theme_uj <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor   = element_blank(),
      panel.grid.major   = element_line(linewidth = 0.3, color = "grey88"),
      plot.title          = element_text(face = "bold", size = rel(1.1)),
      plot.title.position = "plot",
      plot.subtitle       = element_text(color = "grey40", size = rel(0.9)),
      plot.caption        = element_text(color = "grey50", size = rel(0.8), hjust = 0),
      axis.title          = element_text(size = rel(0.95)),
      axis.text           = element_text(size = rel(0.88)),
      legend.position     = "bottom",
      legend.text         = element_text(size = rel(0.88)),
      legend.title        = element_text(size = rel(0.9), face = "bold"),
      strip.text          = element_text(face = "bold", size = rel(0.95))
    )
}

canon_metric <- function(x) dplyr::recode(
  x,
  "advancing_knowledge" = "adv_knowledge",
  "open_science"        = "open_sci",
  "logic_communication" = "logic_comms",
  "global_relevance"    = "gp_relevance",
  "claims_evidence"     = "claims",
  .default = x
)

`%||%` <- function(x, y) if (!is.null(x)) x else y

# Pricing per 1M tokens (USD)
pricing <- tribble(
  ~model,              ~input_per_m, ~output_per_m,
  "GPT-5 Pro",         15.00,        60.00,
  "GPT-5.2 Pro",       15.00,        60.00,
  "GPT-4o-mini",       0.15,         0.60,
  "Claude Sonnet 4",   3.00,         15.00,
  "Claude Opus 4.6",   15.00,        75.00,
  "Gemini 2.0 Flash",  0.075,        0.30
)
```

```{r}
#| label: load-human-data
#| code-fold: true
#| code-summary: "Load human evaluation data"
#| message: false

# Load human ratings from Unjournal
UJmap <- read_delim("data/UJ_map.csv", delim = ";", show_col_types = FALSE) |>
  mutate(label_paper_title = research, label_paper = paper) |>
  select(c("label_paper_title", "label_paper"))

rsx <- read_csv("data/rsx_evalr_rating.csv", show_col_types = FALSE) |>
  clean_names() |>
  mutate(label_paper_title = research) |>
  select(-c("research"))

research <- read_csv("data/research.csv", show_col_types = FALSE) |>
  clean_names() |>
  filter(status == "50_published evaluations (on PubPub, by Unjournal)") |>
  left_join(UJmap, by = c("label_paper_title")) |>
  mutate(doi = str_trim(doi)) |>
  mutate(label_paper = case_when(
    doi == "https://doi.org/10.3386/w31162" ~ "Walker et al. 2023",
    doi == "doi.org/10.3386/w32728" ~ "Hahn et al. 2025",
    doi == "https://doi.org/10.3386/w30011" ~ "Bhat et al. 2022",
    doi == "10.1093/wbro/lkae010" ~ "Crawfurd et al. 2023",
    TRUE ~ label_paper
  )) |>
  left_join(rsx, by = c("label_paper_title"))

key_map <- research |>
  transmute(label_paper_title = str_trim(label_paper_title), label_paper = label_paper) |>
  filter(!is.na(label_paper_title)) |>
  distinct(label_paper_title, label_paper) |>
  group_by(label_paper_title) |>
  slice(1) |>
  ungroup()

rsx_research <- rsx |>
  mutate(label_paper_title = str_trim(label_paper_title)) |>
  left_join(key_map, by = "label_paper_title", relationship = "many-to-one")

metrics_human <- rsx_research |>
  mutate(criteria = canon_metric(criteria)) |>
  filter(criteria %in% c("overall", "claims", "methods", "adv_knowledge", "logic_comms", "open_sci", "gp_relevance")) |>
  transmute(
    paper = label_paper, criteria, evaluator, model = "Human",
    mid = as.numeric(middle_rating),
    lo = suppressWarnings(as.numeric(lower_ci)),
    hi = suppressWarnings(as.numeric(upper_ci))
  ) |>
  filter(!is.na(paper), !is.na(mid)) |>
  mutate(
    lo = ifelse(is.finite(lo), pmax(0, pmin(100, lo)), NA_real_),
    hi = ifelse(is.finite(hi), pmax(0, pmin(100, hi)), NA_real_)
  ) |>
  mutate(across(c(mid, lo, hi), ~ round(.x, 4))) |>
  distinct(paper, criteria, model, evaluator, mid, lo, hi)

human_avg <- metrics_human |>
  filter(criteria == "overall") |>
  group_by(paper) |>
  summarise(
    human_mid = mean(mid, na.rm = TRUE),
    human_lo = mean(lo, na.rm = TRUE),
    human_hi = mean(hi, na.rm = TRUE),
    n_human = n(),
    .groups = "drop"
  )

# Summary stats for inline use
n_human_papers <- n_distinct(metrics_human$paper)
n_human_evaluators <- n_distinct(metrics_human$evaluator)
```

```{r}
#| label: load-llm-data
#| code-fold: true
#| code-summary: "Load LLM evaluation data (all models)"
#| message: false

model_dirs <- list(
  "gpt5_pro_updated_jan2026" = "GPT-5 Pro",
  "gpt52_pro_focal_jan2026" = "GPT-5.2 Pro",
  "gpt_4o_mini_2024_07_18" = "GPT-4o-mini",
  "claude_sonnet_4_20250514" = "Claude Sonnet 4",
  "claude_opus_4_6" = "Claude Opus 4.6",
  "gemini_2.0_flash" = "Gemini 2.0 Flash"
)

parse_response <- function(path, model_name) {
  tryCatch({
    r <- jsonlite::fromJSON(path, simplifyVector = FALSE)
    paper <- basename(path) |>
      str_replace("\\.response\\.json$", "") |>
      str_replace_all("_", " ")

    parsed <- NULL
    if (!is.null(r$parsed) && length(r$parsed) > 0) {
      parsed <- r$parsed
    } else if (!is.null(r$output_text) && nchar(r$output_text) > 0) {
      # Strip markdown code fences if present
      txt <- r$output_text
      txt <- sub("^\\s*```[a-z]*\\s*\n?", "", txt)
      txt <- sub("\\s*```\\s*$", "", txt)
      parsed <- jsonlite::fromJSON(txt, simplifyVector = TRUE)
    } else if (!is.null(r$output)) {
      msg <- purrr::detect(r$output, ~ .x$type == "message", .default = NULL)
      if (!is.null(msg) && length(msg$content) > 0) {
        parsed <- jsonlite::fromJSON(msg$content[[1]]$text, simplifyVector = TRUE)
      }
    }

    if (is.null(parsed)) return(NULL)

    metrics <- parsed$metrics
    metric_rows <- list()
    tier_rows <- list()

    tier_names <- c("tier_should", "tier_will", "journal_should", "journal_will")

    for (nm in names(metrics)) {
      if (nm %in% tier_names) {
        # Normalise journal_should/journal_will → tier_should/tier_will
        tier_kind <- sub("^journal_", "tier_", nm)
        tier_rows[[length(tier_rows) + 1]] <- tibble(
          paper = paper, model = model_name, tier_kind = tier_kind,
          score = metrics[[nm]]$score,
          ci_lower = metrics[[nm]]$ci_lower,
          ci_upper = metrics[[nm]]$ci_upper
        )
      } else {
        metric_rows[[length(metric_rows) + 1]] <- tibble(
          paper = paper, model = model_name, metric = nm,
          midpoint = metrics[[nm]]$midpoint,
          lower_bound = metrics[[nm]]$lower_bound,
          upper_bound = metrics[[nm]]$upper_bound
        )
      }
    }

    input_tok <- r$usage$input_tokens %||% r$input_tokens
    output_tok <- r$usage$output_tokens %||% r$output_tokens
    reasoning_tok <- r$usage$output_tokens_details$reasoning_tokens

    list(
      metrics = bind_rows(metric_rows),
      tiers = bind_rows(tier_rows),
      tokens = tibble(
        paper = paper, model = model_name,
        input_tokens = input_tok %||% NA_integer_,
        output_tokens = output_tok %||% NA_integer_,
        reasoning_tokens = reasoning_tok %||% NA_integer_
      ),
      summary = tibble(
        paper = paper, model = model_name,
        assessment_summary = parsed$assessment_summary
      )
    )
  }, error = function(e) NULL)
}

load_all_llm <- function() {
  all_metrics <- list()
  all_tiers <- list()
  all_tokens <- list()
  all_summaries <- list()

  for (dir_name in names(model_dirs)) {
    model_name <- model_dirs[[dir_name]]
    json_dir <- here("results", dir_name, "json")

    if (dir.exists(json_dir)) {
      files <- list.files(json_dir, pattern = "\\.response\\.json$", full.names = TRUE)

      for (f in files) {
        result <- parse_response(f, model_name)
        if (!is.null(result)) {
          all_metrics[[length(all_metrics) + 1]] <- result$metrics
          all_tiers[[length(all_tiers) + 1]] <- result$tiers
          all_tokens[[length(all_tokens) + 1]] <- result$tokens
          all_summaries[[length(all_summaries) + 1]] <- result$summary
        }
      }
    }
  }

  list(
    metrics = bind_rows(all_metrics) |> mutate(criteria = canon_metric(metric)),
    tiers = bind_rows(all_tiers),
    tokens = bind_rows(all_tokens),
    summaries = bind_rows(all_summaries)
  )
}

llm_data <- load_all_llm()
llm_metrics <- llm_data$metrics
llm_tiers <- llm_data$tiers
llm_tokens <- llm_data$tokens
llm_summaries <- llm_data$summaries

# Summary stats for inline use
n_llm_models <- n_distinct(llm_metrics$model)
n_llm_papers <- n_distinct(llm_metrics$paper)
llm_model_names <- unique(llm_metrics$model)
llm_model_list <- paste(llm_model_names, collapse = ", ")

# Papers with both human and LLM data
matched_papers <- intersect(
 unique(metrics_human$paper),
 unique(llm_metrics$paper)
)
n_matched <- length(matched_papers)

# Cost summary
cost_data <- llm_tokens |>
  left_join(pricing, by = "model") |>
  mutate(
    reasoning_tokens = coalesce(reasoning_tokens, 0L),
    cost_usd = (input_tokens * input_per_m + output_tokens * output_per_m) / 1e6
  )

total_cost <- sum(cost_data$cost_usd, na.rm = TRUE)
total_evaluations <- nrow(llm_tokens)

# Per-model summaries
model_summary <- cost_data |>
  group_by(model) |>
  summarise(
    n_papers = n(),
    avg_input = round(mean(input_tokens, na.rm = TRUE)),
    avg_output = round(mean(output_tokens, na.rm = TRUE)),
    avg_reasoning = round(mean(reasoning_tokens[reasoning_tokens > 0], na.rm = TRUE)),
    total_cost = sum(cost_usd, na.rm = TRUE),
    cost_per_paper = mean(cost_usd, na.rm = TRUE),
    .groups = "drop"
  )
```

We compare quantitative ratings from `r n_llm_models` frontier LLMs (`r llm_model_list`) against human expert reviews from [The Unjournal](https://unjournal.pubpub.org). `r n_matched` papers have both human and LLM evaluations, enabling direct comparison across all seven percentile criteria and two journal-tier predictions.

We submit each PDF through a single, schema-enforced API call. The system prompt mirrors The Unjournal rubric, blocks extrinsic priors (authors, venue, web context), and requires strict JSON: a ~1k-word diagnostic summary plus numeric midpoints and 90% credible intervals for every metric.^[0–100 percentiles relative to a reference group of ~'all serious work you have read in this area in the last three years', overall (global quality/impact), claims_evidence (claim clarity/support), methods (design/identification/robustness), advancing_knowledge (contribution to field/practice), logic_communication (internal consistency and exposition), open_science (reproducibility, data/code availability), and global_relevance (decision and priority usefulness). Journal tiers: (0–5, continuous) capture where the work should publish vs. will publish.] Each numeric field carries a midpoint (model's 50% belief) and a 90% credible interval; wide intervals indicate acknowledged uncertainty.

**Cost and token usage.** Token costs are computed from API-reported input/output tokens (GPT-5 Pro runs with `reasoning` = high).

```{r}
#| label: tbl-cost-overview
#| tbl-cap: "Token usage and estimated cost per model"

if (nrow(llm_tokens) > 0) {

  cost_table <- model_summary |>
    transmute(
      Model = model,
      Papers = n_papers,
      `Avg input` = scales::comma(avg_input),
      `Avg output` = scales::comma(avg_output),
      `Avg reasoning` = ifelse(is.finite(avg_reasoning) & avg_reasoning > 0,
                                scales::comma(avg_reasoning), "—"),
      `Total cost` = scales::dollar(total_cost, accuracy = 0.01),
      `Cost/paper` = scales::dollar(cost_per_paper, accuracy = 0.001)
    ) |>
    arrange(desc(as.numeric(gsub("[$,]", "", `Total cost`))))

  knitr::kable(cost_table, align = c("l", rep("r", 6)))
}
```

Total cost across all `r total_evaluations` LLM evaluations: **`r scales::dollar(total_cost, accuracy = 0.01)`**. Reasoning-capable models (GPT-5 Pro, GPT-5.2 Pro, Claude Opus 4.6) cost substantially more per paper than their lighter counterparts (GPT-4o-mini, Gemini 2.0 Flash), raising practical questions about cost-quality trade-offs for deployment at scale.

```{r}
#| label: scatter-data-prep
#| include: false

# Prepare scatter_data for tbl-agreement (figure in main Results chapter)
scatter_data <- llm_metrics |>
  filter(criteria == "overall") |>
  inner_join(human_avg, by = "paper") |>
  mutate(
    diff = midpoint - human_mid,
    paper_short = str_trunc(paper, 25),
    human_lo = coalesce(human_lo, human_mid),
    human_hi = coalesce(human_hi, human_mid),
    lower_bound = coalesce(lower_bound, midpoint),
    upper_bound = coalesce(upper_bound, midpoint)
  ) |>
  filter(!is.na(human_mid), !is.na(midpoint))
```

**Agreement metrics.** We assess agreement using Pearson *r* (linear correlation), Spearman *ρ* (rank-based, robust to outliers), mean bias (LLM − Human; positive = LLM rates higher), and RMSE/MAE (error magnitude). Full definitions of all statistics appear in [Methods](methods.qmd).

```{r}
#| label: tbl-agreement
#| tbl-cap: "Agreement between each LLM and human mean ratings (overall criterion)"

if (nrow(scatter_data) > 0) {

  agreement_tbl <- scatter_data |>
    group_by(Model = model) |>
    summarise(
      N = n(),
      `Pearson r` = round(cor(human_mid, midpoint, use = "complete.obs"), 3),
      `Spearman ρ` = round(cor(human_mid, midpoint, method = "spearman", use = "complete.obs"), 3),
      `Mean bias` = sprintf("%+.1f", mean(midpoint - human_mid, na.rm = TRUE)),
      RMSE = round(sqrt(mean((midpoint - human_mid)^2, na.rm = TRUE)), 1),
      MAE = round(mean(abs(midpoint - human_mid), na.rm = TRUE), 1),
      .groups = "drop"
    ) |>
    arrange(desc(`Pearson r`))

  knitr::kable(agreement_tbl, align = c("l", rep("r", 6)))
}
```

@tbl-agreement reports correlation and error metrics between each LLM and the human mean overall rating. Positive mean bias indicates the model rates papers higher than humans on average; negative bias indicates harsher ratings. These correlations are bounded above by human inter-rater noise---since the human "mean" is itself noisy, no model can achieve a perfect correlation.

**Human baseline context.** To interpret human-LLM agreement, we compare against human-human agreement as an upper bound. If human evaluators only agree with each other at α = 0.5, expecting LLMs to exceed that would be unrealistic. @tbl-human-baseline shows α~HH~ (Krippendorff's alpha among human evaluators) alongside α~HL~ (between the human mean and each LLM).

```{r}
#| label: tbl-human-baseline
#| tbl-cap: "Human-human vs Human-LLM agreement by criterion (Krippendorff's α)"
#| code-fold: true

if (requireNamespace("irr", quietly = TRUE)) {

  library(irr)

  criteria_list <- c("overall", "claims", "methods", "adv_knowledge",
                     "logic_comms", "open_sci", "gp_relevance")

  criterion_labels <- c(
    overall = "Overall", claims = "Claims", methods = "Methods",
    adv_knowledge = "Adv. Knowledge", logic_comms = "Logic & Comms",
    open_sci = "Open Science", gp_relevance = "Global Relevance"
  )

  # Compute human-human alpha (each evaluator as separate rater)
  compute_hh_alpha <- function(criterion) {
    human_wide <- metrics_human |>
      filter(criteria == criterion, paper %in% matched_papers) |>
      select(paper, evaluator, mid) |>
      distinct() |>
      pivot_wider(names_from = paper, values_from = mid)

    if (nrow(human_wide) < 2 || ncol(human_wide) < 3) return(NA_real_)

    M <- as.matrix(human_wide[, -1, drop = FALSE])
    tryCatch({
      irr::kripp.alpha(M, method = "interval")$value
    }, error = function(e) NA_real_)
  }

  # Compute human-LLM alpha for a specific model
  compute_hl_alpha <- function(criterion, model_name) {
    human_mean <- metrics_human |>
      filter(criteria == criterion, paper %in% matched_papers) |>
      group_by(paper) |>
      summarise(Human = mean(mid, na.rm = TRUE), .groups = "drop")

    llm_rating <- llm_metrics |>
      filter(criteria == criterion, model == model_name, paper %in% matched_papers) |>
      select(paper, midpoint) |>
      rename(LLM = midpoint)

    combined <- inner_join(human_mean, llm_rating, by = "paper")
    if (nrow(combined) < 3) return(NA_real_)

    M <- rbind(Human = combined$Human, LLM = combined$LLM)
    tryCatch({
      irr::kripp.alpha(M, method = "interval")$value
    }, error = function(e) NA_real_)
  }

  # Get unique LLM models
  llm_models <- unique(llm_metrics$model)

  # Build comparison table
  baseline_results <- tibble(Criterion = criterion_labels[criteria_list]) |>
    mutate(criterion_key = criteria_list) |>
    rowwise() |>
    mutate(
      `α_HH` = compute_hh_alpha(criterion_key)
    ) |>
    ungroup()

  # Add columns for each LLM model
  for (mod in llm_models) {
    col_name <- paste0("α_HL (", mod, ")")
    baseline_results <- baseline_results |>
      rowwise() |>
      mutate(!!col_name := compute_hl_alpha(criterion_key, mod)) |>
      ungroup()
  }

  # Format for display
  baseline_display <- baseline_results |>
    select(-criterion_key) |>
    mutate(across(where(is.numeric), ~ ifelse(is.na(.x), "—", sprintf("%.2f", .x))))

  knitr::kable(baseline_display, align = c("l", rep("r", ncol(baseline_display) - 1)))
}
```

Where α~HL~ approaches α~HH~, the LLM is performing at or near the ceiling set by human inter-rater variability. Criteria where α~HH~ is already low (indicating genuine evaluator disagreement) naturally cap how well any model can do. Extended agreement analysis---including CI coverage, bootstrap confidence intervals, mutual information, and per-criterion correlations---is available in the supplementary material.

**Criteria-level patterns.** @fig-criteria-heatmap displays the mean rating each evaluator type assigns across the seven criteria. Criteria where all models and humans converge (similar colour intensity) suggest robust agreement on that dimension; criteria with divergent columns indicate systematic differences in how humans and LLMs weigh that aspect of quality.

```{r}
#| label: fig-criteria-heatmap
#| fig-cap: "Mean rating by evaluator type (rows) and criterion (columns), averaged across all matched papers. Tile colour scales from light (lower mean) to vivid orange (higher mean), clamped to 40\u201380. Criteria where all rows show similar colour intensity indicate broad agreement; divergent columns flag criteria-level systematic differences."
#| fig-width: 11
#| fig-height: 6.5

metric_labels <- c(
  overall = "Overall", claims = "Claims", methods = "Methods",
  adv_knowledge = "Adv. Knowledge", logic_comms = "Logic & Comms",
  open_sci = "Open Science", gp_relevance = "Global Relevance"
)

criteria_means <- bind_rows(
  metrics_human |>
    filter(criteria %in% names(metric_labels), paper %in% matched_papers) |>
    group_by(criteria) |>
    summarise(mean_rating = mean(mid, na.rm = TRUE), .groups = "drop") |>
    mutate(model = "Human"),
  llm_metrics |>
    filter(criteria %in% names(metric_labels), paper %in% matched_papers) |>
    group_by(model, criteria) |>
    summarise(mean_rating = mean(midpoint, na.rm = TRUE), .groups = "drop")
) |>
  mutate(criteria_label = factor(metric_labels[criteria], levels = metric_labels))

if (nrow(criteria_means) > 0) {
  ggplot(criteria_means, aes(x = criteria_label, y = model, fill = mean_rating)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(aes(label = round(mean_rating, 0)), size = 4.2, color = "white", fontface = "bold") +
    scale_fill_gradient(low = "#f0f0f0", high = UJ_ORANGE, limits = c(40, 80), name = "Mean rating") +
    labs(x = NULL, y = NULL) +
    theme_uj() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 11),
          axis.text.y = element_text(size = 11),
          panel.grid = element_blank())
}
```

**Additional models.** @fig-scatter-other shows overall ratings for the two lightweight models---GPT-4o-mini and Gemini 2.0 Flash---against human mean ratings. These models cost 50--200× less per paper than reasoning-capable models (see @tbl-cost-overview) but may sacrifice rating accuracy.

```{r}
#| label: fig-scatter-other
#| fig-cap: "Human mean overall rating (x-axis) vs LLM overall rating (y-axis) for the two lightweight models GPT-4o-mini and Gemini 2.0 Flash. Each point is one paper; the dashed diagonal is the identity line. These models cost 50\u2013200\u00d7 less per paper than reasoning-capable models (see cost table above)."
#| fig-width: 11
#| fig-height: 6

other_models <- c("GPT-4o-mini", "Gemini 2.0 Flash")

scatter_other <- llm_metrics |>
  filter(criteria == "overall", model %in% other_models) |>
  inner_join(human_avg, by = "paper") |>
  mutate(
    diff = midpoint - human_mid,
    paper_short = str_trunc(paper, 25),
    human_lo = coalesce(human_lo, human_mid),
    human_hi = coalesce(human_hi, human_mid),
    lower_bound = coalesce(lower_bound, midpoint),
    upper_bound = coalesce(upper_bound, midpoint)
  ) |>
  filter(!is.na(human_mid), !is.na(midpoint))

if (nrow(scatter_other) > 0) {
  ggplot(scatter_other, aes(x = human_mid, y = midpoint, color = model)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
    geom_point(size = 3.5, alpha = 0.7) +
    ggrepel::geom_text_repel(aes(label = paper_short), size = 2.4, max.overlaps = 8, show.legend = FALSE) +
    scale_color_manual(values = MODEL_COLORS) +
    coord_cartesian(xlim = c(0, 100), ylim = c(0, 100)) +
    facet_wrap(~model, ncol = 2) +
    labs(x = "Human mean rating (0\u2013100)", y = "LLM rating (0\u2013100)", color = NULL) +
    theme_uj() +
    theme(strip.text = element_text(size = 13, face = "bold"))
} else {
  cat("No matching data for lightweight model scatter plot.\n")
}
```

For per-paper scatter plots, rank comparisons, and the gap heatmap showing Human − LLM differences by paper and criterion, see the [main Results chapter](results.qmd). Extended reasoning traces are available in [Appendix C: LLM Traces](appendix_llm_traces.qmd).