---
title: "Results: Critiques & Key Issues"
format:
  html: default
engine: knitr
---

```{r}
#| label: setup-critiques
#| code-summary: "Setup and libraries"
#| code-fold: true
#| message: false
#| warning: false

library("tidyverse")
library("jsonlite")
library("knitr")
library("kableExtra")
library("DT")

# Theme colors (matching results_ratings.qmd)
UJ_ORANGE <- "#f19e4b"
UJ_GREEN  <- "#99bb66"
UJ_BLUE   <- "#4e79a7"

theme_uj <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot",
      legend.position = "bottom"
    )
}
```

This chapter compares **qualitative critiques**: the key methodological and interpretive issues identified by GPT-5.2 Pro against human expert critiques documented by Unjournal evaluation managers in Coda.

## Data Sources

**GPT-5.2 Pro Key Issues**: Structured output from the focal evaluation run (January 2026), where the model was prompted to identify 5-15 key issues per paper, ordered from most to least important.

**Human Expert Critiques**: Curated content from the "Key critiques & issues with paper" column in The Unjournal's internal tracking database (Coda), written by evaluation managers synthesizing evaluator feedback. These use severity labels ("Necessary", "Optional but important", "Unsure") and cite specific evaluator comments.

```{r}
#| label: load-comparison-data
#| code-fold: true
#| code-summary: "Load comparison data"

# Load the matched comparison data
comparison_file <- "results/key_issues_comparison.json"

if (file.exists(comparison_file)) {
  comparison_data <- fromJSON(comparison_file)
  n_papers <- nrow(comparison_data)
} else {
  comparison_data <- NULL
  n_papers <- 0
}
```

We matched **`r n_papers` papers** with both GPT-5.2 Pro key issues and human expert critiques.

## Overview Statistics

```{r}
#| label: tbl-overview-stats
#| tbl-cap: "Summary of key issues comparison data"

if (!is.null(comparison_data) && n_papers > 0) {
  overview <- comparison_data |>
    mutate(
      num_gpt_issues = sapply(gpt_key_issues, length)
    ) |>
    summarise(
      `Papers Compared` = n(),
      `Avg GPT Issues per Paper` = round(mean(num_gpt_issues), 1),
      `Min GPT Issues` = min(num_gpt_issues),
      `Max GPT Issues` = max(num_gpt_issues),
      `Avg Coda Critique Length (chars)` = round(mean(coda_critique_length)),
      `Min Coda Length` = min(coda_critique_length),
      `Max Coda Length` = max(coda_critique_length)
    ) |>
    pivot_longer(everything(), names_to = "Metric", values_to = "Value")

  kable(overview, align = c("l", "r"))
}
```

## Paper-by-Paper Comparison {#sec-paper-comparison}

The sections below show side-by-side comparisons for each matched paper. Human critiques often use severity labels and cite specific evaluators (E1, E2, etc.), while GPT issues are concise bullet points ordered by importance.

```{r}
#| label: render-comparisons
#| results: asis
#| code-fold: true

if (!is.null(comparison_data) && n_papers > 0) {

  for (i in 1:nrow(comparison_data)) {
    row <- comparison_data[i, ]

    # Paper header
    cat(sprintf("\n### %s\n\n", row$gpt_paper))
    cat(sprintf("**Coda title:** %s\n\n", row$coda_title))

    # Two-column layout using Quarto divs
    cat("::: {.columns}\n")
    cat("::: {.column width=\"48%\"}\n")
    cat("#### GPT-5.2 Pro Key Issues\n\n")

    # GPT issues
    gpt_issues <- row$gpt_key_issues[[1]]
    for (issue in gpt_issues) {
      # Clean up numbering if present
      issue_clean <- gsub("^[0-9]+[)\\.]\\s*", "", issue)
      cat(sprintf("- %s\n", issue_clean))
    }

    cat("\n:::\n")
    cat("::: {.column width=\"4%\"}\n")
    cat("\n:::\n")  # spacer
    cat("::: {.column width=\"48%\"}\n")
    cat("#### Human Expert Critiques\n\n")

    # Coda critique (preserve formatting)
    coda_text <- row$coda_critique
    # Convert line breaks to markdown
    coda_text <- gsub("\n", "\n\n", coda_text)
    cat(coda_text)

    cat("\n:::\n")
    cat(":::\n\n")
    cat("---\n\n")
  }
}
```

## Summary and Observations

```{r}
#| label: summary-analysis
#| code-fold: true

if (!is.null(comparison_data) && n_papers > 0) {

  # Calculate some summary statistics
  issue_counts <- comparison_data |>
    mutate(
      num_gpt_issues = sapply(gpt_key_issues, length),
      paper_short = str_trunc(gpt_paper, 30)
    ) |>
    select(paper_short, num_gpt_issues, coda_critique_length)
}
```

### Issue Count Distribution

```{r}
#| label: fig-issue-counts
#| fig-cap: "Number of key issues identified by GPT-5.2 Pro per paper"
#| fig-width: 10
#| fig-height: 5

if (!is.null(comparison_data) && n_papers > 0) {

  issue_counts |>
    ggplot(aes(x = reorder(paper_short, num_gpt_issues), y = num_gpt_issues)) +
    geom_col(fill = UJ_ORANGE, alpha = 0.8) +
    geom_hline(yintercept = mean(issue_counts$num_gpt_issues),
               linetype = "dashed", color = UJ_BLUE, linewidth = 1) +
    coord_flip() +
    labs(
      x = NULL,
      y = "Number of Key Issues",
      subtitle = paste0("Dashed line = mean (", round(mean(issue_counts$num_gpt_issues), 1), " issues)")
    ) +
    theme_uj()
}
```

### Coda Critique Length vs GPT Issue Count

```{r}
#| label: fig-length-scatter
#| fig-cap: "Relationship between human critique detail and GPT issue count"
#| fig-width: 8
#| fig-height: 6

if (!is.null(comparison_data) && n_papers > 0) {

  ggplot(issue_counts, aes(x = coda_critique_length, y = num_gpt_issues)) +
    geom_point(size = 3, color = UJ_ORANGE, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, color = UJ_BLUE, alpha = 0.2) +
    labs(
      x = "Human Critique Length (characters)",
      y = "GPT-5.2 Pro Issue Count"
    ) +
    theme_uj()
}
```

### Observable Structural Differences

The following are **verifiable structural differences** between the two data sources (not assessments of quality or coverage):

| Aspect | GPT-5.2 Pro | Human Expert (Coda) |
|--------|-------------|---------------------|
| Format | Numbered bullet points (array of strings) | Free-form prose |
| Structure | Ordered list (prompted: "most to least important") | Often uses severity labels ("Necessary", "Optional but important") |
| Source attribution | None (single model output) | Often cites specific evaluators (E1, E2, names) |
| Length | Constrained by prompt (~10-12 issues) | Unconstrained, highly variable |

### Questions for Manual Review

The side-by-side comparisons above are provided for **manual expert assessment**. Key questions to investigate:

1. **Coverage**: What proportion of human-identified issues appear in the GPT output (in some form)?
2. **Precision**: Are GPT issues substantive or does the model identify spurious/generic concerns?
3. **Severity alignment**: Does GPT's importance ordering correlate with human severity labels?
4. **Missed issues**: Are there critical human concerns that GPT systematically misses?
5. **Novel issues**: Does GPT surface valid concerns that humans overlooked?

::: {.callout-warning}
## No Automated Assessment Yet

This page displays raw data for manual review. The claims about coverage, precision, and alignment **have not been assessed**—they require either:

- **Manual annotation** using the Issue Annotation Tool (see below)
- LLM-based comparison (run `python compare_key_issues.py --compare` when API is available)

Any conclusions about "high overlap" or "systematic differences" should emerge from that analysis, not be assumed.
:::

## Manual Annotation Tool

To systematically assess concordance between human and LLM critiques, use the **Issue Annotation Tool**:

```bash
# 1. Generate annotation data (parses human critiques into individual issues)
python3 tools/build_issue_annotation_data.py

# 2. Open the annotation UI in a browser
open tools/issue_annotation_ui/index.html
```

**Annotation workflow:**

For each human-identified issue:

1. **Match score (0-1)**: How well do LLM issues capture this concern?
   - 0 = Not addressed at all
   - 0.5 = Partially captured or tangentially related
   - 1 = Fully captured by one or more LLM issues

2. **Confidence (0-1)**: How certain are you of this assessment?

3. **Context flag**: Check if the human critique references information the LLM didn't have (e.g., author responses, evaluator discussions post-evaluation)

4. **Link to LLM issues**: Select which LLM issues (L1, L2, ...) correspond to this human issue

5. **Discussion**: Explain your reasoning, note partial matches, or flag ambiguities

**Export** annotations as JSON/CSV for analysis. Annotations are auto-saved to browser localStorage.

::: {.callout-note}
## Annotation Data Schema

The tool parses human critiques using heuristics:

- **Severity labels** normalized to: `necessary`, `optional`, `unsure`
- **Evaluator attributions** (E1, E2, DR) preserved in issue text
- **Issue boundaries** detected via enumeration, sentence breaks, and section headers

Review parsed issues in the UI and edit/add/remove as needed before annotating.
:::
